---
title: "Regression and Classification Trees"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{RPART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r load,include=TRUE,echo=TRUE,results=FALSE,warning=FALSE,message=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description
This shows the output of PART RPART functions from rwf.
Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)
These functions are not installed by default they are in /rwf/working_functions/OTHER/ML_TREE_RPART.R

## Theoretical Background

### Recursive partitioning

`rpart` implements **CART** (Classification and Regression Trees, Breiman et al., 1984). Starting from the full dataset at the root node, the algorithm searches every predictor and every possible split point on that predictor, and picks the single split that most improves node purity. Each resulting child node is then split again, recursively, until a stopping rule is reached (a minimum number of observations per node, a minimum improvement threshold, or a maximum depth). The result is a **binary tree**: every internal node asks a yes/no question about one predictor, and every terminal node ("leaf") gives a prediction.

### Splitting criteria

The measure of "improvement" depends on the outcome type:

- **Classification** (`method = "class"`, used below for `case`) — by default `rpart` uses the **Gini impurity** of the class distribution within a node; a split is chosen to make the two child nodes as class-pure as possible.
- **Regression** (`method = "anova"`, used below for `medv`) — the criterion is the reduction in **within-node sum of squares**, i.e. the split that most reduces the variance of the outcome within each child node.

### Cost-complexity pruning

A tree grown to full depth almost always **overfits** — it captures noise along with signal. `rpart` addresses this with **cost-complexity pruning**: every subtree $T$ is scored as

$$R_\alpha(T) = R(T) + \alpha \, |T|$$

where $R(T)$ is the tree's error (relative to the root), $|T|$ is the number of terminal nodes, and $\alpha$ (called `CP` in `rpart`'s output) penalizes tree size. Varying $\alpha$ from 0 to 1 produces a nested sequence of subtrees, from the full tree down to a single root node — this sequence is stored in `model$cptable`.

### Cross-validated tree size

Because the *apparent* error of a tree always decreases as it grows (a larger tree can always fit the training data at least as well), `rpart` estimates each subtree's error via internal **k-fold cross-validation**. `cptable` reports, for every candidate `CP` value: the relative error on the training data (`rel error`), the cross-validated error (`xerror`), and its standard deviation (`xstd`). The code below selects the tree whose `CP` gives the **minimum `xerror`** — a common, slightly more permissive alternative is the "1-SE rule" (the smallest tree within one standard error of the minimum), which favors an even simpler tree.

### Variable importance

A predictor's importance is the sum of the improvement in impurity (Gini, for classification) or sum-of-squares (for regression) attributable to it — counting not only splits where it was the *primary* splitter, but also splits where it acted as a **surrogate** (a stand-in used when the primary splitting variable is missing). This is why a variable can show non-trivial importance even if it never appears as a labeled split in the final tree diagram.

```{r create_models,include=TRUE}
infert_formula<-formula(factor(case)~age+parity+education+spontaneous+induced)
boston_formula<-formula(c("medv~",paste(names(MASS::Boston)[1:13],collapse="+")))

print(infert_formula)
print(boston_formula)

# kfolding
train_test_classification<-k_fold(df=infert,
                                  model_formula=infert_formula)
train_test_regression<-k_fold(df=MASS::Boston,
                              model_formula=boston_formula)

# srpart models
rtree_classification<-rpart::rpart(infert_formula,
                                   train_test_classification$f$train$f1,
                                   model=TRUE,x=TRUE,y=TRUE)
rtree_regression<-rpart::rpart(boston_formula,
                               train_test_regression$f$train$f1,
                               model=TRUE,x=TRUE,y=TRUE)

```

## Rpart Classification

The classification tree predicts `case` (secondary infertility, from the classic `infert` case-control study) from `age`, `parity`, `education`, `spontaneous`, and `induced` (prior abortions).

### Growing and pruning the tree

`rtree_classification` was grown to full depth in the `create_models` chunk above. Here it is pruned back to the `CP` value that minimizes the cross-validated error (`xerror`) in its `cptable`.

```{r classification_prune,echo=TRUE}
result<-data.frame(rtree_classification$cptable)
result$nsplit<-factor(result$nsplit+1)
minimun_size<-as.numeric(as.character(result[which.min(result[,"xerror"]),"nsplit"]))
initial_model<-rtree_classification
model<-rpart::prune(rtree_classification,
                    cp=rtree_classification$cptable[which.min(rtree_classification$cptable[,"xerror"]),"CP"])
```

### Variable importance

```{r classification_importance,echo=TRUE,dev='png'}
importance<-model$variable.importance
importance<-data.frame(names=names(importance),importance=importance)
importance$names<-factor(importance$names,levels=rev(as.character(importance$names)))
plot_importance<-ggplot(importance,aes(x=names,y=importance))+
  geom_bar(stat='identity')+
  labs(title="Importance Plot",y="Relative Influence",x="Predictor")+
  theme_bw(base_size=10)+
  scale_x_discrete(limits=rev(levels(names)))+
  coord_flip()
plot_importance
```

Longer bars mean the predictor contributed more to reducing impurity across the tree's splits (as primary splitter or as a surrogate). For `infert`, `age` and `education` typically dominate — consistent with the epidemiological finding that secondary infertility risk in this dataset tracks strongly with age and prior reproductive history.

### The full, unpruned tree

```{r classification_full_tree_text,echo=TRUE}
rtree_classification
```

This is the text representation of the tree *before* pruning — every line is one node: its splitting rule, the number of observations that reach it (`n`), the misclassification loss, and the predicted class with its probability. Compare its size to the pruned `model` used in the plots below — pruning typically collapses several of these nodes back together.

### Choosing the tree size via cross-validation

```{r classification_cv_size,echo=TRUE,dev='png'}
rpart::plotcp(model)
rpart::rsq.rpart(model)
```

`plotcp()` plots cross-validated **relative error** (`xerror`, with one-SE error bars) against `CP` (bottom axis) and tree size (top axis) — the dashed horizontal line marks one standard error above the minimum, the usual reference for the "1-SE rule". `rsq.rpart()` adds two more panels: apparent vs. cross-validated **R-squared** and **relative error** by tree size. Because this is a *classification* tree, these values are derived from the Gini/deviance-based loss rather than explained variance, so read them as "relative predictive error" rather than a literal R²: they still show the same pattern — error falls quickly for the first few splits, then flattens (or the cross-validated curve turns back up), which is exactly the point captured by pruning to minimum `xerror`.

### Error-by-size summary

```{r classification_prune_plot,echo=TRUE,dev='png'}
error<-data.frame(model$cptable)
error$nsplit<-factor(error$nsplit+1)
tree_size<-error[which.min(error[,"xerror"]),"nsplit"]
error<-reshape2::melt(error,id.vars="nsplit")
names(error)<-c("Split","Metric","value")
plot_prune<-ggplot(error,aes(x=Split,y=value,color=Metric))+
  geom_line(aes(group=Metric))+
  geom_point()+
  labs(title=paste("Error Plot","Suggested Size:",tree_size),y="Metric value",x="Size of Tree")+
  theme_bw(base_size=10)
plot_prune
```

This re-plots the same `cptable` metrics (`rel error`, `xerror`, `xstd`) on one axis, as tree size grows from a single node onward. The title reports the tree size (`nsplit`) at which `xerror` is minimized — the size the pruned `model` was cut back to.

### The pruned tree diagram

```{r classification_final_tree,echo=TRUE,dev='png'}
rpart.plot::rpart.plot(model,type=1)
```

Each box is a node: the top line is the predicted class (and, depending on plotting options, the predicted probability), the bottom line is the percentage of the sample that lands in that node. Following the branches down from the root shows the sequence of yes/no rules — e.g. an age threshold, then a threshold on prior spontaneous/induced abortions — that separate higher- from lower-risk groups.

### Raw model components

```{r classification_export,echo=TRUE}
frame<-data.frame(model$frame)
cp<-data.frame(model$cptable)
parameters<-data.frame(parameters=unlist(model$control))
splits<-data.frame(name=row.names(model$splits),model$splits,row.names=NULL)
importance<-data.frame(importance=model$variable.importance)
ordered<-data.frame(ordered=model$ordered)
data<-data.frame(y=model$y,x=model$x,model=model$model)
call<-data.frame(call=call_to_string(model))
result<-list(frame=frame,cp=cp,parameters=parameters,splits=splits,importance=importance,ordered=ordered,call=call)

print(result)
```

`result` bundles the pruned model's internals — node `frame` (one row per node), the `cp` table, control `parameters`, raw `splits`, `importance`, factor-`ordered` flags, and the model `call` — for programmatic inspection or export, without needing to pick through `model`'s S3 structure directly.

## Rpart Regression

The regression tree predicts `medv` (median owner-occupied home value, in $1000s) from the 13 other variables in the classic `MASS::Boston` housing dataset.

### Growing and pruning the tree

```{r regression_prune,echo=TRUE}
result<-data.frame(rtree_regression$cptable)
result$nsplit<-factor(result$nsplit+1)
minimun_size<-as.numeric(as.character(result[which.min(result[,"xerror"]),"nsplit"]))
initial_model<-rtree_regression
model<-rpart::prune(rtree_regression,cp=rtree_regression$cptable[which.min(rtree_regression$cptable[,"xerror"]),"CP"])
```

As with the classification tree, `rtree_regression` is pruned to the `CP` value that minimizes the cross-validated error — here, cross-validated sum-of-squares error rather than a classification loss.

### Variable importance

```{r regression_importance,echo=TRUE,dev='png'}
importance<-model$variable.importance
importance<-data.frame(names=names(importance),importance=importance)
importance$names<-factor(importance$names,levels=rev(as.character(importance$names)))
plot_importance<-ggplot(importance,aes(x=names,y=importance))+
  geom_bar(stat='identity')+
  labs(title="Importance Plot",y="Relative Influence",x="Predictor")+
  theme_bw(base_size=10)+
  scale_x_discrete(limits=rev(levels(names)))+
  coord_flip()
plot_importance
```

For Boston housing, `rm` (average rooms per dwelling) and `lstat` (% lower-status population) are almost always the two dominant predictors of `medv` — this importance ranking is the tree-based analogue of the strong linear associations well known from OLS models fit to this same dataset.

### The full, unpruned tree

```{r regression_full_tree_text,echo=TRUE}
rtree_regression
```

The unpruned tree's text form: each node reports its split rule, the number of training observations reaching it (`n`), the deviance (residual sum of squares) within the node, and the predicted `medv` (the node mean).

### Choosing the tree size via cross-validation

```{r regression_cv_size,echo=TRUE,dev='png'}
rpart::plotcp(model)
rpart::rsq.rpart(model)
```

Here `rsq.rpart()`'s R² panel is a literal R²: the proportion of `medv`'s variance explained, both on the training data (apparent) and under cross-validation. The gap between the two lines is the model's optimism — how much worse the tree performs on unseen data than on the data it was fit to — and it typically widens as the tree grows past the point where pruning is applied.

### Error-by-size summary

```{r regression_prune_plot,echo=TRUE,dev='png'}
error<-data.frame(model$cptable)
error$nsplit<-factor(error$nsplit+1)
tree_size<-error[which.min(error[,"xerror"]),"nsplit"]
error<-reshape2::melt(error,id.vars="nsplit")
names(error)<-c("Split","Metric","value")
plot_prune<-ggplot(error,aes(x=Split,y=value,color=Metric))+
  geom_line(aes(group=Metric))+
  geom_point()+
  labs(title=paste("Error Plot","Suggested Size:",tree_size),y="Metric value",x="Size of Tree")+
  theme_bw(base_size=10)
plot_prune
```

Same three `cptable` series (`rel error`, `xerror`, `xstd`) as in the classification section, plotted against tree size, with the size at minimum `xerror` called out in the title.

### The pruned tree diagram

```{r regression_final_tree,echo=TRUE,dev='png'}
rpart.plot::rpart.plot(model,type=1)
```

Each box shows the node's predicted `medv` and the percentage of observations it contains. Following the splits down — typically starting with `rm` or `lstat` — shows how the tree partitions the housing stock into progressively more homogeneous price segments, e.g. "high rooms, low lstat" neighborhoods landing in the highest-value terminal nodes.

### Raw model components

```{r regression_export,echo=TRUE}
frame<-data.frame(model$frame)
cp<-data.frame(model$cptable)
parameters<-data.frame(parameters=unlist(model$control))
splits<-data.frame(name=row.names(model$splits),model$splits,row.names=NULL)
importance<-data.frame(importance=model$variable.importance)
ordered<-data.frame(ordered=model$ordered)
data<-data.frame(y=model$y,x=model$x,model=model$model)
call<-data.frame(call=call_to_string(model))
result<-list(frame=frame,cp=cp,parameters=parameters,splits=splits,importance=importance,ordered=ordered,call=call)

print(result)
```

Same bundling as the classification section — the pruned regression model's `frame`, `cp` table, `parameters`, `splits`, `importance`, `ordered` flags, and `call`.

