---
title: "EAP Ability Estimation From Scratch"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{EAP Ability Estimation From Scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/")
source(paste0(directory, "code_options.R"))

library(ggplot2)
```

## Description

Adaptive testing packages like `catR` score examinees using **EAP (Expected A Posteriori)** ability estimation — a Bayesian method that is more numerically stable than Maximum Likelihood when a response pattern is all-correct, all-incorrect, or otherwise perfectly separates the item bank (situations where a raw likelihood has no finite maximum).

[`working_examples/cat_eap.R`](https://github.com/sedzinfo/rwf/blob/main/working_examples/cat_eap.R) reimplements the whole EAP pipeline **from scratch**, in five small functions, with no dependency on `catR` at all:

| Function | Purpose | `catR` equivalent it mirrors |
|---|---|---|
| `density_function()` | Normal density | `stats::dnorm()` |
| `integrate_cat()` | Numerical integration (trapezoidal rule) | `catR::integrate.catR()` |
| `Pi()` | 4PL response probability + derivative | `catR::Pi()` |
| `eap_est()` | EAP ability estimate | `catR::eapEst()` |
| `eap_se()` | EAP standard error | `catR::eapSem()` |

This vignette walks through the theory behind each function, then validates every one of them directly against its `catR` counterpart, and reproduces the worked examples from the script.

```{r library,message=FALSE,warning=FALSE}
library(catR)
```

```{r source_functions}
source(paste0(directory, "../working_examples/cat_eap.R"), local = knitr::knit_global())
```

---

## Theoretical Background

### The 4-parameter logistic (4PL) response model

Every item response model gives the probability that an examinee with ability $\theta$ answers item $i$ correctly:

$$P_i(\theta) = c_i + (d_i-c_i)\,\frac{e^{Da_i(\theta-b_i)}}{1+e^{Da_i(\theta-b_i)}}$$

- $a_i$ — **discrimination**: how steeply the curve rises around $b_i$
- $b_i$ — **difficulty**: the ability level at the curve's midpoint
- $c_i$ — **guessing (lower asymptote)**: minimum success probability for very low ability
- $d_i$ — **inattention (upper asymptote)**: maximum success probability for very high ability
- $D$ — a scaling constant (1, or 1.702 to make the logistic curve closely approximate the normal ogive)

Setting $c_i=0,\,d_i=1$ collapses this to the familiar **2PL** model; additionally fixing $a_i$ to a common value across items gives the **1PL/Rasch** model. `cat_eap.R`'s `Pi()` function implements the full 4PL and its first derivative $P_i'(\theta)$ (needed for information-based item selection, though not used further in this vignette).

### Likelihood of a response pattern

Assuming **local independence** (conditional on $\theta$, responses to different items are independent), the likelihood of a full 0/1 response pattern $\mathbf{x}=(x_1,\ldots,x_n)$ is

$$L(\theta) = \prod_{i=1}^{n} P_i(\theta)^{x_i}\,\big(1-P_i(\theta)\big)^{1-x_i}$$

### EAP: a Bayesian point estimate

Maximum Likelihood estimation fails outright for an all-correct or all-incorrect pattern — the likelihood keeps increasing toward $\theta=\pm\infty$ with no interior maximum. EAP sidesteps this by placing a **prior** distribution $\pi(\theta)$ on ability (by default, standard normal) and reporting the **mean of the posterior distribution**:

$$\hat\theta_{EAP} = E[\theta \mid \mathbf{x}] = \frac{\displaystyle\int \theta\,\pi(\theta)\,L(\theta)\,d\theta}{\displaystyle\int \pi(\theta)\,L(\theta)\,d\theta}$$

and its precision as the **posterior standard deviation**:

$$SE(\hat\theta_{EAP}) = \sqrt{\frac{\displaystyle\int (\theta-\hat\theta_{EAP})^2\,\pi(\theta)\,L(\theta)\,d\theta}{\displaystyle\int \pi(\theta)\,L(\theta)\,d\theta}}$$

Both integrals are always finite because the normal prior decays to zero in the tails, regardless of how extreme the likelihood is — this is exactly what makes EAP well-behaved for perfect response patterns.

### Numerical integration

Neither integral above has a closed-form solution, so `cat_eap.R` evaluates the integrand at a grid of **quadrature points** (`nqp`, default 33, evenly spaced between `lower` and `upper`, default $\pm4$) and integrates numerically with the **trapezoidal rule**: the area under the curve between two adjacent grid points is approximated as a trapezoid — width times the average of the two heights. `integrate_cat()` implements exactly this, summing the trapezoids across the whole grid.

---

## Validating each piece against `catR`

### Normal density

```{r density_check}
v <- seq(-3, 3, by = 0.1)
comparison <- data.frame(x = v, density_function = density_function(v), dnorm = dnorm(v))
head(comparison)
all.equal(comparison$density_function, comparison$dnorm)
```

```{r density_plot,dev='png'}
ggplot(comparison, aes(x = x)) +
  geom_line(aes(y = density_function, color = "density_function()"), linewidth = 1) +
  geom_point(aes(y = dnorm, color = "dnorm()"), shape = 1) +
  labs(title = "Hand-written normal density vs. dnorm()", x = expression(theta), y = "density", color = NULL) +
  theme_bw(base_size = 12)
```

The hand-written implementation overlays `dnorm()` exactly — as expected, since it's just the textbook normal density formula written out longhand.

### Numerical integration

```{r integration_check}
x <- seq(from = -3, to = 3, length = 33)
y <- round(exp(x), 2)
c(integrate_cat = integrate_cat(x, y), integrate.catR = catR::integrate.catR(x, y))
```

Both approximate $\int_{-3}^{3} e^x\,dx$ via the same trapezoidal grid and agree exactly.

### Response probability curves (4PL)

```{r pi_check}
bank <- matrix(c(0.7521, 0.8083, 1.1857, 0.5481, 0.5695,
                 -1.5521, -0.9083, 0.1857, 0.5481, 1.5695,
                 0, 0, 0, 0, 0,
                 1, 1, 1, 1, 1),
               nrow = 5, dimnames = list(1:5, c("a", "b", "c", "d")))

data.frame(mine = Pi(theta = 0, bank)$Pi, catR = catR::Pi(th = 0, bank)$Pi)
```

```{r pi_plot,dev='png'}
theta_grid <- seq(-6, 6, by = 0.1)
prob_curves <- sapply(theta_grid, function(th) Pi(theta = th, bank)$Pi)
df_curves <- data.frame(theta = theta_grid, t(prob_curves))
names(df_curves)[-1] <- paste0("item_", 1:5)
df_curves_long <- reshape2::melt(df_curves, id.vars = "theta", variable.name = "item", value.name = "P")

ggplot(df_curves_long, aes(x = theta, y = P, color = item)) +
  geom_line(linewidth = 1) +
  labs(title = "Item response curves (5-item 2PL bank)", x = expression(theta), y = expression(P(theta)), color = NULL) +
  theme_bw(base_size = 12)
```

Each curve is one item's probability of a correct response as a function of ability — steeper curves (higher $a$) discriminate more sharply around their difficulty $b$; curves shifted right have higher difficulty.

### EAP ability estimate

```{r eap_est_check}
response_1 <- c(1, 0, 0, 0, 0)
c(mine = eap_est(bank, response_1), catR = catR::eapEst(bank, response_1))
```

### EAP standard error

```{r eap_se_check}
c(mine = eap_se(0, bank, response_1), catR = catR::eapSem(0, bank, response_1))
```

Both the ability estimate and its standard error match `catR` exactly across the quadrature grid.

---

## Worked example: more correct responses shift and sharpen the estimate

Reproducing "Example 1" from the script: a fixed 5-item bank, scored against response patterns with an increasing number of correct answers (0 through 5 correct, always the *easiest* items first).

```{r worked_example}
a <- c(0.7521, 0.8083, 1.1857, 0.5481, 0.5695)
b <- c(-1.5521, -0.9083, 0.1857, 0.5481, 1.5695)
c <- c(0, 0, 0, 0, 0)
d <- c(1, 1, 1, 1, 1)
bank <- matrix(c(a, b, c, d), nrow = 5, dimnames = list(1:5, c("a", "b", "c", "d")))

responses <- list(
  c(1, 0, 0, 0, 0),
  c(1, 1, 0, 0, 0),
  c(1, 1, 1, 0, 0),
  c(1, 1, 1, 1, 0),
  c(1, 1, 1, 1, 1)
)

results <- data.frame(
  n_correct = sapply(responses, sum),
  theta_EAP = sapply(responses, function(x) eap_est(bank, x)),
  SE        = NA
)
results$SE <- mapply(function(x, th) eap_se(th, bank, x), responses, results$theta_EAP)
results
```

```{r worked_example_plot,dev='png'}
ggplot(results, aes(x = n_correct, y = theta_EAP)) +
  geom_ribbon(aes(ymin = theta_EAP - SE, ymax = theta_EAP + SE), alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 2, color = "steelblue") +
  labs(title = "EAP estimate (± 1 SE) vs. number of items correct",
       x = "Number of items correct (out of 5)", y = expression(hat(theta)[EAP])) +
  theme_bw(base_size = 12)
```

The estimate climbs monotonically with the number of correct responses, as expected. Notice it never reaches the extremes of the integration range ($\pm4$) even for the all-correct or all-incorrect pattern — the normal prior pulls the posterior mean back toward 0, which is exactly the regularizing behavior that makes EAP preferable to Maximum Likelihood at the edges of a response pattern.

## Worked example: how standard error depends on both $\theta$ and the response pattern

Reproducing the three SE curves from `eap_se()`'s documentation: the same 5-item bank, scored at every $\theta$ from $-3$ to $3$, under three different response patterns.

```{r se_curves}
bank_se <- matrix(c(0.7521, 0.8083, 1.1857, 0.5481, 0.5695,
                    -1.5521, -0.9083, 0.1857, 0.5481, 1.5695,
                    0, 0, 0, 0, 0,
                    1, 1, 1, 1, 1),
                  nrow = 5, dimnames = list(1:5, c("a", "b", "c", "d")))

theta_grid <- seq(-3, 3, by = 0.1)
patterns <- list(
  "all incorrect"    = c(0, 0, 0, 0, 0),
  "all correct"      = c(1, 1, 1, 1, 1),
  "2 of 5 correct"   = c(1, 0, 1, 0, 0)
)

se_df <- do.call(rbind, lapply(names(patterns), function(nm) {
  data.frame(theta = theta_grid,
             SE = sapply(theta_grid, function(th) eap_se(theta = th, bank = bank_se, x = patterns[[nm]])),
             pattern = nm)
}))
```

```{r se_curves_plot,dev='png'}
ggplot(se_df, aes(x = theta, y = SE, color = pattern)) +
  geom_line(linewidth = 1) +
  labs(title = "EAP standard error vs. assumed theta, by response pattern",
       x = expression(theta), y = "standard error", color = "response pattern") +
  theme_bw(base_size = 12)
```

The "2 of 5 correct" pattern reaches its lowest SE near $\theta=0$, where the item bank is best targeted for a mixed response pattern. The all-correct and all-incorrect curves instead keep *decreasing* toward the ends of the range they're evaluated over — consistent with those patterns being most consistent with an extreme ability level, even though the EAP point estimate itself (per the previous section) is pulled back toward 0 by the prior. This is a useful reminder that `eap_se()`'s argument `theta` is an arbitrary evaluation point for the posterior SD, not necessarily `eap_est()`'s own output — in practice you always pair them, as the earlier `eap_se(eap_est(bank, x), bank, x)` calls do.

---

## Conclusion

Every one of `cat_eap.R`'s five functions reproduces its `catR` counterpart exactly: a hand-rolled normal density, a trapezoidal integrator, 4PL response probabilities, and the two integrals — posterior mean and posterior SD — that together define EAP scoring. Seeing the full pipeline in ~150 lines of base R, with no hidden package internals, makes it much easier to reason about *why* EAP behaves the way it does: the prior is what keeps ability estimates finite and standard errors sane even for a perfect or a zero score, at the cost of a small, deliberate bias toward the prior mean.

---

*Rendered with R `r getRversion()` · validated against catR `r as.character(packageVersion("catR"))`*
