---
title: "MST"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM MST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}

directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))

```

```{r library,message=FALSE,warning=FALSE}
# install.packages("https://cran.r-project.org/src/contrib/Archive/mstR/mstR_1.2.tar.gz")
source(paste0(directory, "code_rwf.R"))
library(mstR)
library(catR)
```

## Description

This vignette demonstrates Multi-Stage Testing (MST) using the `mstR` package alongside helper functions from `rwf`. MST is an adaptive testing methodology that sits between traditional linear tests and fully adaptive Computerised Adaptive Testing (CAT). Instead of selecting items one at a time, MST groups items into **modules** and selects the next module based on the examinee's performance on the previous one.

A typical MST design has multiple **stages** (e.g., stage 1 → stage 2 → stage 3) and multiple **panels** (parallel test forms). Within each stage, an examinee is routed to an easy, medium, or hard module depending on their estimated ability (θ). This balances measurement precision with practical constraints such as content balancing and test security.

Key concepts used throughout this vignette:

- **Item bank**: a pool of items described by IRT parameters (here, the 2-Parameter Logistic model — discrimination `a` and difficulty `b`).
- **Modules**: subsets of items from the bank assigned to specific positions in the MST design.
- **Transition matrix**: defines which modules can follow which — routing rules between stages.
- **Cutoff matrix**: θ or raw-score thresholds that determine which module an examinee is routed to next.
- **θ (theta)**: the latent ability estimate on the standard normal scale.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Basic Examples

This section constructs a small 1-3-2 MST design by hand and demonstrates the two core routing functions: `startModule` and `nextModule`.

**Item bank construction**: Six groups of items are generated using `genDichoMatrix` under the 2PL model. The difficulty priors (`bPrior`) are shifted so that items in different modules cover different regions of the ability scale — negative difficulties for easy modules and positive difficulties for hard modules.

**Module matrix** (`modules`): A 47×6 binary matrix assigning each of the 47 items to one of six modules. Each column is a module; a `1` indicates that item belongs to that module. `colSums(modules)` confirms the expected module sizes (8, 6, 6, 9, 9, 9).

**Transition matrix** (`trans`): A 6×6 binary matrix defining allowed routing paths. Here, module 1 (stage 1) can route to modules 2 or 3 (stage 2); module 2 routes to modules 4 or 5; module 3 routes to modules 5 or 6. Module 5 is reachable from both stage-2 modules, acting as the medium-difficulty path.

**`startModule`**: Returns which module to administer first given the item bank, module structure, and transition matrix. The optional `fixModule` argument (commented out) would force a specific starting module regardless of the design.

**`thetaEst`**: Estimates θ from the response pattern `x` on the first 8 items using Maximum Likelihood (ML). This estimate is then used to determine the next module.

**`nextModule` with θ cutoffs** (`cut`): Routes to the next module based on whether the ML θ estimate falls below or above the defined thresholds (e.g., θ < 0 → module 2; θ ≥ 0 → module 3). Two calls show routing for the estimated θ and for a fixed θ of −0.2.

**`nextModule` with score cutoffs** (`cut.score`): Routes based on raw number-correct scores rather than θ estimates. This is useful when IRT scoring is not available or when content-balancing rules are score-based.

```{r excamples0,message=FALSE,warning=FALSE,results='hide',dev='png'}
it.MST <- rbind(genDichoMatrix(8, model = "2PL"), 
                genDichoMatrix(6, model = "2PL",bPrior = c("norm", -1, 1)),
                genDichoMatrix(6, model = "2PL",bPrior = c("norm", 1, 1)),
                genDichoMatrix(9, model = "2PL",bPrior = c("norm", -2, 1)),
                genDichoMatrix(9, model = "2PL",bPrior = c("norm", 0, 1)),
                genDichoMatrix(9, model = "2PL",bPrior = c("norm", 2, 1)))
it.MST <- as.matrix(it.MST)

modules <- matrix(0, 47, 6)
modules[1:8, 1] <- 1
modules[9:14, 2] <- 1
modules[15:20, 3] <- 1
modules[21:29, 4] <- 1
modules[30:38, 5] <- 1
modules[39:47, 6] <- 1

colSums(modules)

trans <- matrix(0, 6, 6)
trans[1, 2:3] <- 1
trans[2, 4:5] <- 1
trans[3, 5:6] <- 1

startModule(itemBank = it.MST, modules = modules, transMatrix = trans)
# startModule(itemBank = it.MST, modules = modules, transMatrix = trans, fixModule = 3)

x <- c(1, 1, 1, 1, 0, 0, 0, 0)
th<-mstR::thetaEst(it.MST[1:8, ], x, method = "ML")

cut <- rbind(c(2, 3,  0),
             c(4, 5, -1),
             c(5, 6,  1))

nextModule(itemBank = it.MST, modules = modules,transMatrix = trans, current.module = 1, out = 1:8, cutoff = cut, theta = th)
nextModule(itemBank = it.MST, modules = modules, transMatrix = trans, current.module = 1, out = 1:8, cutoff = cut, theta = -0.2)

cut.score <- rbind(c(2, 3,  5),
                   c(4, 5,  5),
                   c(5, 6, 10))

nextModule(itemBank = it.MST, modules = modules, transMatrix = trans, current.module = 3, out = c(1:8, 15:20), cutoff = cut.score,theta = th)

```

## Simulation Examples

This section runs full MST simulations using `randomMST` and compares MST performance against CAT using `randomCAT`. A 100-item bank with five 2PL items is used, assembled into six modules of 15 items each across a 1-2-2-1 branching design.

**Item bank** (`it.2PL`): Five 2PL items with discrimination values from 0.8 to 1.2 and difficulties spanning −0.8 to 1.1, recycled to fill a 100-item bank. Parameters `c = 0` and `d = 1` indicate no guessing and no slipping.

**Module membership** (A1–F1): Six pre-defined sets of 15 item indices each, assembled into `module1`. A1 is the stage-1 routing module; B1/C1 are stage-2 easy/hard modules; D1/E1/F1 are the three stage-3 modules.

**Simulation 1 — MFI module selection** (`mst.ex1`): A single MST run at true θ = 0 using ML scoring and Maximum Fisher Information (MFI) for module selection. The two `Ii()` calls confirm the routing decision by comparing total Fisher information in module 2 vs. module 3 at the stage-1 θ estimate — the module with higher information should be selected. `plot(mst.ex1)` draws the routing path taken through the MST design.

**Simulation 2 — θ cutoff routing** (`mst.ex2`): Routes modules using θ thresholds (θ < 0 → easy; 0 ≤ θ ≤ 0.3 → medium; θ > 0.3 → hard). Run at true θ = 1 to verify that the design routes to the harder modules.

**Simulation 3 — score cutoff routing** (`mst.ex3`): Routes using raw number-correct score thresholds rather than θ estimates. The scoring method is set to `"score"` for both routing and final estimation — useful when IRT calibration is not available at administration time.

**ASB and RMSE across the ability range**: A large-scale simulation (1000 replications × 9 ability levels from −2 to +2, all routed through the `module1` design) estimates two accuracy metrics across the full latent trait range:

- **ASB (Average Signed Bias)**: mean difference between estimated and true θ — positive values indicate systematic overestimation, negative indicate underestimation.
- **RMSE (Root Mean Square Error)**: overall estimation accuracy combining both bias and variance. Lower values indicate a more precise test design.

**MST vs. CAT comparison**: The final block compares MST (BM scoring, MFI module selection) against CAT (45-item stopping rule, BM scoring, MFI item selection, randomesque = 5 for item exposure control) across 9000 simulated examinees spanning θ from −2 to +2. Both `thFinal` (final ability estimate) and `seFinal` (standard error of measurement) are stored in `RES` for comparison of precision between the two adaptive testing formats.

```{r mst_examples,message=FALSE,warning=FALSE,results='hide',dev='png'}
A1 <- c(7, 13, 14, 15, 22, 28, 34, 38, 51, 59, 70, 84, 85, 87, 89)
B1 <- c(4, 12, 19, 27, 29, 39, 43, 60, 69, 71, 72, 73, 75, 90, 97)
C1 <- c(32, 35, 36, 45, 52, 57, 61, 63, 66, 78, 81,82, 88, 92, 100)
D1 <- c(1, 2, 3, 6, 18, 21, 41, 48, 49, 50, 54, 56, 58, 65, 77)
E1 <- c(5, 16, 24, 25, 26, 37, 40, 44, 53, 68, 74, 80, 86, 91, 93)
F1 <- c(9, 11, 31, 46, 47, 55, 64, 67, 76, 79, 83, 94, 95, 98, 99)

a <- rep(c(0.8, 0.9, 1.0, 1.1, 1.2), length.out = 100)
b <- rep(c(0.0, -0.5, 1.1, -0.8, 0.7), length.out = 100)
it.2PL <- cbind(a, b, c = 0, d = 1)

module1 <- matrix(0, 100, 6)
module1[A1, 1] <- module1[B1, 2] <- module1[C1, 3] <- module1[D1, 4] <- module1[E1, 5] <- module1[F1, 6] <- 1

trans <- matrix(0, 6, 6)
trans[1, 2:3] <- trans[2, 4:5] <- trans[3, 5:6] <- 1

start <- list(theta = 0)

test <- list(method = "ML", moduleSelect = "MFI")
final <- list(method = "ML")
mst.ex1 <- randomMST(trueTheta = 0, itemBank = it.2PL, modules = module1, transMatrix = trans, genSeed = 1, start = start, test = test, final = final)

sum(Ii(0.204, it.2PL[module1[, 2] == 1, ])$Ii)
sum(Ii(0.204, it.2PL[module1[, 3] == 1, ])$Ii)

plot(mst.ex1, border.col = "grey", arrow.col = "grey")

cut <- rbind(c(2, 3,  0.0),
             c(4, 5, -0.3),
             c(5, 6,  0.3))

test2 <- list(method = "ML", cutoff = cut)
mst.ex2 <- randomMST(trueTheta = 1, itemBank = it.2PL, modules = module1, transMatrix = trans, genSeed = 1, start = start, test = test2, final = final)

cut2 <- rbind(c(2, 3,  7.5),
              c(4, 5,  9.5),
              c(5, 6, 20.5))

test3 <- list(method = "score", cutoff = cut2)
final3 <- list(method = "score")
mst.ex3 <- randomMST(trueTheta = 1, itemBank = it.2PL, modules = module1, transMatrix = trans, genSeed = 1, start = start, test = test3, final = final)
```

### Accuracy across the ability range

The block below runs 1000 replications at each of 9 true θ levels (from −2 to +2) through the `module1` design, then computes **ASB** and **RMSE** at every level from the resulting θ estimates.

```{r asb_rmse_sim,message=FALSE,warning=FALSE,results='hide'}
s <- seq(-2, 2, 0.5)
TH <- rep(s, each = 1000)
RES <- matrix(NA, length(TH), 1)
start <- list(theta = 0)
test <- list(method = "BM", moduleSelect = "MFI")
final <- list(method = "ML")

for (i in 1:length(TH)){
  prov <- randomMST(trueTheta = TH[i],
                    itemBank = it.2PL, modules = module1,
                    transMatrix = trans, genSeed = i, start = start,
                    test = test, final = final)
  RES[i, 1] <- prov$thFinal
}

res.ASB <- res.RMSE <- matrix(NA, length(s), 1)
ASB <- function(t, th) mean(t-th)
RMSE <- function(t, th) sqrt(mean((t-th)^2))

for (i in 1:length(s)){
  ind <- which(TH == s[i])
  res.ASB[i, 1] <- ASB(RES[ind, 1], s[i])
  res.RMSE[i, 1] <- RMSE(RES[ind, 1], s[i])
}
```

```{r asb_rmse_plot,dev='png'}
df_asb_rmse <- data.frame(theta = s, ASB = res.ASB[, 1], RMSE = res.RMSE[, 1])
print(df_asb_rmse)

df_asb_rmse_long <- reshape2::melt(df_asb_rmse, id.vars = "theta", variable.name = "Metric", value.name = "value")

ggplot(df_asb_rmse_long, aes(x = theta, y = value, color = Metric)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line() +
  geom_point() +
  labs(title = "MST accuracy across the ability range",
       x = "True theta", y = "Metric value", color = NULL) +
  theme_bw(base_size = 12)
```

ASB near zero at every θ indicates the design is unbiased across the ability range; a systematic upward or downward drift at the extremes would indicate the module design under- or over-estimates examinees at those levels. RMSE typically rises toward both ends of the θ range, since fewer items in the bank concentrate information near the middle of the scale (θ = 0).

### MST vs. CAT comparison

This block compares MST (BM scoring, MFI module selection) against CAT (45-item stopping rule, BM scoring, MFI item selection, `randomesque = 5` for item exposure control) across 9000 simulated examinees spanning θ from −2 to +2.

```{r cat_mst_sim,message=FALSE,warning=FALSE,results='hide'}
s <- seq(-2, 2, 0.5)
TH <- rep(s, each = 1000)
DATA <- genPattern(TH, it.2PL, seed = 1)
RES <- matrix(NA, length(TH), 4)

start.mst <- list(theta = 0)
start.cat <- list(theta = 0, randomesque = 5)
test.cat <- list(method = "BM", itemSelect = "MFI", randomesque = 5)
test.mst <- list(method = "BM", moduleSelect = "MFI")
stop <- list(rule = "length", thr = 45)
final <- list(method = "BM")

for (i in 1:length(TH)){
  pr <- randomCAT(trueTheta = TH[i],
                  itemBank = it.2PL, genSeed = i, start = start.cat,
                  test = test.cat, stop = stop, final = final)
  RES[i,1] <- pr$thFinal
  RES[i,2] <- pr$seFinal
  pr <- randomMST(trueTheta = TH[i],
                  itemBank = it.2PL, genSeed = i, modules = module1,
                  transMatrix = trans, start = start.mst,
                  test = test.mst, final = final)
  RES[i,3] <- pr$thFinal
  RES[i,4] <- pr$seFinal
}
```

```{r cat_mst_summary,dev='png'}
df_compare <- data.frame(
  theta     = TH,
  cat_theta = RES[, 1],
  cat_se    = RES[, 2],
  mst_theta = RES[, 3],
  mst_se    = RES[, 4]
)

summary_compare <- do.call(rbind, lapply(s, function(th) {
  ind <- which(TH == th)
  data.frame(
    theta    = th,
    cat_ASB  = ASB(df_compare$cat_theta[ind], th),
    cat_RMSE = RMSE(df_compare$cat_theta[ind], th),
    cat_SE   = mean(df_compare$cat_se[ind]),
    mst_ASB  = ASB(df_compare$mst_theta[ind], th),
    mst_RMSE = RMSE(df_compare$mst_theta[ind], th),
    mst_SE   = mean(df_compare$mst_se[ind])
  )
}))

print(summary_compare)

summary_long <- reshape2::melt(
  summary_compare[, c("theta", "cat_RMSE", "mst_RMSE", "cat_SE", "mst_SE")],
  id.vars = "theta", variable.name = "Metric", value.name = "value"
)

ggplot(summary_long, aes(x = theta, y = value, color = Metric)) +
  geom_line() +
  geom_point() +
  labs(title = "CAT vs. MST: RMSE and average SE across the ability range",
       x = "True theta", y = "Value", color = NULL) +
  theme_bw(base_size = 12)
```

CAT selects the single best item at every step, so it is expected to match or beat MST's RMSE and average SE at every θ level — MST instead trades some of that per-item precision for the practical advantages of pre-assembled, reviewable modules (content balancing, easier quality control, and no need for real-time item-level adaptation).

---

## Conclusion

This vignette worked through MST from the ground up: routing a single examinee by hand with `startModule`/`nextModule` (via both θ and raw-score cutoffs), running full simulated administrations with `randomMST` under three different routing rules, and evaluating a complete design's accuracy (ASB, RMSE) across the ability range. The final comparison against `catR`'s `randomCAT` quantifies the precision MST gives up relative to fully adaptive CAT in exchange for its practical advantages — fixed, reviewable modules assembled ahead of time rather than item-by-item adaptive selection.






















