---
title: "CAT Simulation"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{CAT Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, echo=FALSE, warning=FALSE, message=FALSE}

directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))

knitr::knit_hooks$set(cex = function(before, options, envir) {
  if (before) par(cex      = options$cex,
                  cex.main = options$cex,
                  cex.lab  = options$cex,
                  cex.axis = options$cex * 0.85)
})
knitr::opts_chunk$set(cex = 2)

```



# Simulating a CAT IRT using catR

This document is a practical exercise illustrating a Computerized Adaptive Testing (CAT) simulation. CAT is grounded in Item Response Theory (IRT) and requires prior estimation of item parameters. Its purpose is to deliver short, tailored questionnaires by selecting items from a pool in real time based on each respondent's estimated ability. For example, if a participant answers a difficult item correctly, the algorithm skips easier items and instead presents a harder one to narrow down the ability estimate more efficiently. As a result, participants do not need to respond to every item in the pool. This simulation uses an item pool of 200 items with a maximum of 40 items administered per respondent.
The `catR` package handles the simulation, while `foreach` and `doSNOW` are used for parallel computation.

```{r set-options,echo=TRUE,results="hide"}

library(catR)
library(foreach)
library(doSNOW)
library(ggplot2)
library(patchwork)

```

## Simulate item pool

This simulation uses the **2PL (two-parameter logistic) model**, which assigns two parameters to each item: a **discrimination parameter** (*a*), reflecting how well the item distinguishes between ability levels, and a **difficulty parameter** (*b*), reflecting how hard the item is. The `genDichoMatrix` function from `catR` generates item parameters for dichotomous (binary) items — items scored as either correct (1) or incorrect (0), such as multiple-choice questions.

The 200-item bank is generated with the following priors:

- **Parameter *a*** follows a normal distribution with mean 2 and SD 0.1, ensuring all discrimination values are positive.
- **Parameter *b*** follows a normal distribution with mean 0 and SD 4, producing items that span a wide range of difficulty — from very easy (large negative *b*) to very hard (large positive *b*).

A high *a* value means the item sharply differentiates between respondents near its difficulty level; a high *b* value means only respondents with high ability are likely to answer correctly.

`genDichoMatrix` also includes *c* (pseudo-guessing) and *d* (inattention) parameters. Since their values default to 0 and 1 respectively, they have no practical effect and the model reduces to a standard 2PL formulation.

```{r bank,echo=TRUE}
itemBank<-genDichoMatrix(items=200,model="2PL",aPrior=c("norm",2,.1),bPrior=c("norm",0,4))
head(itemBank)
summary(itemBank)
```

## Simulate item responses

`catR` provides the `genPattern` function, which generates simulated item responses for a given set of ability values and an item bank. Here, 1,000 respondents with abilities uniformly spaced between −10 and +10 are simulated. Descriptive statistics are shown for the first 9 items to verify the generated response patterns.

```{r simulate_responses,echo=TRUE}
data<-genPattern(th=seq(-10,10,length=1000),it=itemBank,model=NULL,seed=1)
summary(data[,1:9])
```

## Estimate parameters using a third package

Treating the simulated response matrix as real participant data (one row per respondent), the `mirt` package is used to estimate item parameters and respondent ability scores under the assumption that all items were administered. These **full-information scores** serve as a benchmark against the CAT estimates produced in the next step.

```{r estimation,echo=TRUE,out.width=1000,results="hide"}
mirt_model<-mirt::mirt(data=data.frame(data),model=1,itemtype="2PL",technical=list(theta_lim=c(-6,6),NCYCLES=10000))
```

## Simulate CAT administration and scoring

Each respondent is administered a CAT drawn from the 200-item pool, with a maximum of 40 items. The simulation is configured through four control lists:

- **`start`** — defines first-item selection. The initial item is chosen to have a difficulty closest to θ = 0 (`startSelect = "bOpt"`), a neutral starting point before any ability information is available.
- **`test`** — controls item selection and scoring during the adaptive phase. Ability is estimated using Bayesian Modal (BM) estimation (Birnbaum, 1969). Although EAP is generally the recommended estimator for unidimensional IRT, BM was found to yield scores highly correlated with the `mirt` EAP benchmark in this setup. Items are selected by `bOpt`, with a `randomesque = 5` parameter introducing slight randomisation by sampling from the 5 most informative candidates. The estimation range is set to [−20, 20] to prevent ability estimates from collapsing at the boundaries.
- **`final`** — specifies the scoring method applied at test termination (BM, same range).
- **`stop`** — defines termination criteria. The test ends when either 40 items have been administered or the standard error of the provisional ability estimate drops below 0.3 — whichever occurs first.

```{r simulation,echo=TRUE,fig.align="center"}
start<-list(theta=0,nrItems=1,startSelect="bOpt")
test<-list(method="BM",itemSelect="bOpt",randomesque=5,range=c(-20,20))
final<-list(method="BM",range=c(-20,20))
stop<-list(rule=c("length","precision"),thr=c(40,0.3))
cl<-makeCluster(parallel::detectCores())
registerDoSNOW(cl)
result<-foreach(participant=1:nrow(data),.combine=rbind,.packages=c("catR")) %dopar% {
  rc_result<-randomCAT(itemBank=itemBank,
                       model=NULL,
                       responses=data[participant,],
                       start=start,
                       test=test,
                       stop=stop,
                       final=final)
  data.frame(questions=length(rc_result$thetaProv),
             est_theta=rc_result$thFinal,
             se=rc_result$seFinal)
  
}
stopCluster(cl)
result<-data.frame(result,mirt_scores=as.numeric(mirt::fscores(mirt_model)))
```

## Results

The results are summarised graphically across six panels:

- **Plot 1 — CAT vs. mirt scores:** Scatter plot of CAT ability estimates against `mirt` EAP scores. CAT respondents answered at most 40 items, while `mirt` scores are based on all 200 responses. High agreement between the two validates the adaptive procedure.
- **Plot 2 — Items administered vs. CAT theta:** Respondents with ability estimates far from 0 (extreme thetas) tend to require more items before the precision criterion is met, because test information is lower in the tails of the ability distribution.
- **Plot 3 — Items administered vs. SE:** Higher item counts are associated with elevated SE for two reasons: (1) the test was forced to stop at 40 items regardless of precision, and (2) extreme ability levels inherently have higher measurement error. Because extreme scores also require more items, both effects compound.
- **Plot 4 — CAT theta vs. SE:** Standard error increases at the extremes of the ability scale — a direct consequence of lower test information in the tails. Recall that SE is approximately the inverse square root of the information function.
- **Plot 5 — Distribution of CAT estimated scores.**
- **Plot 6 — Distribution of mirt estimated scores.**

```{r results0,echo=TRUE,fig.width=20,fig.height=15,out.width=1000,fig.align="center",dev='png',fig.retina=2}
p1 <- ggplot(result, aes(x = est_theta, y = mirt_scores)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  labs(x     = "CAT Theta",
       y     = "mirt Theta",
       title = paste("Pearson r", round(cor(result$est_theta, result$mirt_scores), 4))
  ) +
  theme_bw(base_size=20)

p2 <- ggplot(result, aes(x = questions, y = est_theta)) +
  geom_point(alpha = 0.5) +
  labs(x = "Questions asked", y = "CAT Theta") +
  theme_bw(base_size=20)

p3 <- ggplot(result, aes(x = questions, y = se)) +
  geom_point(alpha = 0.5) +
  labs(x = "Questions asked", y = "SE") +
  theme_bw(base_size=20)

p4 <- ggplot(result, aes(x = est_theta, y = se)) +
  geom_point(alpha = 0.5) +
  labs(x = "CAT Theta", y = "SE") +
  theme_bw(base_size=20)

p5 <- ggplot(result, aes(x = round(est_theta, 0))) +
  geom_bar() +
  labs(x = "CAT Theta", y = "Frequency") +
  theme_bw(base_size=20)

p6 <- ggplot(result, aes(x = round(mirt_scores, 0))) +
  geom_bar() +
  labs(x = "mirt Theta", y = "Frequency") +
  theme_bw(base_size=20)

(p1 | p2) / (p3 | p4) / (p5 | p6) +
  plot_annotation(theme = theme(plot.background = element_rect(fill = "white")))
```

## Conclusion

The distribution of estimated scores is approximately uniform, which is expected: the response-generation function sampled ability values evenly across the theta range, so all levels are equally represented. In a real-world scenario, ability scores tend to follow a roughly normal distribution, with most respondents clustered near the region of highest test information.

`catR` also provides `simulateRespondents`, a convenience function that runs a complete CAT simulation and computes summary diagnostics — bias, RMSE, and mean SE — as a function of true ability.

```{r results1,echo=TRUE,fig.width=20,fig.height=15,results="hide",out.width=1000,fig.align="center",dev='png',fig.retina=2}
result_simulate_respondents<-simulateRespondents(thetas=seq(-6,6,.01),itemBank=itemBank,genSeed=NULL,start=start,test=test,stop=stop,final=final)
result_simulate_respondents
plot(result_simulate_respondents)

```

