---
title: "Regularized Regression: Ridge, Heteroscedastic Ridge, Lasso, and Elastic Net"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))

set.seed(123)
library(glmnet)
library(dplyr)
library(psych)
```

# What problem is this solving?

Ordinary least squares (OLS) picks coefficients $\beta$ that minimize the residual
sum of squares (RSS): $\sum(y_i - X_i\beta)^2$. That works fine until predictors
are correlated with each other (multicollinearity) or there are many predictors
relative to observations — then OLS coefficients become unstable: tiny changes
in the data can swing them wildly, and the model overfits.

**Regularized regression** fixes this by adding a penalty on the size of the
coefficients to the loss function, so the optimizer is discouraged from letting
any coefficient get too large. Which penalty you choose gives you a different
member of the same family:

| Method      | Penalty added to RSS                  | `glmnet` `alpha` | Effect                                   |
|-------------|----------------------------------------|:---:|-------------------------------------------|
| Ridge       | $\lambda \sum \beta_j^2$ (L2)          | 0   | Shrinks coefficients toward zero, never exactly zero |
| Lasso       | $\lambda \sum \lvert\beta_j\rvert$ (L1) | 1   | Shrinks *and* can zero coefficients out (variable selection) |
| Elastic Net | a weighted mix of L1 and L2            | 0–1 | Compromise: some sparsity, handles correlated predictors better than pure lasso |

All of these introduce one new tuning parameter, $\lambda$ (how strong the
penalty is), which has to be chosen — that choice is a running theme through
this script.

The dataset throughout is the built-in `mtcars`: predicting `mpg` (miles per
gallon) from the other car characteristics (cylinders, weight, horsepower, etc.),
several of which are strongly correlated (e.g. `wt` and `disp`) — exactly the
setting where ridge/lasso earn their keep over plain `lm()`.

```{r data-setup}
# y is centered (mean-subtracted); X will be standardized inside glmnet itself
y <- as.matrix(scale(mtcars$mpg, center = TRUE, scale = FALSE))
X <- as.matrix(mtcars[, !names(mtcars) %in% "mpg"])
dim(X)
colnames(X)
```

# Ridge regression

## Fitting ridge and choosing $\lambda$ by cross-validation

`glmnet(X, y, alpha = 0, ...)` — `alpha = 0` is what tells `glmnet` "this is ridge,
not lasso." The function needs a value (or grid) of `lambda` to try. Rather than
guess one, `cv.glmnet()` performs k-fold cross-validation: it fits ridge on each
grid value of $\lambda$, holds out folds to measure prediction error, and reports
which $\lambda$ generalizes best (`lambda.min`).

```{r ridge-cv}
lambdas_to_try <- 10^seq(-3, 5, length.out = 100)

ridge_cv <- cv.glmnet(X, y, alpha = 0, lambda = lambdas_to_try,
                       standardize = TRUE, nfolds = 10)
plot(ridge_cv)
```

The plot shows cross-validated error (y-axis) against $\log(\lambda)$ (x-axis) —
a U-shape is typical: too small a $\lambda$ overfits, too large underfits, and
the dashed vertical line marks the minimum.

```{r ridge-cv-fit}
lambda_cv <- ridge_cv$lambda.min
lambda_cv

model_cv <- glmnet(X, y, alpha = 0, lambda = lambda_cv, standardize = TRUE)
y_hat_cv <- predict(model_cv, X)
ssr_cv <- t(y - y_hat_cv) %*% (y - y_hat_cv)
rsq_ridge_cv <- cor(y, y_hat_cv)^2
rsq_ridge_cv
```

## Choosing $\lambda$ by information criteria (AIC/BIC) instead

Cross-validation isn't the only way to pick $\lambda$. This section computes AIC
and BIC directly for every candidate $\lambda$, as an alternative to CV.

The subtlety: AIC/BIC need a "degrees of freedom" value, and in ridge regression
that isn't just the number of predictors (since coefficients are shrunk, not
freely estimated) — it's the trace of the **ridge hat matrix**
$H = X(X^\top X + \lambda I)^{-1}X^\top$, which continuously decreases from
$p$ (at $\lambda=0$, same as OLS) toward 0 as $\lambda \to \infty$.

```{r ridge-information-criteria}
X_scaled <- scale(X)
aic <- bic <- c()

for (lambda in seq(lambdas_to_try)) {
  model <- glmnet(X, y, alpha = 0, lambda = lambdas_to_try[lambda], standardize = TRUE)
  betas <- as.vector((as.matrix(coef(model))[-1, ]))
  resid <- y - (X_scaled %*% betas)

  # Ridge hat matrix and its trace = effective degrees of freedom
  ld <- lambdas_to_try[lambda] * diag(ncol(X_scaled))
  H <- X_scaled %*% solve(t(X_scaled) %*% X_scaled + ld) %*% t(X_scaled)
  df <- tr(H)  # tr() comes from the psych package

  aic[lambda] <- nrow(X_scaled) * log(t(resid) %*% resid) + 2 * df
  bic[lambda] <- nrow(X_scaled) * log(t(resid) %*% resid) + 2 * df * log(nrow(X_scaled))
}

plot(log(lambdas_to_try), aic, col = "orange", type = "l", ylim = c(190, 260),
     ylab = "Information Criterion")
lines(log(lambdas_to_try), bic, col = "skyblue3")
legend("bottomright", lwd = 1, col = c("orange", "skyblue3"), legend = c("AIC", "BIC"))
```

BIC penalizes model complexity (degrees of freedom) more heavily than AIC once
you have more than a handful of observations, which is why it typically prefers
a larger $\lambda$ (a simpler, more shrunk model) than AIC does — visible here as
the skyblue curve's minimum sitting to the right of the orange curve's.

```{r ridge-aic-bic-fit}
lambda_aic <- lambdas_to_try[which.min(aic)]
lambda_bic <- lambdas_to_try[which.min(bic)]
c(lambda_aic = lambda_aic, lambda_bic = lambda_bic)

model_aic <- glmnet(X, y, alpha = 0, lambda = lambda_aic, standardize = TRUE)
y_hat_aic <- predict(model_aic, X)
rsq_ridge_aic <- cor(y, y_hat_aic)^2

model_bic <- glmnet(X, y, alpha = 0, lambda = lambda_bic, standardize = TRUE)
y_hat_bic <- predict(model_bic, X)
rsq_ridge_bic <- cor(y, y_hat_bic)^2

c(rsq_ridge_aic = rsq_ridge_aic, rsq_ridge_bic = rsq_ridge_bic)
```

## Visualizing the shrinkage itself

This is the most direct illustration of what ridge actually *does*. Each line is
one predictor's coefficient, traced across every value of $\lambda$ tried. Follow
any single line from left (small $\lambda$, close to the unpenalized OLS
coefficient) to right (large $\lambda$): it bends toward zero, but — unlike
lasso below — none of the lines ever *touch* zero and stay there.

```{r ridge-coefficient-path}
plot(glmnet(X, y, alpha = 0, lambda = lambdas_to_try, standardize = FALSE), xvar = "lambda")
legend("bottomright", lwd = 1, col = 1:6, legend = colnames(X), cex = 0.7)
```

# Heteroscedastic ridge: penalizing each coefficient differently

Plain ridge penalizes every coefficient equally ($\lambda \sum \beta_j^2$, same
$\lambda$ for every $j$). But if some predictors are inherently noisier than
others, treating them identically may not be ideal. Heteroscedastic ridge
instead gives each coefficient its own penalty weight $w_j$, derived from that
predictor's own simple (univariate) regression:

$$\text{minimize} \quad \sum(y_i - X_i\beta)^2 + \lambda \sum w_j \beta_j^2$$

Predictors whose univariate slope estimate is more uncertain (larger standard
error, hence larger $w_j$) get shrunk more; predictors we're more confident
about get shrunk less.

```{r hridge-weights}
# w_j = variance of the coefficient from a simple regression of y on X[,j] alone
weights <- sapply(seq(ncol(X)), function(predictor) {
  uni_model <- lm(y ~ X[, predictor])
  summary(uni_model)$coefficients[2, 2]^2  # squared standard error
})
names(weights) <- colnames(X)
weights
```

```{r hridge-fit}
hridge_loss <- function(betas) {
  sum((y - X %*% betas)^2) + lambda * sum(weights * betas^2)
}

hridge <- function(y, X, lambda, weights) {
  # Regular ridge gives a sensible starting point for the numerical optimizer
  model_init <- glmnet(X, y, alpha = 0, lambda = lambda, standardize = FALSE)
  betas_init <- as.vector(model_init$beta)

  coef <- optim(betas_init, hridge_loss)$par
  fitted <- X %*% coef
  rsq <- cor(y, fitted)^2
  names(coef) <- colnames(X)
  list(coef = coef, fitted = fitted, rsq = rsq)
}

hridge_model <- hridge(y, X, lambda = 0.001, weights = weights)
rsq_hridge_0001 <- hridge_model$rsq
hridge_model$coef
```

Choosing $\lambda$ here could use the same cross-validation or AIC/BIC strategies
shown above for plain ridge — not implemented here, but see
[momisc::hridge](https://github.com/MichalOleszak/momisc/blob/master/R/hridge.R)
for a fuller implementation.

# Lasso: L1 penalty, sparse coefficients

Switching `alpha` from 0 to 1 changes the penalty from L2 ($\sum\beta_j^2$) to
L1 ($\sum\lvert\beta_j\rvert$). The geometry of the L1 penalty (a diamond-shaped
constraint region vs. ridge's circular one) means its corners often land exactly
on an axis — i.e., some coefficients get shrunk *all the way* to zero. That
makes lasso a built-in variable-selection method, not just a shrinkage method.

```{r lasso-cv}
lasso_cv <- cv.glmnet(X, y, alpha = 1, lambda = lambdas_to_try, standardize = TRUE, nfolds = 10)
plot(lasso_cv)

lambda_cv_lasso <- lasso_cv$lambda.min
model_cv_lasso <- glmnet(X, y, alpha = 1, lambda = lambda_cv_lasso, standardize = TRUE)
y_hat_cv_lasso <- predict(model_cv_lasso, X)
rsq_lasso_cv <- cor(y, y_hat_cv_lasso)^2
rsq_lasso_cv
```

```{r lasso-coefficient-path}
res <- glmnet(X, y, alpha = 1, lambda = lambdas_to_try, standardize = FALSE)
plot(res, xvar = "lambda")
legend("bottomright", lwd = 1, col = 1:6, legend = colnames(X), cex = 0.7)
```

Compare this plot to ridge's coefficient path above: here, lines don't just bend
toward zero, they hit zero and stay there as $\lambda$ grows — each one dropping
out is lasso "deciding" that predictor isn't worth keeping at that penalty
strength.

## Comparing R² across methods so far

```{r rsq-comparison}
rsq <- cbind("R-squared" = c(rsq_ridge_cv, rsq_ridge_aic, rsq_ridge_bic,
                              rsq_hridge_0001, rsq_lasso_cv))
rownames(rsq) <- c("ridge cross-validated", "ridge AIC", "ridge BIC",
                    "hridge 0.001", "lasso cross-validated")
print(rsq)
```

These are all in-sample R² (predicting on the same `X` used to fit), so they're
useful for comparing the methods to each other here, but shouldn't be read as an
estimate of real-world predictive accuracy — that's what the cross-validation
step is doing internally when picking $\lambda$, just not reported as a final
held-out number here.

# Elastic net: mixing L1 and L2

Elastic net adds both penalties at once, weighted by `alpha` between 0 (pure
ridge) and 1 (pure lasso):

$$\lambda \left[ \alpha \sum\lvert\beta_j\rvert + (1-\alpha)\sum\beta_j^2 \right]$$

It's useful when you want lasso's variable-selection behavior but have groups of
correlated predictors — pure lasso tends to arbitrarily pick just one predictor
out of a correlated group and zero out the rest, while elastic net's ridge
component lets correlated predictors share the credit more smoothly.

Here, `caret::train()` searches over both `alpha` and `lambda` simultaneously
(`method = "glmnet"`), using repeated cross-validation to score each combination.

Note: the original script builds the training data with `cbind(y, X)` and then
fits `mpg ~ .`. That silently fails here — `y <- scale(mtcars$mpg, ...)` strips
the column name, so `cbind(y, X)` produces a column literally named `""` (`V1`
once coerced to a data frame), not `mpg`, and the formula can't find `mpg`. Naming
the column explicitly fixes it:

```{r elastic-net, results='hide'}
library(caret)
train_control <- trainControl(method = "repeatedcv", number = 5, repeats = 5,
                               search = "random", verboseIter = TRUE)

data_enet <- data.frame(mpg = as.vector(y), X)
elastic_net_model <- train(mpg ~ ., data = data_enet, method = "glmnet",
                            preProcess = c("center", "scale"), tuneLength = 25,
                            trControl = train_control)
```

```{r elastic-net-results}
elastic_net_model$bestTune

y_hat_enet <- predict(elastic_net_model, X)
rsq_enet <- cor(y, y_hat_enet)^2
rsq_enet
```

# Summary

- All four methods (ridge, heteroscedastic ridge, lasso, elastic net) are OLS
  plus a penalty on coefficient size; they differ only in *which* penalty.
- **Ridge** (L2) shrinks all coefficients smoothly toward zero — good for
  multicollinearity, keeps every predictor in the model.
- **Heteroscedastic ridge** lets each predictor be penalized by its own
  confidence (via per-coefficient weights) instead of one blanket penalty.
- **Lasso** (L1) shrinks *and* zeroes out coefficients — a built-in feature
  selector, at the cost of being less stable when predictors are correlated.
- **Elastic net** mixes both, tunable via `alpha`, and is often the practical
  default when you're not sure which of ridge or lasso suits your data.
- In every case, the real modeling decision isn't "ridge vs. lasso" alone — it's
  **how you choose $\lambda$** (cross-validation vs. information criteria), since
  that controls the whole bias/variance trade-off.
