---
title: "EXPLORE_ASSUMPTIONS"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{EXPLORE_ASSUMPTIONS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
library(ggplot2)
library(grid)
```

## Description

This shows the output of functions from the package rwf for checking the normality assumption and finding outliers. The functions draw normality diagnostic plots (histogram, density, boxplot, Q-Q plot), run a battery of eight normality tests, flag outliers with three detection rules, report the percentage of extreme z-scores, and replace outliers with missing values. Each section gives the formulas the function uses.

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/EXPLORE_ASSUMPTIONS.R)

## Data

Three simulated variables with 300 cases each: one normal, one right-skewed (exponential) and one with heavy tails (t distribution with 3 degrees of freedom). Five missing values are added to the normal variable.

```{r data}
set.seed(1)
df <- data.frame(normal = rnorm(300, mean = 50, sd = 10),
                 skewed = rexp(300, rate = 0.1),
                 heavy_tails = 50 + 10 * rt(300, df = 3))
df$normal[sample(300, 5)] <- NA
head(df)
```

## Notation

$x_1,\dots,x_n$ are the non-missing values of a variable and $x_{(1)}\le\dots\le x_{(n)}$ the same values sorted. $\bar{x}$ is the mean, $s$ the sample standard deviation, $M$ the median and $z_i=(x_i-\bar{x})/s$ the z-scores. $Q_p$ is the $p$ quantile (R's default, type 7), $Q_1=Q_{0.25}$, $Q_3=Q_{0.75}$ and $IQR=Q_3-Q_1$. $\Phi$ and $\varphi$ are the standard normal distribution and density functions and $\Phi^{-1}$ the normal quantile function. $F_n(x)=\frac{1}{n}\sum_{i=1}^{n}\mathbf{1}[x_i\le x]$ is the empirical distribution function.

## plot_normality_diagnostics

Draws four base-graphics panels for each numeric variable: histogram, density, boxplot and Q-Q plot. Variables with 2 or fewer values or zero variance are skipped. With `file`, the panels are also saved to a PDF.

**Histogram.** With the default `breaks = "Sturges"` the number of classes is (Sturges, 1926):

$$k=\left\lceil\log_2 n+1\right\rceil$$

**Density.** A Gaussian kernel density estimate with Silverman's rule-of-thumb bandwidth (Silverman, 1986):

$$\hat{f}(x)=\frac{1}{nh}\sum_{i=1}^{n}\varphi\left(\frac{x-x_i}{h}\right),\qquad h=0.9\min\left\{s,\ \frac{IQR}{1.34}\right\}n^{-1/5}$$

**Boxplot.** The box runs from the lower hinge $H_L$ to the upper hinge $H_U$ (Tukey's hinges, close to $Q_1$ and $Q_3$) with a line at the median; the whiskers reach the most extreme values inside the fences and values outside are drawn as points (Tukey, 1977):

$$\left[H_L-1.5\,(H_U-H_L),\ \ H_U+1.5\,(H_U-H_L)\right]$$

**Q-Q plot.** Each sorted value is plotted against the normal quantile of its plotting position (Wilk & Gnanadesikan, 1968), and the reference line passes through the first and third quartiles:

$$\left(\Phi^{-1}(p_i),\ x_{(i)}\right),\qquad p_i=\frac{i-a}{n+1-2a},\qquad a=\begin{cases}3/8 & n\le10\\ 1/2 & n>10\end{cases}$$

$$\text{slope}=\frac{Q_3-Q_1}{\Phi^{-1}(0.75)-\Phi^{-1}(0.25)},\qquad \text{intercept}=Q_1-\text{slope}\cdot\Phi^{-1}(0.25)$$

Points close to the line mean the data are close to normal; a curve means skew and an S shape means heavier or lighter tails than normal.

```{r plot_normality_diagnostics,fig.width=8,fig.height=8,results='hide'}
plot_normality_diagnostics(df = df, title = "Simulated data")
```

## plot_histogram

One `ggplot2` histogram per numeric variable, with the number of observations, mean, SD and median in the caption. `bins` equal-width bins cover the range of the data; `ggplot2` centres the first and last bins on the minimum and maximum, so the bin width is:

$$w=\frac{x_{(n)}-x_{(1)}}{\text{bins}-1}$$

```{r plot_histogram,fig.width=10,fig.height=4}
invisible(plot_multiplot(plotlist = plot_histogram(df = df, bins = 30), cols = 3))
```

`xlims` fixes the x-axis range, `fill` and `color` set the bar colours:

```{r plot_histogram_options,fig.width=5,fig.height=4}
plot_histogram(df = df$skewed, bins = 20, xlims = c(0, 60), fill = "steelblue", color = "white")[[1]]
```

## plot_qq

One `ggplot2` Q-Q plot per numeric variable, with the same plotting positions $p_i$ and quartile reference line as in [plot_normality_diagnostics](#plot_normality_diagnostics).

```{r plot_qq,fig.width=10,fig.height=4}
invisible(plot_multiplot(plotlist = plot_qq(df = df[, c("skewed", "heavy_tails")]), cols = 2))
```

The right-skewed variable bends away from the line at the top; the heavy-tailed variable forms an S shape, with points below the line on the left and above it on the right.

## plot_boxplot

All numeric variables as side-by-side boxplots on one plot. `ggplot2` draws the box from $Q_1$ to $Q_3$ with a line at $M$; whiskers reach the most extreme values inside the fences and values outside are drawn as points:

$$\left[Q_1-1.5\,IQR,\ \ Q_3+1.5\,IQR\right]$$

```{r plot_boxplot,fig.width=8,fig.height=3}
plot_boxplot(df = df, title = "Simulated data")
```

## report_normality_tests

Runs eight normality tests on each variable. Each variable is first converted to z-scores, and variables with fewer than 8 or more than 4999 values are skipped. Small p values mean the data deviate from normality; with large samples even small, unimportant deviations give small p values.

```{r report_normality_tests}
report_normality_tests(df = df)
```

**Shapiro-Wilk** (Shapiro & Wilk, 1965). The weights $a_i$ come from the expected values of normal order statistics; R computes $W$ and its p value with Royston's (1995) approximation:

$$W=\frac{\left(\sum_{i=1}^{n}a_i\,x_{(i)}\right)^2}{\sum_{i=1}^{n}(x_i-\bar{x})^2}$$

**Anderson-Darling** (Anderson & Darling, 1952), with $F$ the distribution function of the null hypothesis:

$$A^2=-n-\frac{1}{n}\sum_{i=1}^{n}(2i-1)\left[\ln F\left(x_{(i)}\right)+\ln\left(1-F\left(x_{(n+1-i)}\right)\right)\right]$$

`report_normality_tests` calls `DescTools::AndersonDarlingTest` with its default null distribution, the uniform distribution on $[0,1]$, so $F(x)=x$ (see the `method` column). For a test of normality $F=\Phi$.

**Cramér-von Mises** (Stephens, 1974). The p value uses the modified statistic $W^2\left(1+\frac{0.5}{n}\right)$:

$$W^2=\frac{1}{12n}+\sum_{i=1}^{n}\left(\Phi\left(z_{(i)}\right)-\frac{2i-1}{2n}\right)^2$$

**Shapiro-Francia** (Shapiro & Francia, 1972). The squared correlation between the sorted data and approximate expected normal order statistics $m_i$; the p value uses Royston's (1993) approximation:

$$W'=\operatorname{cor}\left(x_{(i)},m_i\right)^2,\qquad m_i=\Phi^{-1}\left(\frac{i-3/8}{n+1/4}\right)$$

**Jarque-Bera.** `DescTools::JarqueBeraTest` computes the robust version (Gel & Gastwirth, 2008), which replaces the standard deviation by a robust measure of spread $J$. With $m_k=\frac{1}{n}\sum_{i=1}^{n}(x_i-\bar{x})^k$:

$$RJB=\frac{n}{6}\left(\frac{m_3}{J^3}\right)^2+\frac{n}{64}\left(\frac{m_4}{J^4}-3\right)^2,\qquad J=\sqrt{\frac{\pi}{2}}\,\frac{1}{n}\sum_{i=1}^{n}\left|x_i-M\right|$$

The p value comes from the $\chi^2$ distribution with 2 degrees of freedom. The classic test (Jarque & Bera, 1987) uses $JB=\frac{n}{6}\left(b_1^2+\frac{(b_2-3)^2}{4}\right)$ with $b_1=m_3/m_2^{3/2}$ and $b_2=m_4/m_2^2$.

**Lilliefors** (Lilliefors, 1967). The largest distance between the empirical distribution and the normal distribution with the mean and SD estimated from the data; the p value uses the Dallal & Wilkinson (1986) approximation, and Stephens' (1974) modified statistic when that approximation gives $p>0.1$:

$$D=\max_{i}\left\{\frac{i}{n}-\Phi\left(z_{(i)}\right),\ \Phi\left(z_{(i)}\right)-\frac{i-1}{n}\right\}$$

**Kolmogorov-Smirnov** (Kolmogorov, 1933; Smirnov, 1948). The same distance as Lilliefors:

$$D=\sup_x\left|F_n(x)-\Phi\left(\frac{x-\bar{x}}{s}\right)\right|$$

Its p value comes from the Kolmogorov distribution, which assumes the mean and SD are known in advance. Here they are estimated from the same data, which makes the p value too large; the Lilliefors test corrects for this, which is why the two rows have the same statistic but different p values.

**Pearson $\chi^2$** (Pearson, 1900). The range is split into $k$ classes that are equally likely under the fitted normal distribution, so each has expected count $E_j=N/k$, where $N$ is the number of non-missing values. With $O_j$ the observed count in class $j$:

$$P=\sum_{j=1}^{k}\frac{(O_j-E_j)^2}{E_j},\qquad k=\left\lceil2\,n^{2/5}\right\rceil,\qquad df=k-3$$

Here $n$ is the number of rows of `df`, including missing values (Moore, 1986). Two degrees of freedom are subtracted for the estimated mean and SD.

With `file`, the results are also written to a `.log` file and an `.xlsx` file in which p values $\le .05$ are highlighted.

## plot_outlier

One dot plot per numeric variable with outliers coloured and labelled by row name. Dashed lines show the centre and the two cut-offs. `method` selects the rule:

| `method` | Outlier when | Centre line |
|---|---|---|
| `"mean"` | $\left\lvert x_i-\bar{x}\right\rvert>2s$ | $\bar{x}$ |
| `"median"` | $\left\lvert x_i-M\right\rvert>2\,\dfrac{MAD}{0.6745}$, with $MAD=\operatorname{median}_i\left\lvert x_i-M\right\rvert$ | $M$ |
| `"boxplot"` | $x_i<Q_1-1.5\,IQR$ or $x_i>Q_3+1.5\,IQR$ | $M$ |

Dividing the MAD by 0.6745 $\left(=\Phi^{-1}(0.75)\right)$ makes it estimate the standard deviation for normal data (Leys et al., 2013). The mean and SD are pulled towards outliers themselves, so the median and boxplot rules are more robust. The plots below apply the three rules to the first 100 cases of the skewed variable.

```{r plot_outlier,fig.width=10,fig.height=8}
skewed <- df$skewed[1:100]
invisible(plot_multiplot(plotlist = c(plot_outlier(df = skewed, method = "mean", title = "mean"),
                                      plot_outlier(df = skewed, method = "median", title = "median"),
                                      plot_outlier(df = skewed, method = "boxplot", title = "boxplot")),
                         cols = 3))
```

## outlier_summary

Converts the variable to z-scores and reports the percentage of non-missing values beyond three cut-offs:

$$\text{percent}_c=100\cdot\frac{1}{n}\sum_{i=1}^{n}\mathbf{1}\left[\lvert z_i\rvert\ge c\right],\qquad c\in\{1.96,\ 2.58,\ 3.29\}$$

For normal data about $100\cdot2\left(1-\Phi(c)\right)$ percent are expected: 5%, 1% and 0.1%. Clearly larger percentages point to heavy tails or outliers (Tabachnick & Fidell, 2013).

```{r outlier_summary}
data.frame(t(sapply(df, outlier_summary)))
```

## remove_outliers

Replaces values outside the fences with `NA`. With `probs` $=(p_1,p_2)$:

$$x_i'=\begin{cases}\text{NA} & x_i<Q_{p_1}-1.5\,IQR\ \text{ or }\ x_i>Q_{p_2}+1.5\,IQR\\ x_i & \text{otherwise}\end{cases}$$

$IQR$ is always $Q_{0.75}-Q_{0.25}$, whatever `probs` is; with the default `probs = c(0.25, 0.75)` these are the boxplot fences (Tukey, 1977).

```{r remove_outliers}
cleaned <- data.frame(sapply(df, remove_outliers))
data.frame(missing_before = colSums(is.na(df)),
           missing_after = colSums(is.na(cleaned)))
```

## References

Anderson, T. W., & Darling, D. A. (1952). Asymptotic theory of certain "goodness of fit" criteria based on stochastic processes. *The Annals of Mathematical Statistics, 23*(2), 193–212. https://doi.org/10.1214/aoms/1177729437

Dallal, G. E., & Wilkinson, L. (1986). An analytic approximation to the distribution of Lilliefors's test statistic for normality. *The American Statistician, 40*(4), 294–296. https://doi.org/10.1080/00031305.1986.10475419

Gel, Y. R., & Gastwirth, J. L. (2008). A robust modification of the Jarque-Bera test of normality. *Economics Letters, 99*(1), 30–32. https://doi.org/10.1016/j.econlet.2007.05.022

Jarque, C. M., & Bera, A. K. (1987). A test for normality of observations and regression residuals. *International Statistical Review, 55*(2), 163–172. https://doi.org/10.2307/1403192

Kolmogorov, A. N. (1933). Sulla determinazione empirica di una legge di distribuzione. *Giornale dell'Istituto Italiano degli Attuari, 4*, 83–91.

Leys, C., Ley, C., Klein, O., Bernard, P., & Licata, L. (2013). Detecting outliers: Do not use standard deviation around the mean, use absolute deviation around the median. *Journal of Experimental Social Psychology, 49*(4), 764–766. https://doi.org/10.1016/j.jesp.2013.03.013

Lilliefors, H. W. (1967). On the Kolmogorov-Smirnov test for normality with mean and variance unknown. *Journal of the American Statistical Association, 62*(318), 399–402. https://doi.org/10.1080/01621459.1967.10482916

Moore, D. S. (1986). Tests of chi-squared type. In R. B. D'Agostino & M. A. Stephens (Eds.), *Goodness-of-fit techniques* (pp. 63–95). Marcel Dekker.

Pearson, K. (1900). On the criterion that a given system of deviations from the probable in the case of a correlated system of variables is such that it can be reasonably supposed to have arisen from random sampling. *The London, Edinburgh, and Dublin Philosophical Magazine and Journal of Science, 50*(302), 157–175. https://doi.org/10.1080/14786440009463897

Royston, P. (1993). A pocket-calculator algorithm for the Shapiro-Francia test for non-normality: An application to medicine. *Statistics in Medicine, 12*(2), 181–184. https://doi.org/10.1002/sim.4780120209

Royston, P. (1995). Remark AS R94: A remark on algorithm AS 181: The W-test for normality. *Journal of the Royal Statistical Society: Series C (Applied Statistics), 44*(4), 547–551. https://doi.org/10.2307/2986146

Shapiro, S. S., & Francia, R. S. (1972). An approximate analysis of variance test for normality. *Journal of the American Statistical Association, 67*(337), 215–216. https://doi.org/10.1080/01621459.1972.10481232

Shapiro, S. S., & Wilk, M. B. (1965). An analysis of variance test for normality (complete samples). *Biometrika, 52*(3–4), 591–611. https://doi.org/10.1093/biomet/52.3-4.591

Silverman, B. W. (1986). *Density estimation for statistics and data analysis*. Chapman & Hall.

Smirnov, N. (1948). Table for estimating the goodness of fit of empirical distributions. *The Annals of Mathematical Statistics, 19*(2), 279–281. https://doi.org/10.1214/aoms/1177730256

Stephens, M. A. (1974). EDF statistics for goodness of fit and some comparisons. *Journal of the American Statistical Association, 69*(347), 730–737. https://doi.org/10.1080/01621459.1974.10480196

Sturges, H. A. (1926). The choice of a class interval. *Journal of the American Statistical Association, 21*(153), 65–66. https://doi.org/10.1080/01621459.1926.10502161

Tabachnick, B. G., & Fidell, L. S. (2013). *Using multivariate statistics* (6th ed.). Pearson.

Tukey, J. W. (1977). *Exploratory data analysis*. Addison-Wesley.

Wilk, M. B., & Gnanadesikan, R. (1968). Probability plotting methods for the analysis of data. *Biometrika, 55*(1), 1–17. https://doi.org/10.1093/biomet/55.1.1
