---
title: "FUNCTIONS_RECODE"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{FUNCTIONS_RECODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
```

## Description

This shows the output of recoding functions from the package rwf. The functions stack a list of data frames into one data frame, reverse-score a vector (for example reverse-keyed Likert items), turn multiple-response answers into binary dummy columns, and collapse rare factor levels into a single "Other" level. Each section gives the rule the function applies.

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/FUNCTIONS_RECODE.R)

## Notation

$x_1,\dots,x_n$ are the values of the input vector and $\mathbf{1}[\cdot]$ is 1 when the condition is true and 0 otherwise. $|S|$ is the number of elements of a set $S$.

## flatten_list

Stacks the elements of a list $L=\{L_1,\dots,L_K\}$ into one data frame. Each element $L_k$ is converted to a data frame with $n_k$ rows, the data frames are placed under each other, and a `.id` column records the name of the element each row came from:

$$\text{rows}=\sum_{k=1}^{K}n_k,\qquad \texttt{.id}_i=\operatorname{name}(L_k)\ \text{ for every row } i \text{ that comes from } L_k$$

Columns with the same name are matched; a column missing from an element is filled with `NA`.

```{r flatten_list}
results <- list(sample_a = data.frame(mean = 10.2, sd = 1.1, n = 30),
                sample_b = data.frame(mean = 11.8, sd = 1.4, n = 25),
                sample_c = data.frame(mean = 9.7, sd = 0.9, n = 41))
flatten_list(results)
```

Vectors become one-column data frames:

```{r flatten_list_vectors}
flatten_list(list(group1 = 1:3, group2 = 4:5))
```

## swap

Reverse-scores a vector by mirroring its distinct observed values. With $v_{(1)}<v_{(2)}<\dots<v_{(m)}$ the $m$ distinct non-missing values in sorted order, each value is replaced by the value at the mirrored position:

$$x_i=v_{(r)}\ \Rightarrow\ x_i'=v_{(m+1-r)}$$

When the observed values are all the consecutive integers from $a$ to $b$, this is the usual reverse-scoring formula:

$$x_i'=a+b-x_i$$

Missing values stay missing, and the result has the same class as the input (numeric, integer or character).

```{r swap}
likert <- c(1, 2, 3, 4, 5, 5, 4, 1)
swap(likert)
all.equal(swap(likert), 1 + 5 - likert)
```

Only the **observed** values are mirrored, not the full scale. If a 1–5 item has no answers of 4, the observed values are 1, 2, 3, 5: 1 and 5 swap, 2 and 3 swap, and $a+b-x$ no longer applies:

```{r swap_observed}
likert_gap <- c(1, 2, 3, 5, 5)
swap(likert_gap)
6 - likert_gap
```

Numbers are sorted by value and text alphabetically, so text categories are mirrored in alphabetical order, not in their meaning:

```{r swap_text}
swap(c("low", "mid", "high", "low"))
```

## dummy_arrange

Turns multiple-response answers into binary dummy columns. Each element $i$ holds one or more responses separated by commas, giving a set $S_i$ (an empty element gives $S_i=\varnothing$). The columns are all responses given by anyone, sorted as text:

$$\mathcal{K}=\bigcup_{i=1}^{n}S_i,\qquad D_{ik}=\mathbf{1}\left[k\in S_i\right],\quad i=1,\dots,n,\quad k\in\mathcal{K}$$

Row sums give the number of responses each person chose, $\sum_{k}D_{ik}=|S_i|$, and column sums give how often each response was chosen. The function returns the 1s and 0s as text, so convert them to numbers before summing:

```{r dummy_arrange}
answers <- c("Email,Phone", "Phone", "Post,Email,Phone", "", "Email")
dummies <- dummy_arrange(answers)
dummies
dummies <- data.frame(lapply(dummies, as.numeric))
rowSums(dummies)
colSums(dummies)
```

Single responses work the same way. Because columns are sorted as text, numeric responses sort as `1`, `10`, `2`:

```{r dummy_arrange_numeric}
dummy_arrange(c("1,10", "2", "10,2"))
```

`generate_multiple_response_vector` from rwf simulates multiple-response data; `gsub` removes the spaces after the commas:

```{r dummy_arrange_generated}
set.seed(1)
responses <- gsub(" ", "", generate_multiple_response_vector(responses = c("Agree", "Hi", "All"), responded = 1:3, length = 6), fixed = TRUE)
responses
dummy_arrange(responses)
```

## drop_levels

Collapses rare levels of each factor into a single level `"Other"` and removes unused levels. With $n_\ell=\sum_{i=1}^{n}\mathbf{1}[x_i=\ell]$ the frequency of level $\ell$ and $c$ = `minimum_frequency`:

$$x_i'=\begin{cases}\text{"Other"} & n_{x_i}\le c\\ x_i & n_{x_i}>c\end{cases}$$

Levels with no observations ($n_\ell=0\le c$) are removed. All factor columns are processed, or only the columns in `factor_index`; other columns are left unchanged.

```{r drop_levels}
df <- data.frame(id = 1:12,
                 city = factor(c(rep("Athens", 6), rep("Patras", 3), rep("Volos", 2), "Chania"),
                               levels = c("Athens", "Patras", "Volos", "Chania", "Larissa")),
                 group = factor(rep(c("x", "y"), 6)))
table(df$city)
```

With $c=2$, Volos (2) and Chania (1) become "Other" and the empty level Larissa is removed:

```{r drop_levels_2}
table(drop_levels(df, factor_index = 2, minimum_frequency = 2)$city)
```

The comparison is $\le$, so with $c=3$ Patras (exactly 3) is collapsed too:

```{r drop_levels_3}
table(drop_levels(df, factor_index = 2, minimum_frequency = 3)$city)
```

Without `factor_index`, every factor is processed. With $c=6$ no level of either factor has more than 6 cases (Athens and both levels of `group` have exactly 6), so each factor collapses into a single "Other" level:

```{r drop_levels_all}
result <- drop_levels(df, minimum_frequency = 6)
table(result$city)
table(result$group)
```
