---
title: "FUNCTIONS TRAIN TEST"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{FUNCTIONS TRAIN TEST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
options(future.show.progress=FALSE)
options(width=1000)
knitr::opts_chunk$set(echo=TRUE)
hook_output<-function(x, options) {
  paste0('<pre class="r-output">',knitr::knit_print(x),'</pre>')
}
knitr::knit_hooks$set(output=hook_output)
```

```{r load,include=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description
This shows the output of useful functions for training and testing from rwf.  
Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

## Plot Receiver Operating Characteristic ROC Curve
```{r plot_roc,fig.width=8,fig.height=8,echo=TRUE,message=FALSE,warning=FALSE,results='hide',dev='png',fig.retina=2}
# Observed vector binary
observed<-round(abs(rnorm(100,m=0,sd=.5)))
# Predicted vector continuous
predicted<-abs(rnorm(100,m=0,sd=.5))
plot_roc(observed=observed,predicted=predicted)

# An example with a good prediction
df1<-data.frame(matrix(.999,ncol=2,nrow=2))
correlation_martix<-as.matrix(df1)
diag(correlation_martix)<-1
df1<-generate_correlation_matrix(correlation_martix,nrows=1000)
df1$X1<-ifelse(abs(df1$X1)<1,0,1)
df1$X2<-abs(df1$X2)
df1$X2<-(df1$X2-min(df1$X2))/(max(df1$X2)-min(df1$X2))
plot_roc(observed=round(abs(df1$X1),0),predicted=abs(df1$X2))
```

## Plot Confusion Matrix
```{r plot_confusion,fig.width=8,fig.height=8,dev='png',fig.retina=2}
plot_confusion(observed=c(1,2,3,1,2,3),predicted=c(1,2,3,1,2,3))
observed<-c(rep("male",10),rep("female",10),"male","male")
predicted<-c(rep("male",10),rep("female",10),"female","female")
plot_confusion(observed=observed,predicted=predicted)
```

## Plot Separability
This function generates a separability plot using ggplot2. It shows the density distribution of predicted probabilities for different observed categories. The plot helps to visualize how well the predicted probabilities separate the different observed categories.

```{r plot_separability,fig.width=8,fig.height=8,dev='png',fig.retina=2}
df1<-data.frame(matrix(.999,ncol=2,nrow=2))
correlation_martix<-as.matrix(df1)
diag(correlation_martix)<-1
df1<-generate_correlation_matrix(correlation_martix,nrows=1000)
df1$X1<-ifelse(abs(df1$X1)<1,0,1)
df1$X2<-abs(df1$X2)
df1$X2<-(df1$X2-min(df1$X2))/(max(df1$X2)-min(df1$X2))
plot_separability(observed=round(abs(df1$X1),0),predicted=abs(df1$X2))
```

## Plot Confusion Matrix Performance
This function evaluates the performance of a confusion matrix at different cut-off points. It iterates through a range of cut-off points, calculates the confusion matrix,and evaluates the proportion of correct classifications for each cut-off.

```{r plot_confusion_performance,fig.width=9,fig.height=8,dev='png',fig.retina=2}
df<-data.frame(matrix(.999,ncol=2,nrow=2))
correlation_martix<-as.matrix(df)
diag(correlation_martix)<-1
df<-generate_correlation_matrix(correlation_martix,nrows=1000)
df$X1<-ifelse(abs(df$X1)<1,0,1)
df$X2<-abs(df$X2)
df$X2<-(df$X2-min(df$X2))/(max(df$X2)-min(df$X2))
result_confusion_performance(observed=round(abs(df$X1),0),
                             predicted=abs(df$X2),
                             step=.01)
result_confusion_performance(observed=c(1,2,3,1,2,3),predicted=abs(rnorm(6,0,sd=.1)))
```

## KFOLD
Sample data for train test validation procedures

```{r kfold}
infert_formula<-as.formula(factor(case)~age+parity+education+spontaneous+induced)
result<-k_fold(infert,k=10,model_formula=infert_formula)
model_formula<-as.formula(mpg~cyl+disp+hp+drat+wt+qsec+vs+am+gear+carb)
result<-k_fold(mtcars,k=2,model_formula=model_formula)
result
```

## KSAMPLE
Sample data for train test validation procedures

```{r ksample}
infert_formula<-as.formula(factor(case)~age+parity+education+spontaneous+induced)
result<-k_sample(df=infert,k=10,model_formula=infert_formula)
model_formula<-as.formula(mpg~cyl+disp+hp+drat+wt+qsec+vs+am+gear+carb)
result<-k_sample(df=mtcars,k=10,model_formula=model_formula)
result
```

## Scaling
Scale all data between 0 and 1 

```{r scaling}
result<-recode_scale_dummy(infert)
head(result)
```

## Confusion Matrix
```{r confusion_classification}
confusion(observed=c(1,2,3,4,5,10),predicted=c(1,2,3,4,5,11))
confusion(observed=c(1,2,2,2,2),predicted=c(1,1,2,2,2))
```

## Proportion Accutate
```{r proportion_accurate}
proportion_accurate(observed=c(1,2,3,4,5,10),predicted=c(1,2,3,4,5,11))
```

## Confusion Matrix Percent
```{r confusion_percent}
observed<-factor(round(rnorm(10000,m=10,sd=1)))
predicted<-factor(round(rnorm(10000,m=10,sd=1)))
confusion_matrix_percent(observed,predicted)
```







