---
title: "GLM SEM"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM SEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette documents the Structural Equation Modeling (SEM) / Confirmatory Factor Analysis (CFA) functions from `working_functions/GLM_SEM.R`: a dependency-free (no `semPlot`) path diagram plotter, a full CFA report (fit indices, parameter estimates, modification indices, predicted scores), and a sample-size-planning simulation that tracks how model fit behaves as sample size grows.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### Confirmatory Factor Analysis

CFA specifies, in advance, which observed indicators load onto which latent factors (unlike EFA, which discovers the structure). For a single-indicator relationship,

$$x_i = \lambda_i \eta + \epsilon_i$$

where $\eta$ is the latent factor, $\lambda_i$ is the factor loading (how strongly indicator $i$ reflects the factor), and $\epsilon_i$ is measurement error. The model is fit by finding parameter values that best reproduce the observed covariance matrix among indicators.

### Model fit indices

- **$\chi^2$ test**: tests whether the model-implied covariance matrix differs significantly from the observed one. With large samples, this test is almost always significant even for good models, so it's rarely used alone.
- **CFI / TLI** (Comparative Fit Index / Tucker-Lewis Index): compare the fitted model to a null (no-correlation) baseline; conventionally $\geq 0.95$ indicates good fit.
- **RMSEA** (Root Mean Square Error of Approximation): a badness-of-fit index that penalizes model complexity; conventionally $\leq 0.06$ (sometimes $0.08$) indicates good fit, with an accompanying confidence interval.
- **SRMR** (Standardized Root Mean Square Residual): the average standardized difference between observed and model-implied correlations; conventionally $\leq 0.08$ indicates good fit.

### Modification indices

A modification index estimates how much $\chi^2$ would drop if a currently-fixed parameter (typically a residual covariance, or a cross-loading) were freed and re-estimated. Large indices flag places where the model doesn't match the data — but freeing parameters purely to improve fit, without substantive justification, risks overfitting to sample-specific noise rather than reflecting genuine structure.

### Sample size and power for SEM

SEM fit indices and standard errors are sample-size dependent: small samples can fail to reject a poorly-fitting model (low power) while very large samples can flag trivial misfit as statistically significant. `simulate_cfa_fit` directly visualizes how fit indices stabilize as sample size grows, supporting sample-size planning before data collection.

---

## Model Diagrams

`plot_cfa_gg` draws a path diagram directly in `ggplot2` (no `semPlot`/Rgraphviz system dependency), with latent variables as filled rectangles, observed variables as light rectangles, and edges labeled with loadings.

```{r cfa_model,results='hide'}
model_syntax <- "LATENT1=~X1+X2+X3
                 LATENT2=~X4+X5+X6"
df_cfa <- lavaan::simulateData(model = model_syntax, model.type = "cfa", return.type = "data.frame", sample.nobs = 300)
fit <- lavaan::cfa(model_syntax, data = df_cfa)
```

```{r plot_cfa_std,fig.width=8,fig.height=6}
plot_cfa_gg(fit, what = "std", layout = "tree")
```

```{r plot_cfa_est,fig.width=8,fig.height=6}
plot_cfa_gg(fit, what = "est", layout = "circle")
```

`what = "std"` shows standardized loadings (comparable across indicators regardless of their original scale); `what = "est"` shows raw, unstandardized estimates (in the original metric of each indicator). Three `layout` options (`tree`, `circle`, `spring`) control node placement via `igraph`.

`plot_cfa` is a batch wrapper that runs every combination of `layout` × `what` (9 diagrams total) in one call — useful for quickly picking the clearest arrangement for a report:

```{r plot_cfa_batch,results='hide'}
all_plots <- plot_cfa(fit)
```

```{r plot_cfa_batch_names}
names(all_plots)
```

## Full CFA Report

`report_cfa` bundles fit indices, R², parameter estimates (unstandardized and standardized), modification indices, sample covariance, model-implied predicted scores, and the path diagrams into one structured report.

```{r report_cfa,results='hide'}
result <- report_cfa(model = fit)
```

```{r report_cfa_output}
result$fit_indices
result$r_squared
head(result$parameters)
head(result$modification_indices[order(-result$modification_indices$mi), ])
```

- **`fit_indices`**: the full `lavaan` fit table (CFI, TLI, RMSEA + CI, SRMR, AIC, BIC, etc.).
- **`r_squared`**: proportion of each indicator's variance explained by its latent factor.
- **`parameters`**: every estimated parameter (loadings, variances, covariances) with SE, z, p, and confidence intervals.
- **`modification_indices`**: sorted here by size — the largest values are the parameters most likely to improve fit if freed (again, only worth freeing with a substantive rationale).

### Multi-group models

When the fitted model has more than one group, `report_cfa` additionally returns a `group` summary table (group labels, sample sizes per group, and total N) — omitted here since the example above is single-group.

## Sample Size Planning

`simulate_cfa_fit` refits a CFA model repeatedly across a range of sample sizes, in parallel, and returns fit indices at each size — either generating data from known population parameters (`model_sim`) or resampling from an observed dataset's correlation structure (`df`).

```{r simulate_cfa_fit,results='hide'}
model_sim <- "LATENT =~ 1*X1 + 0.5*X2 + 1.5*X3 + 1.5*X4 + X5"
model_est <- "LATENT =~ X1 + X2 + X3 + X4 + X5"
result_sim <- simulate_cfa_fit(model_sim = model_sim, model = model_est, minnobs = 50, maxnobs = 1000, stepping = 100)
```

```{r simulate_cfa_fit_plot,fig.width=12,fig.height=8}
plot_multiplot(plotlist = result_sim[[2]], cols = 3)
```

Each panel tracks one fit index against sample size. Indices like RMSEA and SRMR typically stabilize and tighten (less sample-to-sample variability) as $n$ grows — the point at which a given index's variability becomes acceptably small for your purposes is a data-driven answer to "how large a sample do I need?" for this specific model.

---

## Conclusion

`plot_cfa_gg`/`plot_cfa` visualize a fitted measurement model's structure without requiring `semPlot`'s system dependencies; `report_cfa` collects every quantity needed to write up a CFA (fit, parameters, R², modification indices) in one call; and `simulate_cfa_fit` turns "what sample size do I need" into an empirical, visualizable question rather than a rule of thumb.

---

*Rendered with R `r getRversion()` · lavaan `r as.character(packageVersion("lavaan"))`*
