---
title: "GLM SEM"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM SEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
library(ggplot2)
library(grid)

```

## Description

This shows the output of Structural Equation Modeling (SEM) and Confirmatory Factor Analysis (CFA) functions from the package rwf. The functions include a path diagram plotter that does not need `semPlot`, a full CFA report (fit indices, parameter estimates, modification indices, predicted scores), and a sample-size planning simulation that tracks how model fit changes as sample size grows.

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/GLM_SEM.R)

## Theoretical Background

### Confirmatory Factor Analysis

CFA specifies, in advance, which observed indicators load onto which latent factors (unlike EFA, which discovers the structure). For a single-indicator relationship,

$$x_i = \lambda_i \eta + \epsilon_i$$

where $\eta$ is the latent factor, $\lambda_i$ is the factor loading (how strongly indicator $i$ reflects the factor), and $\epsilon_i$ is measurement error. The model is fit by finding parameter values that best reproduce the observed covariance matrix among indicators.

### Model fit indices

Every index in the report is listed with its reference in [Fit indices in the report](#fit-indices-in-the-report).

### Modification indices

A modification index estimates how much $\chi^2$ would drop if a currently-fixed parameter (typically a residual covariance, or a cross-loading) were freed and re-estimated. Large indices flag places where the model doesn't match the data — but freeing parameters purely to improve fit, without substantive justification, risks overfitting to sample-specific noise rather than reflecting genuine structure.

### Sample size and power for SEM

SEM fit indices and standard errors are sample-size dependent: small samples can fail to reject a poorly-fitting model (low power) while very large samples can flag trivial misfit as statistically significant. `simulate_cfa_fit` directly visualizes how fit indices stabilize as sample size grows, supporting sample-size planning before data collection.

---

## Model Diagrams

`plot_cfa_gg` draws a path diagram directly in `ggplot2` (no `semPlot`/Rgraphviz system dependency), with latent variables as filled rectangles, observed variables as light rectangles, and edges labeled with loadings.

```{r cfa_model,results='hide'}
model_syntax <- "LATENT1=~X1+X2+X3
                 LATENT2=~X4+X5+X6"
df_cfa <- lavaan::simulateData(model = model_syntax, model.type = "cfa", return.type = "data.frame", sample.nobs = 300)
fit <- lavaan::cfa(model_syntax, data = df_cfa)
```

```{r plot_cfa_std,fig.width=8,fig.height=6}
plot_cfa_gg(fit, what = "std", layout = "tree")
```

```{r plot_cfa_est,fig.width=8,fig.height=6}
plot_cfa_gg(fit, what = "est", layout = "circle")
```

`what = "std"` shows standardized loadings (comparable across indicators regardless of their original scale); `what = "est"` shows raw, unstandardized estimates (in the original metric of each indicator). Three `layout` options (`tree`, `circle`, `spring`) control node placement via `igraph`.

### Equality constraints

`what = "eq"` labels each loading with its parameter label, so loadings constrained to be equal (sharing the same label) are easy to spot; loadings without a label show their estimate. It needs a model with labelled parameters. Here the loadings of X2 and X3 are constrained to be equal (`a`), and so are those of X5 and X6 (`b`):

```{r plot_cfa_eq,fig.width=8,fig.height=6}
model_eq <- "LATENT1=~X1+a*X2+a*X3
             LATENT2=~X4+b*X5+b*X6"
fit_eq <- lavaan::cfa(model_eq, data = df_cfa)
plot_cfa_gg(fit_eq, what = "eq", layout = "spring")
```

### Styling

`label_size` and `edge_label_size` set the text size of the node and path labels; `color_latent` and `color_observed` set the fill colour of the latent and observed nodes:

```{r plot_cfa_style,fig.width=8,fig.height=6}
plot_cfa_gg(fit, what = "std", layout = "tree",
            label_size = 4, edge_label_size = 3.5,
            color_latent = "#2c3e50", color_observed = "#ecf0f1")
```

### All layouts at once

`plot_cfa` is a batch wrapper that runs every combination of `layout` × `what` in one call and returns them as a named list — useful for quickly picking the clearest arrangement for a report. The three `eq` diagrams are only produced for models with labelled parameters, so `fit` gives 6 diagrams and `fit_eq` gives all 9:

```{r plot_cfa_batch,results='hide'}
all_plots <- plot_cfa(fit)
all_plots_eq <- plot_cfa(fit_eq)
```

```{r plot_cfa_batch_names}
names(all_plots)
names(all_plots_eq)
```

Any diagram can be taken from the list by name:

```{r plot_cfa_batch_one,fig.width=8,fig.height=6}
all_plots$spring_standard_estimates
```

## Full CFA Report

`report_cfa` bundles fit indices, R², parameter estimates (unstandardized and standardized), modification indices, sample covariance, model-implied predicted scores, and the path diagrams into one structured report.

```{r report_cfa,results='hide'}
result <- report_cfa(model = fit)
```

```{r report_cfa_output}
result$fit_indices
result$r_squared
result$parameters
result$modification_indices[order(-result$modification_indices$mi), ]
```

- **`fit_indices`**: the full `lavaan` fit table (CFI, TLI, RMSEA + CI, SRMR, AIC, BIC, etc.).
- **`r_squared`**: proportion of each indicator's variance explained by its latent factor.
- **`parameters`**: every estimated parameter (loadings, variances, covariances) with SE, z, p, and confidence intervals.
- **`modification_indices`**: sorted here by size — the largest values are the parameters most likely to improve fit if freed (again, only worth freeing with a substantive rationale).

The rest of the report holds the input data and model matrices:

```{r report_cfa_output_rest}
result$sample_covariance
result$standardized_estimates
head(result$predict)
result$call
```

- **`sample_covariance`**: the observed covariance matrix of the indicators that the model tries to reproduce.
- **`unstandardized_estimates`** / **`standardized_estimates`**: the model matrices — `lambda` (loadings), `theta` (residual variances and covariances) and `psi` (factor variances and covariances).
- **`predict`**: the observed data with each case's predicted factor scores added as extra columns.
- **`call`**: the `lavaan` call that fitted the model.

With `file`, the report is also saved: `report_cfa(fit, file = "cfa")` writes every table to `cfa.xlsx` and all diagrams to `cfa_diagram.pdf`, with `w` and `h` setting the PDF page size in inches.

### Fit indices in the report

`result$fit_indices` holds every fit measure `lavaan` computes (Rosseel, 2012). Cut-offs are given only where the literature gives one; "lower is better" indices are for comparing models fitted to the same data. The `_nomean` variants ignore the mean structure, so they equal the plain index when the model has no means, as here.

| Row in `fit_indices` | Index | Good fit | Reference |
|---|---|---|---|
| `npar` | Number of free parameters | — | — |
| `fmin` | Minimum of the fit function, $\chi^2/(2N)$ | — | Jöreskog (1969) |
| `chisq`, `df`, `pvalue` | $\chi^2$ test of exact fit | $p > .05$ | Jöreskog (1969) |
| `baseline.chisq`, `baseline.df`, `baseline.pvalue` | $\chi^2$ test of the baseline (independence) model, used by the incremental indices | — | Bentler & Bonett (1980) |
| `cfi` | Comparative Fit Index | $\geq .95$ | Bentler (1990); Hu & Bentler (1999) |
| `tli`, `nnfi` | Tucker-Lewis Index / Non-Normed Fit Index (same value) | $\geq .95$ | Tucker & Lewis (1973); Bentler & Bonett (1980); Hu & Bentler (1999) |
| `rfi` | Relative Fit Index | close to 1 | Bollen (1986) |
| `nfi` | Normed Fit Index | $\geq .90$ | Bentler & Bonett (1980) |
| `pnfi` | Parsimony Normed Fit Index | higher is better | James et al. (1982) |
| `ifi` | Incremental Fit Index | close to 1 | Bollen (1989a) |
| `rni` | Relative Noncentrality Index | $\geq .95$ | McDonald & Marsh (1990); Hu & Bentler (1999) |
| `logl`, `unrestricted.logl` | Log-likelihood of the model and of the saturated model | — | Jöreskog (1969) |
| `aic` | Akaike Information Criterion | lower is better | Akaike (1974) |
| `bic` | Bayesian Information Criterion | lower is better | Schwarz (1978) |
| `ntotal` | Total sample size | — | — |
| `bic2` | Sample-size adjusted BIC | lower is better | Sclove (1987) |
| `rmsea` | Root Mean Square Error of Approximation | $\leq .06$ | Steiger & Lind (1980); Steiger (1990); Hu & Bentler (1999) |
| `rmsea.ci.lower`, `rmsea.ci.upper`, `rmsea.ci.level` | 90% confidence interval of RMSEA | — | Browne & Cudeck (1992) |
| `rmsea.pvalue`, `rmsea.close.h0` | Test of close fit, $H_0$: RMSEA $\leq .05$ | $p > .05$ | Browne & Cudeck (1992) |
| `rmsea.notclose.pvalue`, `rmsea.notclose.h0` | Test of not-close fit, $H_0$: RMSEA $\geq .08$ | $p < .05$ | MacCallum et al. (1996) |
| `rmr`, `rmr_nomean` | Root Mean Square Residual (unstandardized, depends on the scale of the variables) | — | Jöreskog & Sörbom (1981) |
| `srmr`, `srmr_bentler`, `srmr_bentler_nomean` | Standardized Root Mean Square Residual | $\leq .08$ | Bentler (1995); Hu & Bentler (1999) |
| `crmr`, `crmr_nomean` | Correlation Root Mean Square Residual (residual correlations, diagonal excluded) | — | Bollen (1989b) |
| `srmr_mplus`, `srmr_mplus_nomean` | SRMR as computed by Mplus | $\leq .08$ | Asparouhov & Muthén (2018) |
| `gfi`, `gfi.ci.lower`, `gfi.ci.upper`, `gfi.ci.level` | Goodness of Fit Index, with confidence interval | close to 1 | Jöreskog & Sörbom (1981) |
| `cn_05`, `cn_01` | Hoelter's critical N at $\alpha = .05$ and $.01$ | $> 200$ | Hoelter (1983) |
| `gfi_lisrel`, `agfi_lisrel` | GFI and Adjusted GFI as computed by LISREL | close to 1 | Jöreskog & Sörbom (1981) |
| `pgfi` | Parsimony Goodness of Fit Index | higher is better | Mulaik et al. (1989) |
| `mfi` | McDonald's Fit Index | close to 1 | McDonald (1989) |
| `ecvi` | Expected Cross-Validation Index | lower is better | Browne & Cudeck (1989) |

### Multi-group models

When the fitted model has more than one group, `report_cfa` additionally returns a `group` summary table (group labels, sample sizes per group, and total N) — omitted here since the example above is single-group.

## Sample Size Planning

`simulate_cfa_fit` refits a CFA model repeatedly across a range of sample sizes, in parallel, and returns fit indices at each size — either generating data from known population parameters (`model_sim`) or resampling from an observed dataset's correlation structure (`df`).

```{r simulate_cfa_fit,results='hide'}
model_sim <- "LATENT =~ 1*X1 + 0.5*X2 + 1.5*X3 + 1.5*X4 + X5"
model_est <- "LATENT =~ X1 + X2 + X3 + X4 + X5"
result_sim <- simulate_cfa_fit(model_sim = model_sim, model = model_est, minnobs = 50, maxnobs = 10000, stepping = 100)
```

The first element is a table with one row per sample size and one column per fit index; the second is a named list of plots, one per fit index:

```{r simulate_cfa_fit_table}
head(result_sim[[1]][, c("observations", "cfi", "tli", "rmsea", "srmr")])
```

```{r simulate_cfa_fit_plots,fig.width=10,fig.height=8}
plot_multiplot(plotlist = result_sim[[2]][c("observations_cfi", "observations_tli",
                                            "observations_rmsea", "observations_srmr")],
               cols = 2)
```

Each panel tracks one fit index against sample size. Indices like RMSEA and SRMR typically stabilize and tighten (less sample-to-sample variability) as $n$ grows — the point at which a given index's variability becomes acceptably small for your purposes is a data-driven answer to "how large a sample do I need?" for this specific model.

---

## Conclusion

`plot_cfa_gg`/`plot_cfa` visualize a fitted measurement model's structure without requiring `semPlot`'s system dependencies; `report_cfa` collects every quantity needed to write up a CFA (fit, parameters, R², modification indices) in one call; and `simulate_cfa_fit` turns "what sample size do I need" into an empirical, visualizable question rather than a rule of thumb.

## References

Akaike, H. (1974). A new look at the statistical model identification. *IEEE Transactions on Automatic Control, 19*(6), 716–723. https://doi.org/10.1109/TAC.1974.1100705

Asparouhov, T., & Muthén, B. (2018). *SRMR in Mplus* [Technical report]. Muthén & Muthén. https://www.statmodel.com/download/SRMR2.pdf

Bentler, P. M. (1990). Comparative fit indexes in structural models. *Psychological Bulletin, 107*(2), 238–246. https://doi.org/10.1037/0033-2909.107.2.238

Bentler, P. M. (1995). *EQS structural equations program manual*. Multivariate Software.

Bentler, P. M., & Bonett, D. G. (1980). Significance tests and goodness of fit in the analysis of covariance structures. *Psychological Bulletin, 88*(3), 588–606. https://doi.org/10.1037/0033-2909.88.3.588

Bollen, K. A. (1986). Sample size and Bentler and Bonett's nonnormed fit index. *Psychometrika, 51*(3), 375–377. https://doi.org/10.1007/BF02294061

Bollen, K. A. (1989a). A new incremental fit index for general structural equation models. *Sociological Methods & Research, 17*(3), 303–316. https://doi.org/10.1177/0049124189017003004

Bollen, K. A. (1989b). *Structural equations with latent variables*. Wiley. https://doi.org/10.1002/9781118619179

Browne, M. W., & Cudeck, R. (1989). Single sample cross-validation indices for covariance structures. *Multivariate Behavioral Research, 24*(4), 445–455. https://doi.org/10.1207/s15327906mbr2404_4

Browne, M. W., & Cudeck, R. (1992). Alternative ways of assessing model fit. *Sociological Methods & Research, 21*(2), 230–258. https://doi.org/10.1177/0049124192021002005

Hoelter, J. W. (1983). The analysis of covariance structures: Goodness-of-fit indices. *Sociological Methods & Research, 11*(3), 325–344. https://doi.org/10.1177/0049124183011003003

Hu, L., & Bentler, P. M. (1999). Cutoff criteria for fit indexes in covariance structure analysis: Conventional criteria versus new alternatives. *Structural Equation Modeling, 6*(1), 1–55. https://doi.org/10.1080/10705519909540118

James, L. R., Mulaik, S. A., & Brett, J. M. (1982). *Causal analysis: Assumptions, models, and data*. Sage.

Jöreskog, K. G. (1969). A general approach to confirmatory maximum likelihood factor analysis. *Psychometrika, 34*(2), 183–202. https://doi.org/10.1007/BF02289343

Jöreskog, K. G., & Sörbom, D. (1981). *LISREL V: Analysis of linear structural relationships by maximum likelihood and least squares methods*. National Educational Resources.

MacCallum, R. C., Browne, M. W., & Sugawara, H. M. (1996). Power analysis and determination of sample size for covariance structure modeling. *Psychological Methods, 1*(2), 130–149. https://doi.org/10.1037/1082-989X.1.2.130

McDonald, R. P. (1989). An index of goodness-of-fit based on noncentrality. *Journal of Classification, 6*(1), 97–103. https://doi.org/10.1007/BF01908590

McDonald, R. P., & Marsh, H. W. (1990). Choosing a multivariate model: Noncentrality and goodness of fit. *Psychological Bulletin, 107*(2), 247–255. https://doi.org/10.1037/0033-2909.107.2.247

Mulaik, S. A., James, L. R., Van Alstine, J., Bennett, N., Lind, S., & Stilwell, C. D. (1989). Evaluation of goodness-of-fit indices for structural equation models. *Psychological Bulletin, 105*(3), 430–445. https://doi.org/10.1037/0033-2909.105.3.430

Rosseel, Y. (2012). lavaan: An R package for structural equation modeling. *Journal of Statistical Software, 48*(2), 1–36. https://doi.org/10.18637/jss.v048.i02

Schwarz, G. (1978). Estimating the dimension of a model. *The Annals of Statistics, 6*(2), 461–464. https://doi.org/10.1214/aos/1176344136

Sclove, S. L. (1987). Application of model-selection criteria to some problems in multivariate analysis. *Psychometrika, 52*(3), 333–343. https://doi.org/10.1007/BF02294360

Steiger, J. H. (1990). Structural model evaluation and modification: An interval estimation approach. *Multivariate Behavioral Research, 25*(2), 173–180. https://doi.org/10.1207/s15327906mbr2502_4

Steiger, J. H., & Lind, J. C. (1980). *Statistically based tests for the number of common factors* [Paper presentation]. Annual meeting of the Psychometric Society, Iowa City, IA, United States.

Tucker, L. R., & Lewis, C. (1973). A reliability coefficient for maximum likelihood factor analysis. *Psychometrika, 38*(1), 1–10. https://doi.org/10.1007/BF02291170

---

*Rendered with R `r getRversion()` · lavaan `r as.character(packageVersion("lavaan"))`*
