---
title: "GLM COMPUTE AOV ES"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM COMPUTE AOV ES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
options(contrasts = c("contr.sum", "contr.poly"))
```

## Description

This shows the output of `compute_aov_es` from the package rwf, which takes a between-subjects `aov` model, builds its ANOVA table with Type I, II or III sums of squares and adds six effect sizes for every factor and interaction: eta-squared, partial eta-squared, omega-squared, partial omega-squared, epsilon-squared and Cohen's f. Each section gives the formulas the function uses, computes them by hand and checks the results against `effectsize`, `sjstats`, `car` and `compute_one_way_test`.

The computation of partial omega-squared is adapted from the answer of Stephen Martin (2014) to the question "Omega squared for measure of effect in R?" on Cross Validated (https://stats.stackexchange.com/a/126520). The Type II and III sums of squares come from `car::Anova` (Fox & Weisberg, 2019).

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/GLM_ANOVA.R)

Type III sums of squares need sum-to-zero contrasts ([below](#type-iii-and-contrasts)), so this document sets them before any model is fitted:

```{r contrasts,eval=FALSE}
options(contrasts = c("contr.sum", "contr.poly"))
```

## Notation

A term is a factor or an interaction of factors. For a term with $df_{\text{effect}}$ degrees of freedom, $SS_{\text{effect}}$ is its sum of squares and $MS_{\text{effect}}=SS_{\text{effect}}/df_{\text{effect}}$ its mean square. $SS_{\text{error}}$, $df_{\text{error}}$ and $MS_{\text{error}}$ belong to the residuals. $SS_{\text{total}}$ is the sum of the sums of squares of all the terms and the residuals in the ANOVA table, and $N$ the number of observations used by the model. $SS(A\mid B)$ is the reduction in the residual sum of squares when $A$ is added to a model that already contains $B$.

## Types of sums of squares

In a factorial design with factors $A$ and $B$, the sum of squares of each term is the variance it explains **after** some other terms are already in the model. The three types differ in which terms come first:

| Term | Type I (sequential) | Type II | Type III |
|---|---|---|---|
| $A$ | $SS(A)$ | $SS(A\mid B)$ | $SS(A\mid B, AB)$ |
| $B$ | $SS(B\mid A)$ | $SS(B\mid A)$ | $SS(B\mid A, AB)$ |
| $AB$ | $SS(AB\mid A, B)$ | $SS(AB\mid A, B)$ | $SS(AB\mid A, B)$ |

- **Type I** adds the terms in the order of the formula, so the result depends on that order. It suits designs where the order has a meaning, such as a covariate entered first.
- **Type II** adjusts each term for every other term that does not contain it. Main effects are not adjusted for their interactions. It is the most powerful test of main effects when there is no interaction (Fox & Weisberg, 2019).
- **Type III** adjusts each term for all the others, including the interactions. It tests main effects averaged over the levels of the other factors and is the default of SPSS and SAS.

In a **balanced** design (equal numbers in every cell) the three types give the same sums of squares. In an **unbalanced** design they differ. `mtcars` has between 2 and 12 cars per cell of cylinders by transmission:

```{r unbalanced}
cars <- transform(mtcars, cyl = factor(cyl), am = factor(am))
table(cars$cyl, cars$am)
model_cars <- stats::aov(mpg ~ cyl * am, data = cars)
compute_aov_es(model = model_cars, ss = "I")[, c("comparisons", "Df", "Sum Sq", "F value", "Pr(>F)")]
compute_aov_es(model = model_cars, ss = "II")[, c("comparisons", "Df", "Sum Sq", "F value", "Pr(>F)")]
compute_aov_es(model = model_cars, ss = "III")[, c("comparisons", "Df", "Sum Sq", "F value", "Pr(>F)")]
```

Type I depends on the order of the terms. With `am` first, `am` gets all the variance it shares with `cyl`:

```{r type_i_order}
compute_aov_es(model = stats::aov(mpg ~ am * cyl, data = cars), ss = "I")[, c("comparisons", "Sum Sq")]
```

Each sum of squares is a comparison of two models. The Type II sum of squares of `cyl` is the drop in the residual sum of squares when `cyl` is added to a model with `am`:

```{r type_ii_model_comparison}
stats::anova(stats::lm(mpg ~ am, data = cars), stats::lm(mpg ~ am + cyl, data = cars))
```

### Type III and contrasts

Type III tests a main effect with the interaction in the model, so it depends on how the factor levels are coded. With sum-to-zero contrasts (`contr.sum`), the main effect is averaged over the levels of the other factor. With R's default treatment contrasts (`contr.treatment`), it becomes the effect at the reference level of the other factor, which is rarely the question asked. The interaction is the same either way:

```{r type_iii_contrasts}
model_treatment <- stats::aov(mpg ~ cyl * am, data = cars,
                              contrasts = list(cyl = "contr.treatment", am = "contr.treatment"))
data.frame(term = c("cyl", "am", "cyl:am"),
           ss_contr_sum = compute_aov_es(model = model_cars, ss = "III")$`Sum Sq`[1:3],
           ss_contr_treatment = compute_aov_es(model = model_treatment, ss = "III")$`Sum Sq`[1:3])
```

Set `options(contrasts = c("contr.sum", "contr.poly"))` **before** fitting the model whenever you use `ss = "III"`. Types I and II do not depend on the contrasts.

## When to use it and assumptions

### When to use it

Use `compute_aov_es` to report the size of each effect of a between-subjects ANOVA: one or more factors, each participant in one cell only. It works with any `aov` model whose ANOVA table `summary` (Type I) or `car::Anova` (Types II and III) can compute.

Choosing the type of sums of squares:

- **Balanced design:** any type; they agree.
- **Unbalanced design, no interaction of interest:** Type II.
- **Unbalanced design with an interaction:** Type III with sum-to-zero contrasts, or interpret the simple effects instead of the main effects.
- **Terms with a natural order** (for example a covariate entered before the factor of interest): Type I, with the terms in that order.

Do not use it when:

- **The same participants are measured more than once.** Models with an `Error()` term have several error strata, and the function only reads the first.
- **The model has aliased terms**, such as an interaction that is confounded with blocks. Type I still works, but `car::Anova` cannot compute Type II or III ([example below](#aliased-terms)).

### Assumptions

The effect sizes describe the ANOVA, so they inherit its assumptions:

1. **Independent observations.** Each participant contributes one observation to one cell.
2. **Normally distributed residuals.** With moderate cell sizes the F tests are fairly robust to non-normality.
3. **Equal variances across cells.** With unequal cell sizes, unequal variances can bias the F tests.

## Why effect sizes

An F test answers one question: if a factor had no effect, how surprising would data like these be? It does not say how large the effect is. Because $F$ grows with the sample size, a trivial effect becomes "significant" in a large sample, and a large effect can be "non-significant" in a small one.

An effect size answers a different question: how much of the variance in the outcome does each factor explain? Effect sizes are useful because they:

- **measure magnitude.** They separate practical importance from statistical significance.
- **do not grow with $N$.** The same population effect gives roughly the same effect size in a study of 30 people or 3,000.
- **can be compared across studies.** This is what meta-analysis combines.
- **drive power analysis.** The sample size a new study needs depends on the effect size it expects.
- **are expected in reports.** The APA Task Force on Statistical Inference recommends reporting effect sizes for primary outcomes (Wilkinson & Task Force on Statistical Inference, 1999; see also Lakens, 2013).

## Effect sizes

### Formulas

$$\eta^2=\frac{SS_{\text{effect}}}{SS_{\text{total}}},\qquad \eta^2_p=\frac{SS_{\text{effect}}}{SS_{\text{effect}}+SS_{\text{error}}}$$

$$\omega^2=\frac{SS_{\text{effect}}-df_{\text{effect}}\,MS_{\text{error}}}{SS_{\text{total}}+MS_{\text{error}}},\qquad \omega^2_p=\frac{df_{\text{effect}}\left(MS_{\text{effect}}-MS_{\text{error}}\right)}{SS_{\text{effect}}+\left(N-df_{\text{effect}}\right)MS_{\text{error}}}$$

$$\varepsilon^2=\frac{SS_{\text{effect}}-df_{\text{effect}}\,MS_{\text{error}}}{SS_{\text{total}}},\qquad f=\sqrt{\frac{\eta^2_p}{1-\eta^2_p}}$$

- **Eta-squared** $\eta^2$ is the proportion of the total variance explained by the term.
- **Omega-squared** $\omega^2$ (Hays, 1963) and **epsilon-squared** $\varepsilon^2$ (Kelley, 1935) subtract $df_{\text{effect}}\,MS_{\text{error}}$, the part of $SS_{\text{effect}}$ expected by chance. They estimate the proportion of variance explained in the population and remove most of the upward bias of $\eta^2$. They differ only in the denominator.
- **Partial** measures (Olejnik & Algina, 2003) divide by the variance of the term and the error only, leaving out the variance explained by the other terms.
- **Cohen's f** is computed from $\eta^2_p$, so it is a partial measure. It is the effect size used in power analysis (Cohen, 1988).

### Partial and non-partial

The non-partial measures share one denominator, so they add up to at most 1 across the terms and depend on the other factors in the design. The partial measures do not add up to anything meaningful, but the value of a term does not shrink when other factors explain a lot of variance, which makes them easier to compare between studies with different designs. In a one-way design the two are the same.

`ToothGrowth` is balanced, so the sum of squares of `supp` is the same with or without `dose` in the model. Adding `dose`, which explains most of the variance, leaves $\eta^2$ of `supp` unchanged but raises $\eta^2_p$, because the error variance shrinks:

```{r partial}
tooth <- transform(ToothGrowth, dose = factor(dose))
rbind(compute_aov_es(model = stats::aov(len ~ supp, data = tooth))[1, c("comparisons", "etasq", "partial_etasq")],
      compute_aov_es(model = stats::aov(len ~ supp * dose, data = tooth))[1, c("comparisons", "etasq", "partial_etasq")])
```

### Negative values

$\omega^2$, $\omega^2_p$ and $\varepsilon^2$ are negative when $F<1$, when a term explains less variance than expected by chance. `compute_aov_es` returns them as they are, as `sjstats::anova_stats` does, and `effectsize` reports them as 0 ([example below](#negative-values-1)). Report a negative value as 0 or as it is, but do not read it as a negative effect.

### Interpretation

Multiplied by 100, $\eta^2$, $\omega^2$ and $\varepsilon^2$ read as the percentage of variance explained. Small, medium and large are the benchmarks of Cohen (1988). Sawilowsky (2009) added very small, very large and huge benchmarks for Cohen's $d$, and the table converts them with $\eta^2=d^2/(d^2+4)$ and $f=d/2$, the relations for two groups of equal size:

| Magnitude | $\eta^2$, $\omega^2$, $\varepsilon^2$ | Cohen's $f$ | Cohen's $d$ |
|---|---|---|---|
| tiny | < 0.01 | < 0.10 | < 0.2 |
| small | 0.01 to < 0.06 | 0.10 to < 0.25 | 0.2 |
| medium | 0.06 to < 0.14 | 0.25 to < 0.40 | 0.5 |
| large | 0.14 to < 0.26 | 0.40 to < 0.60 | 0.8 |
| very large | 0.26 to < 0.50 | 0.60 to < 1.00 | 1.2 |
| huge | ≥ 0.50 | ≥ 1.00 | 2.0 |

These are rough guides. What counts as a meaningful effect depends on the field and the outcome. Partial measures are usually larger than non-partial ones, so apply the benchmarks to the measure you report and say which one it is.

## Data

`ToothGrowth` records the length of the odontoblasts of 60 guinea pigs, each given vitamin C at one of three doses (0.5, 1 and 2 mg/day) by one of two delivery methods (orange juice or ascorbic acid). The design is a balanced 2 by 3 factorial with 10 animals per cell.

```{r data}
head(tooth)
table(tooth$supp, tooth$dose)
```

## compute_aov_es

```{r compute_aov_es}
model_tooth <- stats::aov(len ~ supp * dose, data = tooth)
compute_aov_es(model = model_tooth, ss = "I")
compute_aov_es(model = model_tooth, ss = "II")
compute_aov_es(model = model_tooth, ss = "III")
```

The design is balanced, so the three types give the same effect sizes. Type III also reports the intercept.

The output columns are:

| Column | Meaning |
|---|---|
| `call` | the model formula |
| `ss` | type of sums of squares |
| `comparisons` | the term |
| `Df`, `Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)` | the ANOVA table |
| `etasq`, `partial_etasq` | $\eta^2$, $\eta^2_p$ |
| `omegasq`, `partial_omegasq` | $\omega^2$, $\omega^2_p$ |
| `epsilonsq` | $\varepsilon^2$ |
| `cohens_f` | $f$ from $\eta^2_p$ |

The effect sizes are `NA` for the residuals and the intercept.

## Step by step

Every effect size computed by hand from the Type II table of the unbalanced `mtcars` model, where the sums of squares of the terms and the residuals do not add up to the total variance of `mpg`:

```{r step_by_step}
table_ii <- as.data.frame(car::Anova(model_cars, type = "II"))
terms <- rownames(table_ii) != "Residuals"
ss_effect <- table_ii$`Sum Sq`[terms]
df_effect <- table_ii$Df[terms]
ms_effect <- ss_effect / df_effect
ss_error <- table_ii["Residuals", "Sum Sq"]
ms_error <- ss_error / table_ii["Residuals", "Df"]
ss_total <- sum(table_ii$`Sum Sq`)
N <- nrow(stats::model.frame(model_cars))
c(ss_total_table = ss_total, ss_total_mpg = sum((cars$mpg - mean(cars$mpg))^2))
partial_etasq <- ss_effect / (ss_effect + ss_error)
by_hand <- data.frame(comparisons = rownames(table_ii)[terms],
                      etasq = ss_effect / ss_total,
                      partial_etasq = partial_etasq,
                      omegasq = (ss_effect - df_effect * ms_error) / (ss_total + ms_error),
                      partial_omegasq = df_effect * (ms_effect - ms_error) / (ss_effect + (N - df_effect) * ms_error),
                      epsilonsq = (ss_effect - df_effect * ms_error) / ss_total,
                      cohens_f = sqrt(partial_etasq / (1 - partial_etasq)))
by_hand
rwf <- compute_aov_es(model = model_cars, ss = "II")
all.equal(by_hand[, -1], rwf[!is.na(rwf$etasq), names(by_hand)[-1]], check.attributes = FALSE)
```

## Checks against other packages

### effectsize and sjstats

`effectsize` and `sjstats` compute the same effect sizes from the `aov` model (Type I) or from the `car::Anova` table (Types II and III):

```{r check_packages}
effectsize::eta_squared(car::Anova(model_cars, type = 3), partial = FALSE, ci = NULL)
effectsize::eta_squared(car::Anova(model_cars, type = 3), partial = TRUE, ci = NULL)
effectsize::omega_squared(car::Anova(model_cars, type = 3), partial = FALSE, ci = NULL)
effectsize::omega_squared(car::Anova(model_cars, type = 3), partial = TRUE, ci = NULL)
effectsize::epsilon_squared(car::Anova(model_cars, type = 3), partial = FALSE, ci = NULL)
effectsize::cohens_f(car::Anova(model_cars, type = 3), partial = TRUE, ci = NULL)
sjstats::anova_stats(car::Anova(model_cars, type = 3), digits = 22)
compute_aov_es(model = model_cars, ss = "III")
```

### compute_one_way_test

In a one-way design, `compute_aov_es` and `compute_one_way_test` (with `var.equal = TRUE`) give the same $\eta^2$, $\omega^2$ and Cohen's $f$:

```{r check_one_way}
one_way <- compute_one_way_test(formula = weight ~ feed, df = chickwts, var.equal = TRUE)
aov_es <- compute_aov_es(model = stats::aov(weight ~ feed, data = chickwts))
data.frame(statistic = c("etasq", "omegasq", "partial_omegasq", "cohens_f"),
           compute_one_way_test = c(one_way$etasq, one_way$omegasq, one_way$partial.omegasq, one_way$cohens.f),
           compute_aov_es = c(aov_es$etasq[1], aov_es$omegasq[1], aov_es$partial_omegasq[1], aov_es$cohens_f[1]))
```

### A test function

`check_aov_es` puts each rwf value next to the value from `effectsize` and `sjstats` and returns the differences, one row per term. `effectsize` reports negative values as 0, so the rwf values are truncated at 0 before the comparison with `effectsize`.

```{r check_function}
check_aov_es <- function(model, ss) {
  rwf <- compute_aov_es(model = model, ss = ss)
  rwf <- rwf[!is.na(rwf$etasq), ]
  table <- if (ss == "I") model else car::Anova(model, type = ss)
  effectsize_value <- function(fun, ...) {
    values <- as.data.frame(suppressMessages(suppressWarnings(fun(table, ci = NULL, ...))))
    values[match(rwf$comparisons, values$Parameter), 2]
  }
  sjstats_table <- as.data.frame(suppressMessages(suppressWarnings(sjstats::anova_stats(table, digits = 22))))
  sjstats_table <- sjstats_table[match(rwf$comparisons, sjstats_table$term), ]
  data.frame(term = rwf$comparisons,
             etasq_effectsize = rwf$etasq - effectsize_value(effectsize::eta_squared, partial = FALSE),
             partial_etasq_effectsize = rwf$partial_etasq - effectsize_value(effectsize::eta_squared, partial = TRUE),
             omegasq_effectsize = pmax(0, rwf$omegasq) - effectsize_value(effectsize::omega_squared, partial = FALSE),
             partial_omegasq_effectsize = pmax(0, rwf$partial_omegasq) - effectsize_value(effectsize::omega_squared, partial = TRUE),
             epsilonsq_effectsize = pmax(0, rwf$epsilonsq) - effectsize_value(effectsize::epsilon_squared, partial = FALSE),
             cohens_f_effectsize = rwf$cohens_f - effectsize_value(effectsize::cohens_f, partial = TRUE),
             etasq_sjstats = rwf$etasq - sjstats_table$etasq,
             partial_etasq_sjstats = rwf$partial_etasq - sjstats_table$partial.etasq,
             omegasq_sjstats = rwf$omegasq - sjstats_table$omegasq,
             partial_omegasq_sjstats = rwf$partial_omegasq - sjstats_table$partial.omegasq,
             epsilonsq_sjstats = rwf$epsilonsq - sjstats_table$epsilonsq,
             cohens_f_sjstats = rwf$cohens_f - sjstats_table$cohens.f,
             row.names = NULL)
}
check_aov_es(model = model_cars, ss = "III")
```

The same check on five models, balanced and unbalanced, with one to four terms, for all three types of sums of squares. `max_abs_difference` is the largest absolute difference across all the comparisons and terms:

```{r check_models}
models <- list(
  "len ~ supp * dose" = model_tooth,
  "mpg ~ cyl * am" = model_cars,
  "breaks ~ wool * tension" = stats::aov(breaks ~ wool * tension, data = warpbreaks),
  "yield ~ block + N * P + K" = stats::aov(yield ~ block + N * P + K, data = npk),
  "weight ~ feed" = stats::aov(weight ~ feed, data = chickwts)
)
do.call(rbind, lapply(names(models), function(name) {
  do.call(rbind, lapply(c("I", "II", "III"), function(ss) {
    check <- check_aov_es(model = models[[name]], ss = ss)
    data.frame(model = name, ss = ss, terms = nrow(check),
               max_abs_difference = max(abs(as.matrix(check[, -1]))))
  }))
}))
```

## Bias of eta-squared

In a 2 by 3 design with 5 observations per cell and **no** effects at all, the average $\eta^2$ and $\eta^2_p$ are well above 0, while $\omega^2$, $\omega^2_p$ and $\varepsilon^2$ are close to 0:

```{r simulation_bias}
set.seed(1)
null_effects <- do.call(rbind, replicate(1000, {
  sim <- data.frame(a = factor(rep(1:2, each = 15)), b = factor(rep(1:3, times = 10)), y = rnorm(30))
  compute_aov_es(model = stats::aov(y ~ a * b, data = sim))[1:3, c("comparisons", "etasq", "partial_etasq", "omegasq", "partial_omegasq", "epsilonsq")]
}, simplify = FALSE))
aggregate(cbind(etasq, partial_etasq, omegasq, partial_omegasq, epsilonsq) ~ comparisons, data = null_effects, FUN = mean)
```

The bias grows with the degrees of freedom of the term: `b` and `a:b` have 2 degrees of freedom and `a` has 1. With small samples, report $\omega^2$, $\omega^2_p$ or $\varepsilon^2$.

## Negative values

A factor of random labels explains less variance than expected by chance here ($F<1$). rwf and `sjstats` return negative $\omega^2$, $\omega^2_p$ and $\varepsilon^2$, while `effectsize` reports 0:

```{r negative_values}
set.seed(5)
tooth_noise <- transform(tooth, noise = factor(sample(c("a", "b"), nrow(tooth), replace = TRUE)))
model_noise <- stats::aov(len ~ dose + noise, data = tooth_noise)
compute_aov_es(model = model_noise)[, c("comparisons", "F value", "omegasq", "partial_omegasq", "epsilonsq")]
as.data.frame(sjstats::anova_stats(model_noise, digits = 22))[, c("term", "omegasq", "partial.omegasq", "epsilonsq")]
effectsize::omega_squared(model_noise, partial = FALSE, ci = NULL)
```

## Missing values

`aov` removes the rows with a missing value in any variable of the model before fitting, and `compute_aov_es` takes $N$ from the rows the model used:

```{r missing_values}
tooth_missing <- tooth
tooth_missing$len[c(1, 31)] <- NA
model_missing <- stats::aov(len ~ supp * dose, data = tooth_missing)
nrow(stats::model.frame(model_missing))
check_aov_es(model = model_missing, ss = "II")
```

## Aliased terms

In `npk` the three-way interaction `N:P:K` is confounded with the blocks, so it cannot be estimated. Type I sums of squares still work, but `car::Anova` cannot compute Types II and III:

```{r aliased}
model_npk <- stats::aov(yield ~ block + N * P * K, data = npk)
compute_aov_es(model = model_npk, ss = "I")[, c("comparisons", "Df", "Sum Sq", "etasq", "partial_etasq")]
tryCatch(compute_aov_es(model = model_npk, ss = "III"), error = conditionMessage)
```

Drop the aliased term (here `yield ~ block + N * P + K`, as in the [checks above](#a-test-function)) to use Types II and III.

## References

Ben-Shachar, M. S., Lüdecke, D., & Makowski, D. (2020). effectsize: Estimation of effect size indices and standardized parameters. *Journal of Open Source Software, 5*(56), 2815. https://doi.org/10.21105/joss.02815

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Fox, J., & Weisberg, S. (2019). *An R companion to applied regression* (3rd ed.). Sage. https://www.john-fox.ca/Companion/

Hays, W. L. (1963). *Statistics for psychologists*. Holt, Rinehart and Winston.

Kelley, T. L. (1935). An unbiased correlation ratio measure. *Proceedings of the National Academy of Sciences, 21*(9), 554–559. https://doi.org/10.1073/pnas.21.9.554

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate cumulative science: A practical primer for t-tests and ANOVAs. *Frontiers in Psychology, 4*, 863. https://doi.org/10.3389/fpsyg.2013.00863

Martin, S. (2014, December 3). Answer to "Omega squared for measure of effect in R?" *Cross Validated*. https://stats.stackexchange.com/a/126520

Olejnik, S., & Algina, J. (2003). Generalized eta and omega squared statistics: Measures of effect size for some common research designs. *Psychological Methods, 8*(4), 434–447. https://doi.org/10.1037/1082-989X.8.4.434

Sawilowsky, S. S. (2009). New effect size rules of thumb. *Journal of Modern Applied Statistical Methods, 8*(2), 597–599. https://doi.org/10.22237/jmasm/1257035100

Wilkinson, L., & Task Force on Statistical Inference. (1999). Statistical methods in psychology journals: Guidelines and explanations. *American Psychologist, 54*(8), 594–604. https://doi.org/10.1037/0003-066X.54.8.594
