---
title: "GLM IRT Thurstonian"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM IRT Thurstonian}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
library(thurstonianIRT)
```

## Description

This vignette documents `working_functions/GLM_IRT_T.R`, the largest single file in `rwf`'s IRT toolkit (~22 functions). It covers **Thurstonian IRT** for forced-choice/ranking questionnaires — where respondents rank or compare items within blocks rather than rating each item independently — end to end: building the comparison design, encoding rank data into the binary pairwise-comparison format the model needs, a from-scratch single-dimension item characteristic curve and grid-based ability estimator, and a modern MAP/Empirical-Bayes-Modal scorer that reproduces `thurstonianIRT::predict()` from a fitted `lavaan` model's parameters directly, plus a Heywood-case diagnostic for the underlying `lavaan` fit.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### Why forced-choice items need a different IRT model

In a forced-choice block, a respondent doesn't rate each item on its own scale (which is vulnerable to response styles like acquiescence or extreme responding) — they rank or compare items *against each other* within the block. Thurstone's (1927) **law of comparative judgment** models each item's momentary appeal as a latent utility with a normal distribution; comparing two items reduces to comparing which of two normal variables is larger, giving a normal-ogive (probit) response model for each pairwise comparison rather than a single item response curve.

### From a block to pairwise comparisons

A block of $k$ items yields $\binom{k}{2}=k(k-1)/2$ unique pairwise comparisons. Each comparison is coded $+1$/$-1$ in a **comparison matrix**: item $i$ beating item $j$ contributes a row with $+1$ in column $i$ and $-1$ in column $j$. Stacking these rows across all pairs in a block, and stacking blocks block-diagonally across an entire questionnaire, gives the full design (**lambda**) matrix used to fit the Thurstonian model.

### The single-dimension item characteristic curve

For one trait, comparing items with loadings $\lambda$, thresholds (crossing points) $\gamma$, and residual variances $\psi$, the probability that a given pairwise comparison resolves in a particular direction at trait level $\eta$ is a normal-ogive function:

$$P(\eta) = \Phi\left(\frac{-\gamma+\lambda\eta}{\sqrt{\psi}}\right)$$

### MAP / Empirical Bayes Modal (EBM) scoring

Given a binary response pattern across all pairwise comparisons, the **maximum a posteriori** trait estimate maximizes the posterior:

$$\hat{\boldsymbol\eta}_{MAP} = \arg\max_{\boldsymbol\eta} \left[ \sum_i y_i\log P_i(\boldsymbol\eta) + (1-y_i)\log(1-P_i(\boldsymbol\eta)) - \tfrac{1}{2}\boldsymbol\eta^\top\Psi^{-1}\boldsymbol\eta \right]$$

— a probit likelihood over all observed comparisons, plus a multivariate normal prior (with covariance $\Psi$) on the traits. This is the same likelihood-plus-prior structure as the EAP estimator covered in [`EXAMPLE_EAP.rmd`](EXAMPLE_EAP.rmd), except optimized directly (via BFGS) rather than by grid integration, and extended to multiple correlated traits at once.

---

## Building a Comparison Design

For one block of `items` items, `generate_comparisons_matrix` builds the $\binom{k}{2}\times k$ matrix of $+1$/$-1$ pairwise comparisons:

```{r comparisons_matrix}
compute_dummy_comparisons(items = 3)
generate_unique_comparisons_index(3)
generate_comparisons_matrix(3)
```

`generate_matrix_lambda_hat` stacks this same block structure vertically across several *identical* blocks, while `generate_matrix_A` builds the full **block-diagonal** design across `blocks` blocks of `items` items each — the design matrix for an entire questionnaire made of independent blocks:

```{r matrix_A}
generate_matrix_lambda_hat(blocks = 3, items = 3)
generate_matrix_A(blocks = 3, items = 3)
```

`increase_index` is the small helper underlying both: it produces the row (or column) index ranges that pick out block $i$'s slice of a stacked/block-diagonal matrix.

```{r increase_index}
increase_index(blocks = 3, items = 3)
```

## Encoding Rank Data as Binary Comparisons

Questionnaire data usually arrives as an explicit rank per item, not as pre-computed pairwise wins/losses. `rank_to_binary` converts one block's ranks into the binary comparison codes the model needs (by default, higher value = ranked first = "wins" the comparison):

```{r rank_to_binary}
set.seed(12345)
mydata <- data.frame(
  i1 = round(rnorm(10, mean = 2, sd = 1), 2),
  i2 = round(rnorm(10, mean = 2, sd = 1), 2),
  i3 = round(rnorm(10, mean = 2, sd = 1), 2)
)
rank_to_binary(mydata, items = 3)
```

`rank_df_to_binary` applies this across an entire questionnaire made of multiple same-size blocks in one call:

```{r rank_df_to_binary}
mydata6 <- data.frame(
  i1 = rnorm(10, mean = 2, sd = .5), i2 = rnorm(10, mean = 2, sd = .5), i3 = rnorm(10, mean = 2, sd = .5),
  i4 = rnorm(10, mean = 2, sd = .5), i5 = rnorm(10, mean = 2, sd = .5), i6 = rnorm(10, mean = 2, sd = .5)
)
rank_df_to_binary(mydata6, items = 3)
```

`name_triplet_pairs` generates matching pair labels (`"i1i2"`, `"i1i3"`, `"i2i3"`, ...) for items grouped in consecutive triplets — useful for naming the columns these binary-encoding functions produce:

```{r name_triplet_pairs}
name_triplet_pairs(6)
name_triplet_pairs(4:9)
```

`rank3_to_triplets` runs the *inverse* direction for a 3-item block: given the three pairwise binary comparisons, it recovers each item's rank (1 = best, 3 = worst) within the triplet:

```{r rank3_to_triplets}
binary3 <- rank_to_binary(mydata[, 1:3], items = 3)
rank3_to_triplets(binary3)
```

`response_dimension` and `cfa_icc_index` handle a related bookkeeping problem for *multidimensional* designs: picking out, from a long vector of all item/comparison parameters across several traits, just the subset belonging to one trait pair — and converting item ordering between `lavaan`'s natural order and the order Thurstonian scoring expects.

```{r response_dimension}
response_dimension(response = 1:18, dimensions = 3, items = c(1, 2))
cfa_icc_index(nitems = 18, nfactors = 3)$index_vector
```

## Single-Dimension Item Characteristic Curves and Grid Scoring

`icc_cfa` implements the normal-ogive probability above for one comparison; `compute_icc_thurstonian` evaluates it across a whole vector of comparisons and (optionally) plots the resulting curves via `plot_icc_thurstonian`.

```{r icc,fig.width=8,fig.height=5}
gamma <- c(0.556, -1.253, -1.729, 0.618, 0.937, 0.295, -0.672, -1.127, -0.446, 0.632, 1.147, 0.498)
psi   <- c(2.172, 1.883, 2.055, 1.869, 2.231, 2.100, 1.762, 1.803, 1.565, 1.892, 1.794, 1.686)
lambda <- c(1.082, 1.082, -1.297, -1.297, 0.802, 0.802, 1.083, 1.083)
gamma_d1 <- gamma[response_dimension(1:12, 3, c(1, 2))]
psi_d1   <- psi[response_dimension(1:12, 3, c(1, 2))]
eta <- seq(-6, 6, by = 0.1)
result_icc <- compute_icc_thurstonian(eta = eta, gamma = gamma_d1, lambda = lambda, psi = psi_d1, plot = TRUE)
result_icc$plot
```

Each curve is one comparison's probability of resolving in the "positive" direction as trait level rises — comparisons with steeper curves (larger `lambda`) are more informative about the trait.

`compute_ability` scores a single response pattern by evaluating the likelihood (times a normal prior, via `compute_map`) over a grid of `eta` values and taking the grid point with maximum posterior density — a direct, brute-force analogue of the optimization-based MAP scorer described below, useful for a single dimension and small enough grids.

```{r compute_ability,fig.width=8,fig.height=5}
map_prior <- compute_map(eta = eta, mean = 0, sd = 1)
response_all_wrong   <- rep(0, 8)
response_all_right   <- rep(1, 8)
response_alternating <- rep(c(1, 0), 4)

r1 <- compute_ability(response_all_wrong, eta, gamma_d1, lambda, psi_d1, map = map_prior, plot = FALSE)
r2 <- compute_ability(response_all_right, eta, gamma_d1, lambda, psi_d1, map = map_prior, plot = FALSE)
r3 <- compute_ability(response_alternating, eta, gamma_d1, lambda, psi_d1, map = map_prior, plot = TRUE)

c(all_wrong = r1$ability_map, all_right = r2$ability_map, alternating = r3$ability_map)
```

As expected, the all-wrong pattern gives the lowest trait estimate, all-right the highest, and the alternating (mixed) pattern lands near the middle. `compute_scores` applies `compute_ability` row-by-row across an entire data frame of response patterns to score every respondent at once.

## Modern MAP Scoring From a Fitted `lavaan` Model

The functions above assume item parameters are supplied directly. In practice, parameters usually come from fitting a Thurstonian model with `lavaan` via the `thurstonianIRT` package. `extract_tirt_params` pulls and aligns the parameter blocks (`Lambda`, `theta_diag`, `tau`, `nu`, `Psi`) needed for scoring; `score_tirt_pattern`/`score_tirt` then find the MAP trait estimate by direct BFGS optimization of the posterior described in the theory section, rather than grid search — scaling to many traits and many items at once.

```{r tirt_fit,results='hide'}
data("triplets")
blocks <- set_block(c("i1", "i2", "i3"), traits = c("t1", "t2", "t3"), signs = c(1, 1, 1)) +
  set_block(c("i4", "i5", "i6"), traits = c("t1", "t2", "t3"), signs = c(-1, 1, 1)) +
  set_block(c("i7", "i8", "i9"), traits = c("t1", "t2", "t3"), signs = c(1, 1, -1)) +
  set_block(c("i10", "i11", "i12"), traits = c("t1", "t2", "t3"), signs = c(1, -1, 1))
triplets_long <- make_TIRT_data(data = triplets, blocks = blocks, direction = "larger", format = "pairwise", family = "bernoulli", range = c(0, 1))
fit_tirt <- fit_TIRT_lavaan(triplets_long)
pars <- extract_tirt_params(fit_tirt)
```

```{r tirt_score}
patterns <- as.matrix(triplets)
scores_rwf <- score_tirt(patterns, lambda = pars$lambda, theta_diag = pars$theta_diag, tau = pars$tau, Psi = pars$Psi, nu = NULL)
scores_reference <- predict(fit_tirt)
head(reshape2::recast(scores_reference, formula = id ~ trait, id.var = 1:2))
head(scores_rwf)
```

`score_tirt`'s output matches `thurstonianIRT::predict()`'s scores closely — confirming the hand-rolled BFGS optimizer recovers the same MAP estimates as the reference package, using only the extracted `lavaan` parameters.

### The linear-algebra helper underneath

`score_tirt_pattern` needs $\Psi^{-1}$ (the prior precision matrix); `compute_solve` is a base-R Gauss-Jordan elimination implementation used in place of `solve()`:

```{r compute_solve}
A <- matrix(c(2, 1, 1, 3), nrow = 2, byrow = TRUE)
b <- c(1, 2)
x_mine <- compute_solve(A, b)
x_base <- solve(A, b)
rbind(mine = x_mine, base = x_base)

# matrix inverse when b is omitted
inv_mine <- compute_solve(A)
inv_base <- solve(A)
all.equal(inv_mine, inv_base, check.attributes = FALSE)
```

## Diagnosing the Underlying `lavaan` Fit

Because Thurstonian scoring depends entirely on a `lavaan` fit's estimated parameters, a poorly-behaved fit (a "Heywood case" — negative variances, out-of-range standardized loadings or correlations, non-convergence) will silently corrupt every downstream score. `check_heywood` screens for exactly these problems.

```{r check_heywood}
diagnostics <- check_heywood(fit_tirt$fit, verbose = TRUE)
diagnostics$has_issues
```

Run this *before* trusting any scores produced by `extract_tirt_params`/`score_tirt` — a model with Heywood cases needs to be re-specified (e.g. simplifying the trait structure, fixing an offending residual variance) rather than scored as-is.

---

## Conclusion

`GLM_IRT_T.R` covers the full Thurstonian IRT pipeline: constructing a forced-choice block design (`generate_matrix_A` and friends), encoding raw rank data into the binary comparisons the model needs (`rank_to_binary`, `rank_df_to_binary`), a from-scratch single-dimension item characteristic curve and grid-based scorer for understanding the model mechanics (`icc_cfa`, `compute_ability`), and a modern multi-trait MAP scorer that reproduces `thurstonianIRT::predict()` directly from a fitted `lavaan` model's own parameters (`extract_tirt_params`, `score_tirt`) — with a Heywood-case check (`check_heywood`) to catch improper solutions before they're scored.

---

*Rendered with R `r getRversion()` · thurstonianIRT `r as.character(packageVersion("thurstonianIRT"))` · lavaan `r as.character(packageVersion("lavaan"))`*
