---
title: "GLM KRUSKAL WALLIS"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM KRUSKAL WALLIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
load(paste0(directory, "data/df_blood_pressure.rda"))
```

## Description

This shows the output of `compute_kruskal_wallis_test` from the package rwf, which runs the Kruskal-Wallis rank-sum test (Kruskal & Wallis, 1952), the non-parametric alternative to the one-way ANOVA. The function returns the test statistic $H$ with its degrees of freedom and p value, two effect sizes (rank eta-squared $\eta^2_H$ and rank epsilon-squared $\varepsilon^2_R$) and, optionally, bootstrap confidence intervals for both effect sizes. Each section gives the formulas the function uses, computes them by hand and checks the results against `stats`, `coin`, `rcompanion`, `rstatix` and `effectsize`.

The computation of $H$, including the correction for ties, is adapted from `stats::kruskal.test` (R Core Team, 2026), and the effect sizes use the same formulas as `rstatix::kruskal_effsize` (Kassambara, 2026) and `effectsize::rank_epsilon_squared` (Ben-Shachar, Lüdecke & Makowski, 2020).

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/GLM_ANOVA.R)

## Notation

There are $k$ independent groups. Group $j$ has $n_j$ observations and the total sample size is $n=\sum_{j=1}^{k}n_j$. All $n$ observations are pooled and ranked from 1 to $n$. Tied values get the mean of the ranks they span (mid-ranks). $r_{ij}$ is the rank of observation $i$ in group $j$, $R_j=\sum_{i=1}^{n_j}r_{ij}$ is the rank sum of group $j$, $\bar{r}_j=R_j/n_j$ is its mean rank and $\bar{r}=(n+1)/2$ is the mean of all ranks. $t_1,\dots,t_m$ are the sizes of the $m$ groups of tied values.

## The Kruskal-Wallis test

The Kruskal-Wallis test is the rank-based alternative to the one-way ANOVA. It replaces the observations with their ranks and asks whether the mean ranks differ between groups more than they would by chance. It does not assume normality, so it suits skewed or ordinal outcomes. Its null hypothesis is that all groups come from the same distribution.

### The H statistic

Without ties the statistic is

$$H_0=\frac{12}{n(n+1)}\sum_{j=1}^{k}\frac{R_j^2}{n_j}-3(n+1)$$

Ties shrink the variance of the ranks, so $H_0$ is divided by a correction factor:

$$C=1-\frac{\sum_{l=1}^{m}\left(t_l^3-t_l\right)}{n^3-n},\qquad H=\frac{H_0}{C}$$

Values that occur once have $t_l=1$ and add nothing to the sum, so $C=1$ when there are no ties.

### Degrees of freedom and p value

Under the null hypothesis $H$ approximately follows a chi-squared distribution with $k-1$ degrees of freedom:

$$df=k-1,\qquad p=P\left(\chi^2_{k-1}\ge H\right)$$

### H as an ANOVA on the ranks

The tie-corrected $H$ can also be written with sums of squares of the ranks:

$$H=(n-1)\,\frac{SS_{\text{between}}}{SS_{\text{total}}}=(n-1)\,\frac{\sum_{j=1}^{k}n_j\left(\bar{r}_j-\bar{r}\right)^2}{\sum_{j=1}^{k}\sum_{i=1}^{n_j}\left(r_{ij}-\bar{r}\right)^2}$$

So the Kruskal-Wallis test is a one-way ANOVA run on the ranks. $SS_{\text{between}}/SS_{\text{total}}$ is the share of rank variance explained by the groups. This is where the effect sizes below come from.

## When to use it and assumptions

### When to use it

Use the Kruskal-Wallis test to compare two or more independent groups on one outcome when the one-way ANOVA is not a good fit:

- **The outcome is ordinal**, such as Likert items, rankings or grades, so means and variances are not meaningful.
- **The outcome is continuous but clearly not normal within groups**, for example strongly skewed reaction times or incomes, and the groups are small. With large groups the ANOVA is fairly robust to non-normality.
- **There are outliers** that would dominate the group means. Ranks limit the influence of any single value.

Do not use it when:

- **The data are normal with equal variances.** The ANOVA is then slightly more powerful. The asymptotic relative efficiency of Kruskal-Wallis against the ANOVA under normality is $3/\pi\approx0.955$, so little is lost by using it (Hollander, Wolfe & Chicken, 2014).
- **The same participants are measured more than once.** Use the Friedman test for repeated measures, `compute_friedman_test`, which reports Kendall's $W$ as its effect size.
- **The groups have different spreads and the question is about means.** Use the Welch ANOVA (`compute_one_way_test(var.equal = FALSE)`). Kruskal-Wallis is not a fix for unequal variances.
- **The design has several factors or covariates.** The test handles only one grouping variable.

With two groups the Kruskal-Wallis test is the Wilcoxon-Mann-Whitney rank-sum test. Both give the same p value when the Wilcoxon test uses the normal approximation without continuity correction:

```{r two_groups}
stats::kruskal.test(bp_before ~ sex, data = df_blood_pressure)$p.value
stats::wilcox.test(bp_before ~ sex, data = df_blood_pressure, exact = FALSE, correct = FALSE)$p.value
```

### Assumptions

1. **Independent observations.** Each observation belongs to one group only and does not depend on any other observation.
2. **An outcome that is at least ordinal.** The values must be orderable for the ranks to mean anything.
3. **Similar shape and spread across groups, to compare medians.** If the group distributions have the same shape and differ only in location, a significant result means the medians differ. If shapes or spreads differ, the test still detects whether some groups tend to have larger values than others (stochastic dominance), but it is no longer a test of medians. Unequal spreads with equal medians can also make it reject the null hypothesis too often.
4. **Groups large enough for the chi-squared approximation.** A common rule is at least 5 observations per group. With smaller groups use an exact or permutation p value, such as `coin::kruskal_test(distribution = coin::approximate(nresample = 10000))`.

Normality and equal variances are **not** assumed.

The assumptions in `df_blood_pressure`: 40 independent patients per age group, and an outcome measured in whole numbers. The boxplots and the Fligner-Killeen test of equal spread check whether the groups have a similar shape and spread:

```{r assumptions,fig.width=7,fig.height=5}
boxplot(bp_before ~ agegrp, data = df_blood_pressure, xlab = "age group", ylab = "blood pressure before")
table(df_blood_pressure$agegrp)
stats::fligner.test(bp_before ~ agegrp, data = df_blood_pressure)
```

### After a significant result

A significant $H$ means that at least one group differs from the others, not which ones. Follow it with pairwise comparisons on the ranks, such as Dunn's test (Dunn, 1964), with an adjustment for multiple comparisons.

## Why effect sizes

A p value answers one question: if the groups did not differ, how surprising would data like these be? It does not say how large the difference is. Because $H$ grows with the sample size, a trivial difference becomes "significant" in a large sample, and a large difference can be "non-significant" in a small one.

An effect size answers a different question: how much do the groups differ? Effect sizes are useful because they:

- **measure magnitude.** They separate practical importance from statistical significance.
- **do not grow with $n$.** The same population difference gives roughly the same effect size in a study of 30 people or 3,000.
- **can be compared across studies.** This is what meta-analysis combines.
- **drive power analysis.** The sample size a new study needs depends on the effect size it expects.
- **are expected in reports.** The APA Task Force on Statistical Inference recommends reporting effect sizes for primary outcomes (Wilkinson & Task Force on Statistical Inference, 1999; see also Lakens, 2013).

The [simulation below](#p-values-shrink-effect-sizes-do-not) shows this with data.

## Effect sizes for Kruskal-Wallis

### Rank epsilon-squared

$$\varepsilon^2_R=\frac{H}{\left(n^2-1\right)/(n+1)}=\frac{H}{n-1}=\frac{SS_{\text{between}}}{SS_{\text{total}}}$$

$\varepsilon^2_R$ is the proportion of rank variance explained by group membership, the $R^2$ of a one-way ANOVA on the ranks (King, Rosopa & Minium, 2018). It lies in $[0,1]$. Because the groups' mean ranks never match exactly in a sample, it is biased upwards. Under the null hypothesis $E[H]\approx k-1$, so $E[\varepsilon^2_R]\approx(k-1)/(n-1)$ rather than 0.

### Rank eta-squared

$$\eta^2_H=\frac{H-k+1}{n-k}$$

$\eta^2_H$ (Tomczak & Tomczak, 2014) subtracts $k-1$, the expected value of $H$ under the null hypothesis, before scaling. This removes most of the bias: it is close to 0 on average when there is no effect. It can therefore be **negative** when $H<k-1$. rwf returns the negative value as it is, while `rstatix` and `effectsize` truncate it at 0.

### Interpretation

Multiplied by 100, both measures read as the percentage of rank variance explained by the groups. Small, medium and large are the $\eta^2$ benchmarks of Cohen (1988). Sawilowsky (2009) added very small, very large and huge benchmarks for Cohen's $d$, and the table converts them to the $\eta^2$ scale with $\eta^2=d^2/(d^2+4)$:

| Magnitude | $\eta^2_H$, $\varepsilon^2_R$ | Cohen's $d$ |
|---|---|---|
| tiny | < 0.01 | < 0.2 |
| small | 0.01 to < 0.06 | 0.2 |
| medium | 0.06 to < 0.14 | 0.5 |
| large | 0.14 to < 0.26 | 0.8 |
| very large | 0.26 to < 0.50 | 1.2 |
| huge | ≥ 0.50 | 2.0 |

```{r cutoffs}
d <- c(0.2, 0.5, 0.8, 1.2, 2)
data.frame(d, eta_squared = d^2 / (d^2 + 4))
```

These are rough guides. What counts as a meaningful effect depends on the field and the outcome.

## Data

`df_blood_pressure` records the blood pressure of 120 patients in three age groups of 40. The formula `bp_before ~ agegrp` asks whether blood pressure before treatment differs between age groups. Blood pressure is recorded in whole numbers, so there are many ties.

```{r data}
head(df_blood_pressure)
form <- formula(bp_before ~ agegrp)
table(df_blood_pressure$agegrp)
```

## compute_kruskal_wallis_test

```{r compute_kruskal_wallis_test}
result <- compute_kruskal_wallis_test(formula = form, df = df_blood_pressure)
result
```

The output columns are:

| Column | Formula |
|---|---|
| `formula` | the model formula |
| `method` | the test name |
| `etasq` | $\eta^2_H=(H-k+1)/(n-k)$ |
| `epsilonsq` | $\varepsilon^2_R=H/(n-1)$ |
| `H` | tie-corrected Kruskal-Wallis statistic |
| `df` | $k-1$ |
| `p` | $P\left(\chi^2_{k-1}\ge H\right)$ |

## Step by step

Every quantity from the formulas above, computed by hand.

```{r step_ranks}
x <- df_blood_pressure$bp_before
g <- factor(df_blood_pressure$agegrp)
n <- length(x)
k <- nlevels(g)
r <- rank(x)
data.frame(n_j = tapply(r, g, length),
           R_j = tapply(r, g, sum),
           mean_rank = tapply(r, g, mean))
```

$H$ without ties, the tie correction and the corrected $H$:

```{r step_H}
R_j <- tapply(r, g, sum)
n_j <- tapply(r, g, length)
H_0 <- 12 / (n * (n + 1)) * sum(R_j^2 / n_j) - 3 * (n + 1)
t_l <- table(x)
C <- 1 - sum(t_l^3 - t_l) / (n^3 - n)
H <- H_0 / C
data.frame(H_0, C, H)
```

The same $H$ from the sums of squares of the ranks:

```{r step_H_ss}
ss_between <- sum(n_j * (tapply(r, g, mean) - mean(r))^2)
ss_total <- sum((r - mean(r))^2)
(n - 1) * ss_between / ss_total
```

Degrees of freedom, p value and effect sizes:

```{r step_effect_sizes}
by_hand <- data.frame(etasq = (H - k + 1) / (n - k),
                      epsilonsq = H / (n - 1),
                      H = H,
                      df = k - 1,
                      p = pchisq(H, k - 1, lower.tail = FALSE))
by_hand
all.equal(by_hand, result[, c("etasq", "epsilonsq", "H", "df", "p")], check.attributes = FALSE)
```

## Checks against other packages

### stats and coin

`stats::kruskal.test` and `coin::kruskal_test` give the same $H$, degrees of freedom and p value:

```{r check_stats}
stats::kruskal.test(formula = form, data = df_blood_pressure)
coin::kruskal_test(bp_before ~ factor(agegrp), data = df_blood_pressure)
```

### ANOVA on the ranks

$\varepsilon^2_R$ is the $R^2$ of a linear model of the ranks on the groups:

```{r check_lm}
summary(lm(rank(bp_before) ~ agegrp, data = df_blood_pressure))$r.squared
result$epsilonsq
```

### rcompanion, rstatix and effectsize

`rcompanion::epsilonSquared`, `rstatix::kruskal_effsize` and `effectsize::rank_epsilon_squared` compute $\varepsilon^2_R$. `rstatix::kruskal_effsize` (its default method) and `effectsize::rank_eta_squared` compute $\eta^2_H$. The confidence intervals are bootstrapped, so they change slightly from run to run. The [confidence intervals](#confidence-intervals) section compares them with the intervals from rwf.

```{r check_effect_sizes}
rcompanion::epsilonSquared(x = df_blood_pressure$bp_before, g = df_blood_pressure$agegrp,
                           group = "row", ci = TRUE, conf = 0.95, type = "perc", R = 1000, digits = 3)
rstatix::kruskal_effsize(df_blood_pressure, form, ci = TRUE, conf.level = 0.95, ci.type = "perc", nboot = 100)
rstatix::kruskal_effsize(df_blood_pressure, form, method = "epsilon2")
effectsize::rank_epsilon_squared(form, data = df_blood_pressure)
effectsize::rank_eta_squared(form, data = df_blood_pressure)
```

### A test function

`check_kruskal_wallis` puts each rwf value next to the value from the other packages and returns the difference. `rstatix` and `effectsize` truncate $\eta^2_H$ at 0, so the rwf value is truncated in the same way before the comparison.

```{r check_function}
check_kruskal_wallis <- function(formula, df) {
  y <- all.vars(formula)[1]
  g <- all.vars(formula)[2]
  df <- stats::na.omit(data.frame(y = df[[y]], g = factor(df[[g]])))
  rwf <- compute_kruskal_wallis_test(formula = y ~ g, df = df)
  kw <- stats::kruskal.test(y ~ g, data = df)
  rwf_etasq <- max(0, min(1, rwf$etasq))
  comparison <- data.frame(
    statistic = c("H", "H", "df", "p", "epsilonsq", "epsilonsq", "epsilonsq", "epsilonsq", "etasq", "etasq"),
    package = c("stats::kruskal.test", "coin::kruskal_test", "stats::kruskal.test", "stats::kruskal.test",
                "lm on ranks", "rcompanion::epsilonSquared", "rstatix::kruskal_effsize", "effectsize::rank_epsilon_squared",
                "rstatix::kruskal_effsize", "effectsize::rank_eta_squared"),
    rwf = c(rwf$H, rwf$H, rwf$df, rwf$p,
            rwf$epsilonsq, rwf$epsilonsq, rwf$epsilonsq, rwf$epsilonsq,
            rwf_etasq, rwf_etasq),
    other = c(unname(kw$statistic),
              unname(coin::statistic(coin::kruskal_test(y ~ g, data = df))),
              unname(kw$parameter),
              kw$p.value,
              summary(lm(rank(y) ~ g, data = df))$r.squared,
              unname(rcompanion::epsilonSquared(x = df$y, g = df$g, digits = 15)),
              rstatix::kruskal_effsize(df, y ~ g, method = "epsilon2")$effsize,
              effectsize::rank_epsilon_squared(y ~ g, data = df, ci = NULL)$rank_epsilon_squared,
              rstatix::kruskal_effsize(df, y ~ g)$effsize,
              effectsize::rank_eta_squared(y ~ g, data = df, ci = NULL)$rank_eta_squared))
  comparison$difference <- comparison$rwf - comparison$other
  comparison
}
check_kruskal_wallis(formula = form, df = df_blood_pressure)
```

The same check on six datasets with different numbers of groups, sample sizes and amounts of ties. `max_abs_difference` is the largest absolute difference across all the comparisons for that dataset:

```{r check_datasets}
datasets <- list(
  list(formula = bp_before ~ agegrp, df = df_blood_pressure),
  list(formula = qsec ~ cyl, df = mtcars),
  list(formula = weight ~ group, df = PlantGrowth),
  list(formula = count ~ spray, df = InsectSprays),
  list(formula = weight ~ feed, df = chickwts),
  list(formula = len ~ dose, df = ToothGrowth)
)
do.call(rbind, lapply(datasets, function(d) {
  rwf <- compute_kruskal_wallis_test(formula = d$formula, df = d$df)
  check <- check_kruskal_wallis(formula = d$formula, df = d$df)
  data.frame(formula = deparse(d$formula),
             n = nrow(d$df),
             k = rwf$df + 1,
             H = rwf$H,
             etasq = rwf$etasq,
             epsilonsq = rwf$epsilonsq,
             max_abs_difference = max(abs(check$difference)))
}))
```

## Confidence intervals

With `ci = TRUE` the function adds percentile bootstrap confidence intervals for both effect sizes. It draws `nboot` samples of $n$ rows with replacement and computes $\eta^2_H$ and $\varepsilon^2_R$ in each, giving $\hat{\theta}^{*}_1,\dots,\hat{\theta}^{*}_B$. The interval is

$$\left[\hat{\theta}^{*}_{(\alpha/2)},\ \hat{\theta}^{*}_{(1-\alpha/2)}\right],\qquad \alpha=1-\texttt{conf.level}$$

where $\hat{\theta}^{*}_{(q)}$ is the $q$ quantile of the bootstrap estimates. Unlike `rstatix` and `effectsize`, rwf does not truncate the bootstrap values of $\eta^2_H$ at 0, so its lower limit can be negative when the effect is small.

```{r confidence_intervals}
set.seed(1)
compute_kruskal_wallis_test(formula = form, df = df_blood_pressure, ci = TRUE, conf.level = 0.95, nboot = 5000)
```

The other packages use the same percentile method. Each bootstrap draws different resamples, so the limits agree only to about two decimals, and they get closer as `nboot` grows:

```{r confidence_intervals_check}
set.seed(1)
rwf <- compute_kruskal_wallis_test(formula = form, df = df_blood_pressure, ci = TRUE, nboot = 5000)
set.seed(1)
rcompanion_epsilonsq <- rcompanion::epsilonSquared(x = df_blood_pressure$bp_before, g = df_blood_pressure$agegrp,
                                                   ci = TRUE, type = "perc", R = 5000, digits = 15)
set.seed(1)
rstatix_etasq <- rstatix::kruskal_effsize(df_blood_pressure, form, ci = TRUE, ci.type = "perc", nboot = 5000)
set.seed(1)
effectsize_epsilonsq <- effectsize::rank_epsilon_squared(form, data = df_blood_pressure, alternative = "two.sided", iterations = 5000)
set.seed(1)
effectsize_etasq <- effectsize::rank_eta_squared(form, data = df_blood_pressure, alternative = "two.sided", iterations = 5000)
data.frame(
  statistic = c("etasq", "etasq", "etasq", "epsilonsq", "epsilonsq", "epsilonsq"),
  package = c("rwf", "rstatix::kruskal_effsize", "effectsize::rank_eta_squared",
              "rwf", "rcompanion::epsilonSquared", "effectsize::rank_epsilon_squared"),
  estimate = c(rwf$etasq, rstatix_etasq$effsize, effectsize_etasq$rank_eta_squared,
               rwf$epsilonsq, rcompanion_epsilonsq$epsilon.squared, effectsize_epsilonsq$rank_epsilon_squared),
  lower = c(rwf$etasq_lower, rstatix_etasq$conf.low, effectsize_etasq$CI_low,
            rwf$epsilonsq_lower, rcompanion_epsilonsq$lower.ci, effectsize_epsilonsq$CI_low),
  upper = c(rwf$etasq_upper, rstatix_etasq$conf.high, effectsize_etasq$CI_high,
            rwf$epsilonsq_upper, rcompanion_epsilonsq$upper.ci, effectsize_epsilonsq$CI_high))
```

`rstatix` rounds its limits to two decimals. `effectsize` reports one-sided intervals by default, so `alternative = "two.sided"` is set here.

## p values shrink, effect sizes do not

Three groups are drawn from normal distributions whose means differ by 0.3 standard deviations, with 10 to 1,000 observations per group. The population effect is the same in every row. As the sample grows, $H$ grows and the p value drops towards 0, while $\eta^2_H$ and $\varepsilon^2_R$ settle around the same value:

```{r simulation_sample_size}
set.seed(1)
do.call(rbind, lapply(c(10, 30, 100, 300, 1000), function(n_per_group) {
  sim <- data.frame(group = rep(c("a", "b", "c"), each = n_per_group),
                    score = rnorm(3 * n_per_group, mean = rep(c(0, 0.3, 0.6), each = n_per_group)))
  cbind(n_per_group, compute_kruskal_wallis_test(formula = score ~ group, df = sim)[, c("H", "p", "etasq", "epsilonsq")])
}))
```

A p value from one study says little about the size of the effect. The effect size does.

## Bias of epsilon-squared

When there is no effect at all, the average $\varepsilon^2_R$ is close to $(k-1)/(n-1)$, while the average $\eta^2_H$ is close to 0. With $k=3$ groups of 10 the bias is $2/29\approx0.069$, which is already above the "medium" benchmark:

```{r simulation_bias}
set.seed(1)
null_effects <- t(replicate(2000, {
  sim <- data.frame(group = rep(c("a", "b", "c"), each = 10), score = rnorm(30))
  unlist(compute_kruskal_wallis_test(formula = score ~ group, df = sim)[, c("etasq", "epsilonsq")])
}))
data.frame(mean_etasq = mean(null_effects[, "etasq"]),
           mean_epsilonsq = mean(null_effects[, "epsilonsq"]),
           expected_epsilonsq = (3 - 1) / (30 - 1),
           proportion_negative_etasq = mean(null_effects[, "etasq"] < 0))
```

With small samples, report $\eta^2_H$ or remember that $\varepsilon^2_R$ is inflated by roughly $(k-1)/(n-1)$.

## Negative eta-squared

If the groups have identical rank distributions, $H=0$ and $\eta^2_H=(0-k+1)/(n-k)$ is negative. rwf returns the negative value, while `rstatix` and `effectsize` report 0:

```{r negative_etasq}
identical_groups <- data.frame(group = rep(c("a", "b", "c"), each = 5), score = rep(1:5, times = 3))
compute_kruskal_wallis_test(formula = score ~ group, df = identical_groups)
rstatix::kruskal_effsize(identical_groups, score ~ group)
effectsize::rank_eta_squared(score ~ group, data = identical_groups, ci = NULL)
```

A negative value means the groups differ less than expected by chance. Report it as 0 or as it is, but do not read it as a negative effect.

## Missing values

`compute_kruskal_wallis_test` removes rows with a missing value in the outcome or the grouping variable before the test, as `stats::kruskal.test` does. Missing values in other columns of `df` are ignored. Two blood pressure values are set to missing here, and both functions use the remaining 118 rows:

```{r missing_values}
df_missing <- df_blood_pressure
df_missing$bp_before[c(1, 50)] <- NA
df_missing$bp_after[2] <- NA
compute_kruskal_wallis_test(formula = form, df = df_missing)
stats::kruskal.test(formula = form, data = df_missing)
check_kruskal_wallis(formula = form, df = df_missing)
```

## References

Ben-Shachar, M. S., Lüdecke, D., & Makowski, D. (2020). effectsize: Estimation of effect size indices and standardized parameters. *Journal of Open Source Software, 5*(56), 2815. https://doi.org/10.21105/joss.02815

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Dunn, O. J. (1964). Multiple comparisons using rank sums. *Technometrics, 6*(3), 241–252. https://doi.org/10.1080/00401706.1964.10490181

Hollander, M., Wolfe, D. A., & Chicken, E. (2014). *Nonparametric statistical methods* (3rd ed.). Wiley.

Kassambara, A. (2026). *rstatix: Pipe-friendly framework for basic statistical tests* (R package version 1.1.0). https://doi.org/10.32614/CRAN.package.rstatix

King, B. M., Rosopa, P. J., & Minium, E. W. (2018). *Statistical reasoning in the behavioral sciences* (7th ed.). Wiley.

Kruskal, W. H., & Wallis, W. A. (1952). Use of ranks in one-criterion variance analysis. *Journal of the American Statistical Association, 47*(260), 583–621. https://doi.org/10.1080/01621459.1952.10483441

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate cumulative science: A practical primer for t-tests and ANOVAs. *Frontiers in Psychology, 4*, 863. https://doi.org/10.3389/fpsyg.2013.00863

R Core Team. (2026). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria. https://doi.org/10.32614/R.manuals

Sawilowsky, S. S. (2009). New effect size rules of thumb. *Journal of Modern Applied Statistical Methods, 8*(2), 597–599. https://doi.org/10.22237/jmasm/1257035100

Tomczak, M., & Tomczak, E. (2014). The need to report effect size estimates revisited. An overview of some recommended measures of effect size. *Trends in Sport Sciences, 1*(21), 19–25.

Wilkinson, L., & Task Force on Statistical Inference. (1999). Statistical methods in psychology journals: Guidelines and explanations. *American Psychologist, 54*(8), 594–604. https://doi.org/10.1037/0003-066X.54.8.594
