---
title: "GLM ONE WAY"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM ONE WAY}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
load(paste0(directory, "data/df_blood_pressure.rda"))
```

## Description

This shows the output of `compute_one_way_test` from the package rwf, which runs the one-way analysis of variance for two or more independent groups, either assuming equal variances (Fisher's F test) or not (Welch's F test, Welch, 1951). The function returns the sums of squares and mean squares, the F statistic with its degrees of freedom and p value, five effect sizes (eta-squared, partial eta-squared, omega-squared, partial omega-squared and Cohen's f) and the observed power. Each section gives the formulas the function uses, computes them by hand and checks the results against `stats`, `car`, `lsr`, `effectsize`, `sjstats` and `pwr`.

The group statistics and the Welch test (the weights, the Welch F statistic and its degrees of freedom) are adapted from `stats::oneway.test` (R Core Team, 2026). The sums of squares, effect sizes and power are added in rwf.

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/GLM_ANOVA.R)

## Notation

There are $k$ independent groups. Group $j$ has $n_j$ observations $y_{1j},\dots,y_{n_jj}$, mean $\bar{y}_j$ and sample variance $s_j^2=\frac{1}{n_j-1}\sum_{i=1}^{n_j}(y_{ij}-\bar{y}_j)^2$. $N=\sum_{j=1}^{k}n_j$ is the total sample size and $\bar{y}$ the mean of all $N$ observations.

## The one-way ANOVA

The one-way ANOVA tests whether the means of $k$ independent groups are equal. It splits the total variability of the outcome into a part between the group means and a part within the groups, and compares the two. Its null hypothesis is $\mu_1=\mu_2=\dots=\mu_k$.

### Assuming equal variances (Fisher)

Sums of squares and degrees of freedom:

$$SS_{\text{effect}}=\sum_{j=1}^{k}n_j\left(\bar{y}_j-\bar{y}\right)^2,\qquad df_{\text{effect}}=k-1$$

$$SS_{\text{error}}=\sum_{j=1}^{k}(n_j-1)\,s_j^2,\qquad df_{\text{error}}=N-k$$

$$SS_{\text{total}}=SS_{\text{effect}}+SS_{\text{error}}=\sum_{j=1}^{k}\sum_{i=1}^{n_j}\left(y_{ij}-\bar{y}\right)^2$$

Mean squares, the F statistic and its p value:

$$MS_{\text{effect}}=\frac{SS_{\text{effect}}}{df_{\text{effect}}},\qquad MS_{\text{error}}=\frac{SS_{\text{error}}}{df_{\text{error}}},\qquad F=\frac{MS_{\text{effect}}}{MS_{\text{error}}}$$

$$p=P\left(F_{df_{\text{effect}},\,df_{\text{error}}}\ge F\right)$$

$MS_{\text{error}}$ pools the variances of all groups into one estimate, which is why this version assumes that the groups have equal variances.

### Not assuming equal variances (Welch)

Welch's test weights each group by the precision of its mean, $w_j=n_j/s_j^2$, so groups with a large variance or few observations count less:

$$W=\sum_{j=1}^{k}w_j,\qquad \bar{y}_w=\frac{1}{W}\sum_{j=1}^{k}w_j\bar{y}_j,\qquad \Lambda=\frac{1}{k^2-1}\sum_{j=1}^{k}\frac{\left(1-w_j/W\right)^2}{n_j-1}$$

$$F_W=\frac{\sum_{j=1}^{k}w_j\left(\bar{y}_j-\bar{y}_w\right)^2}{(k-1)\left(1+2(k-2)\Lambda\right)},\qquad df_{\text{effect}}=k-1,\qquad df_{\text{error}}=\frac{1}{3\Lambda}$$

The error degrees of freedom are usually not a whole number.

Welch's test has no sums of squares of its own. To fill the same output columns, rwf sets

$$MS_{\text{effect}}=\sum_{j=1}^{k}w_j\left(\bar{y}_j-\bar{y}_w\right)^2,\qquad MS_{\text{error}}=(k-1)\left(1+2(k-2)\Lambda\right)$$

and $SS=MS\times df$, so that $F_W=MS_{\text{effect}}/MS_{\text{error}}$. These are not variance components of the data, only quantities that reproduce $F_W$. The effect sizes computed from them are the usual conversions of an F statistic into an effect size ([Welch effect sizes](#welch-effect-sizes)).

## When to use it and assumptions

### When to use it

Use the one-way ANOVA to compare the means of two or more independent groups on an outcome measured on an interval or ratio scale, such as blood pressure, reaction time or a test score.

Prefer Welch's version (`var.equal = FALSE`) unless you have a good reason to believe the variances are equal. It controls the Type I error rate when variances differ and loses very little power when they do not (Delacre, Leys, Mora & Lakens, 2019). The [simulation below](#fisher-and-welch-with-unequal-variances) shows why.

Do not use it when:

- **The outcome is ordinal, or clearly not normal with small groups.** Use the Kruskal-Wallis test, `compute_kruskal_wallis_test`.
- **The same participants are measured more than once.** Use a repeated measures ANOVA, or the Friedman test, `compute_friedman_test`.
- **There are covariates or several factors.** Use an ANCOVA or a factorial ANOVA.

With two groups the one-way ANOVA is the independent samples t test, and $F=t^2$. Fisher's F matches Student's t and Welch's F matches Welch's t, including the degrees of freedom:

```{r two_groups}
fisher <- compute_one_way_test(formula = bp_before ~ sex, df = df_blood_pressure, var.equal = TRUE)
welch <- compute_one_way_test(formula = bp_before ~ sex, df = df_blood_pressure, var.equal = FALSE)
student_t <- stats::t.test(bp_before ~ sex, data = df_blood_pressure, var.equal = TRUE)
welch_t <- stats::t.test(bp_before ~ sex, data = df_blood_pressure, var.equal = FALSE)
data.frame(test = c("Fisher", "Welch"),
           F_rwf = c(fisher$statistic, welch$statistic),
           t_squared = c(student_t$statistic^2, welch_t$statistic^2),
           df_error_rwf = c(fisher$df_error, welch$df_error),
           df_t = c(student_t$parameter, welch_t$parameter))
```

### Assumptions

1. **Independent observations.** Each observation belongs to one group only and does not depend on any other observation.
2. **An interval or ratio outcome.** Means and variances must be meaningful.
3. **Normally distributed residuals within each group.** With moderate or large groups the F test is fairly robust to non-normality, because the group means are close to normal. With small groups and skewed data use the Kruskal-Wallis test.
4. **Equal variances across groups (Fisher only).** Fisher's F is fairly robust when the groups are of equal size. When they are not, the Type I error rate is inflated if the smaller groups have the larger variances, and the test becomes conservative if they have the smaller variances. Welch's F does not make this assumption.
5. **No extreme outliers.** A single extreme value can move a group mean and inflate its variance.

In `df_blood_pressure` there are 40 independent patients per age group. The group standard deviations, the Q-Q plot of the residuals and Levene's test (with the median, the Brown-Forsythe version) check normality and equal variances:

```{r assumptions,fig.width=10,fig.height=5}
data.frame(n = tapply(df_blood_pressure$bp_before, df_blood_pressure$agegrp, length),
           mean = tapply(df_blood_pressure$bp_before, df_blood_pressure$agegrp, mean),
           sd = tapply(df_blood_pressure$bp_before, df_blood_pressure$agegrp, sd))
model <- stats::aov(bp_before ~ agegrp, data = df_blood_pressure)
par(mfrow = c(1, 2))
boxplot(bp_before ~ agegrp, data = df_blood_pressure, xlab = "age group", ylab = "blood pressure before")
stats::qqnorm(stats::residuals(model))
stats::qqline(stats::residuals(model))
stats::shapiro.test(stats::residuals(model))
car::leveneTest(bp_before ~ factor(agegrp), data = df_blood_pressure, center = median)
```

The standard deviations are similar and Levene's test finds no evidence of unequal variances. The Shapiro-Wilk test detects a departure from normality, but the Q-Q plot shows it is mild, and with 40 observations per group the F test is robust to it. Significance tests of assumptions become more sensitive as the sample grows, so judge the size of the departure from the plots, not only the p value.

### After a significant result

A significant $F$ means that at least one group mean differs from the others, not which ones. Follow it with post hoc comparisons: Tukey's HSD when the variances are equal, Games-Howell when they are not. `report_oneway` in rwf returns both.

## Why effect sizes

A p value answers one question: if the group means were equal, how surprising would data like these be? It does not say how large the difference is. Because $F$ grows with the sample size, a trivial difference becomes "significant" in a large sample, and a large difference can be "non-significant" in a small one.

An effect size answers a different question: how much do the groups differ? Effect sizes are useful because they:

- **measure magnitude.** They separate practical importance from statistical significance.
- **do not grow with $N$.** The same population difference gives roughly the same effect size in a study of 30 people or 3,000.
- **can be compared across studies.** This is what meta-analysis combines.
- **drive power analysis.** The sample size a new study needs depends on the effect size it expects.
- **are expected in reports.** The APA Task Force on Statistical Inference recommends reporting effect sizes for primary outcomes (Wilkinson & Task Force on Statistical Inference, 1999; see also Lakens, 2013).

The [simulation below](#p-values-shrink-effect-sizes-do-not) shows this with data.

## Effect sizes

### Eta-squared

$$\eta^2=\frac{SS_{\text{effect}}}{SS_{\text{total}}},\qquad \eta^2_p=\frac{SS_{\text{effect}}}{SS_{\text{effect}}+SS_{\text{error}}}$$

$\eta^2$ is the proportion of the total variance explained by the groups, the $R^2$ of the model. Partial eta-squared $\eta^2_p$ leaves out the variance explained by other factors. A one-way design has no other factors, so $SS_{\text{effect}}+SS_{\text{error}}=SS_{\text{total}}$ and $\eta^2_p=\eta^2$. Both are biased upwards in small samples. Under the null hypothesis $E[\eta^2]\approx(k-1)/(N-1)$ rather than 0.

### Omega-squared

$$\omega^2=\frac{SS_{\text{effect}}-df_{\text{effect}}\,MS_{\text{error}}}{SS_{\text{total}}+MS_{\text{error}}},\qquad \omega^2_p=\frac{df_{\text{effect}}\left(MS_{\text{effect}}-MS_{\text{error}}\right)}{df_{\text{effect}}\,MS_{\text{effect}}+\left(N-df_{\text{effect}}\right)MS_{\text{error}}}$$

$\omega^2$ (Hays, 1963) estimates the proportion of variance explained in the population. It subtracts the part of $SS_{\text{effect}}$ expected by chance, $df_{\text{effect}}\,MS_{\text{error}}$, which removes most of the bias of $\eta^2$. It is negative when $F<1$. Partial omega-squared $\omega^2_p$ (Olejnik & Algina, 2003) is again the same as $\omega^2$ in a one-way design with equal variances, because the two denominators are equal when $N-df_{\text{effect}}=df_{\text{error}}+1$.

### Cohen's f

$$f=\sqrt{\frac{\eta^2}{1-\eta^2}}$$

Cohen's $f$ is the standard deviation of the group means divided by the common within-group standard deviation. It is the effect size used in power analysis (Cohen, 1988).

### Interpretation

Multiplied by 100, $\eta^2$ and $\omega^2$ read as the percentage of variance explained by the groups. Small, medium and large are the benchmarks of Cohen (1988). Sawilowsky (2009) added very small, very large and huge benchmarks for Cohen's $d$, and the table converts them with $\eta^2=d^2/(d^2+4)$ and $f=d/2$, the relations for two groups of equal size:

| Magnitude | $\eta^2$, $\omega^2$ | Cohen's $f$ | Cohen's $d$ |
|---|---|---|---|
| tiny | < 0.01 | < 0.10 | < 0.2 |
| small | 0.01 to < 0.06 | 0.10 to < 0.25 | 0.2 |
| medium | 0.06 to < 0.14 | 0.25 to < 0.40 | 0.5 |
| large | 0.14 to < 0.26 | 0.40 to < 0.60 | 0.8 |
| very large | 0.26 to < 0.50 | 0.60 to < 1.00 | 1.2 |
| huge | ≥ 0.50 | ≥ 1.00 | 2.0 |

```{r cutoffs}
d <- c(0.2, 0.5, 0.8, 1.2, 2)
eta_squared <- d^2 / (d^2 + 4)
data.frame(d, eta_squared, f = sqrt(eta_squared / (1 - eta_squared)))
```

These are rough guides. What counts as a meaningful effect depends on the field and the outcome.

## Observed power

`power` is the probability that an F test at $\alpha=0.05$ rejects the null hypothesis if the population effect were exactly the observed $f$. It uses the noncentral F distribution with noncentrality parameter

$$\lambda=f^2N,\qquad \text{power}=P\left(F_{df_{\text{effect}},\,df_{\text{error}},\,\lambda}>F_{0.95;\,df_{\text{effect}},\,df_{\text{error}}}\right)$$

where $F_{0.95;\,df_{\text{effect}},\,df_{\text{error}}}$ is the critical value of the central F distribution.

Observed power adds no information beyond the p value (Hoenig & Heisey, 2001). Both come from the same $F$, so a study with $p=0.05$ always has an observed power of about 0.5, and a non-significant result always has low observed power. It cannot show that a non-significant study "had enough power". Each point below is one simulated study:

```{r observed_power,fig.width=7,fig.height=5}
set.seed(1)
studies <- do.call(rbind, lapply(1:500, function(i) {
  sim <- data.frame(group = rep(c("a", "b", "c"), each = 20),
                    score = rnorm(60, mean = rep(c(0, 0, runif(1, 0, 1)), each = 20)))
  compute_one_way_test(formula = score ~ group, df = sim)[, c("p", "power")]
}))
plot(studies$p, studies$power, xlab = "p value", ylab = "observed power", pch = 16, col = "grey40")
abline(v = 0.05, h = 0.5, lty = 2)
```

Power is useful **before** a study, with an effect size chosen from theory or earlier studies rather than from the data. For example, the number of participants per group needed to detect a medium effect ($f=0.25$) with three groups and 80% power:

```{r power_planning}
pwr::pwr.anova.test(k = 3, f = 0.25, sig.level = 0.05, power = 0.80)
```

## Data

`df_blood_pressure` records the blood pressure of 120 patients in three age groups of 40. The formula `bp_before ~ agegrp` asks whether mean blood pressure before treatment differs between age groups.

```{r data}
head(df_blood_pressure)
form <- formula(bp_before ~ agegrp)
table(df_blood_pressure$agegrp)
```

## compute_one_way_test

```{r compute_one_way_test}
fisher <- compute_one_way_test(formula = form, df = df_blood_pressure, var.equal = TRUE)
welch <- compute_one_way_test(formula = form, df = df_blood_pressure, var.equal = FALSE)
rbind(fisher, welch)
```

The output columns are:

| Column | Fisher (`var.equal = TRUE`) | Welch (`var.equal = FALSE`) |
|---|---|---|
| `formula`, `method` | model formula and test | model formula and test |
| `ss_effect`, `ss_error` | $SS_{\text{effect}}$, $SS_{\text{error}}$ | $MS\times df$ (not variance components) |
| `ms_effect`, `ms_error` | $MS_{\text{effect}}$, $MS_{\text{error}}$ | numerator and denominator of $F_W$ |
| `etasq`, `partial.etasq` | $\eta^2$, $\eta^2_p$ | $F\,df_{\text{effect}}/(F\,df_{\text{effect}}+df_{\text{error}})$ |
| `omegasq`, `partial.omegasq` | $\omega^2$, $\omega^2_p$ | see [Welch effect sizes](#welch-effect-sizes) |
| `cohens.f` | $\sqrt{\eta^2/(1-\eta^2)}$ | $\sqrt{F\,df_{\text{effect}}/df_{\text{error}}}$ |
| `power` | observed power, $\lambda=f^2N$ | observed power, $\lambda=f^2N$ |
| `statistic` | $F$ | $F_W$ |
| `df_effect`, `df_error` | $k-1$, $N-k$ | $k-1$, $1/(3\Lambda)$ |
| `p` | p value | p value |

## Step by step

### Fisher

Every quantity from the formulas above, computed by hand:

```{r step_fisher}
y <- df_blood_pressure$bp_before
g <- factor(df_blood_pressure$agegrp)
k <- nlevels(g)
n_j <- tapply(y, g, length)
mean_j <- tapply(y, g, mean)
var_j <- tapply(y, g, var)
N <- sum(n_j)
ss_effect <- sum(n_j * (mean_j - mean(y))^2)
ss_error <- sum((n_j - 1) * var_j)
ss_total <- sum((y - mean(y))^2)
df_effect <- k - 1
df_error <- N - k
ms_effect <- ss_effect / df_effect
ms_error <- ss_error / df_error
F_value <- ms_effect / ms_error
etasq <- ss_effect / ss_total
omegasq <- (ss_effect - df_effect * ms_error) / (ss_total + ms_error)
cohens_f <- sqrt(etasq / (1 - etasq))
power <- pf(qf(0.95, df_effect, df_error), df_effect, df_error, ncp = cohens_f^2 * N, lower.tail = FALSE)
by_hand <- data.frame(ss_effect, ss_error, ms_effect, ms_error,
                      etasq, partial.etasq = ss_effect / (ss_effect + ss_error),
                      omegasq, partial.omegasq = df_effect * (ms_effect - ms_error) / (df_effect * ms_effect + (N - df_effect) * ms_error),
                      cohens.f = cohens_f, power,
                      statistic = F_value, df_effect, df_error,
                      p = pf(F_value, df_effect, df_error, lower.tail = FALSE))
by_hand
c(ss_total = ss_total, ss_effect_plus_error = ss_effect + ss_error)
all.equal(by_hand, fisher[, names(by_hand)], check.attributes = FALSE)
```

### Welch

```{r step_welch}
w_j <- n_j / var_j
W <- sum(w_j)
mean_w <- sum(w_j * mean_j) / W
Lambda <- sum((1 - w_j / W)^2 / (n_j - 1)) / (k^2 - 1)
F_welch <- sum(w_j * (mean_j - mean_w)^2) / ((k - 1) * (1 + 2 * (k - 2) * Lambda))
df_error_welch <- 1 / (3 * Lambda)
data.frame(F = F_welch, df_effect = k - 1, df_error = df_error_welch,
           p = pf(F_welch, k - 1, df_error_welch, lower.tail = FALSE))
welch[, c("statistic", "df_effect", "df_error", "p")]
```

## Checks against other packages

### stats and car

`stats::oneway.test` gives $F$, the degrees of freedom and the p value for both versions. `stats::aov` and `car::Anova` give the sums of squares and mean squares of the Fisher version:

```{r check_stats}
stats::oneway.test(formula = form, data = df_blood_pressure, var.equal = TRUE)
stats::oneway.test(formula = form, data = df_blood_pressure, var.equal = FALSE)
summary(stats::aov(form, data = df_blood_pressure))
car::Anova(stats::aov(form, data = df_blood_pressure), type = 2)
```

### Effect sizes and power

`lsr::etaSquared` and `effectsize` give $\eta^2$, $\omega^2$ and Cohen's $f$, `sjstats::anova_stats` gives all the effect sizes and the observed power, and `pwr::pwr.anova.test` gives the power for groups of equal size:

```{r check_effect_sizes}
lsr::etaSquared(stats::aov(form, data = df_blood_pressure), type = 3, anova = TRUE)
effectsize::eta_squared(stats::aov(form, data = df_blood_pressure), partial = FALSE, ci = NULL)
effectsize::omega_squared(stats::aov(form, data = df_blood_pressure), partial = FALSE, ci = NULL)
effectsize::cohens_f(stats::aov(form, data = df_blood_pressure), ci = NULL)
sjstats::anova_stats(stats::lm(form, data = df_blood_pressure), digits = 22)
pwr::pwr.anova.test(k = 3, n = 40, f = fisher$cohens.f, sig.level = 0.05)$power
fisher$power
```

### Welch effect sizes

For Welch's test, `etasq`, `omegasq` and `cohens.f` are the conversions of the F statistic into effect sizes with the Welch degrees of freedom, the same as `effectsize::F_to_eta2`, `F_to_omega2` and `F_to_f`:

$$\eta^2=\frac{F\,df_{\text{effect}}}{F\,df_{\text{effect}}+df_{\text{error}}},\qquad \omega^2=\frac{df_{\text{effect}}(F-1)}{F\,df_{\text{effect}}+df_{\text{error}}+1},\qquad f=\sqrt{\frac{F\,df_{\text{effect}}}{df_{\text{error}}}}$$

```{r check_welch_effect_sizes}
data.frame(
  statistic = c("etasq", "omegasq", "cohens.f"),
  rwf = c(welch$etasq, welch$omegasq, welch$cohens.f),
  effectsize = c(effectsize::F_to_eta2(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Eta2_partial,
                 effectsize::F_to_omega2(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Omega2_partial,
                 effectsize::F_to_f(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Cohens_f_partial))
```

`partial.omegasq` for Welch uses the actual sample size $N$ in place of $df_{\text{effect}}+df_{\text{error}}+1$, so it differs from `omegasq` (`r round(welch$partial.omegasq, 3)` against `r round(welch$omegasq, 3)` here). These Welch effect sizes are approximations; report `omegasq` or `etasq` with the Welch degrees of freedom, and treat `partial.omegasq` with caution.

### A test function

`check_one_way` puts each rwf value next to the value from the other packages and returns the difference. The Fisher version is compared with `stats`, `effectsize` and `sjstats`, the Welch version with `stats` and `effectsize`.

```{r check_function}
check_one_way <- function(formula, df) {
  vars <- all.vars(formula)
  df <- stats::na.omit(data.frame(y = df[[vars[1]]], g = factor(df[[vars[2]]])))
  fisher <- compute_one_way_test(formula = y ~ g, df = df, var.equal = TRUE)
  welch <- compute_one_way_test(formula = y ~ g, df = df, var.equal = FALSE)
  model <- stats::aov(y ~ g, data = df)
  anova_table <- summary(model)[[1]]
  oneway_fisher <- stats::oneway.test(y ~ g, data = df, var.equal = TRUE)
  oneway_welch <- stats::oneway.test(y ~ g, data = df, var.equal = FALSE)
  sjstats_table <- as.data.frame(sjstats::anova_stats(stats::lm(y ~ g, data = df), digits = 22))
  comparison <- data.frame(
    test = c(rep("Fisher", 11), rep("Welch", 6)),
    statistic = c("ss_effect", "ss_error", "statistic", "df_error", "p", "etasq", "omegasq", "partial.omegasq", "cohens.f", "power", "etasq",
                  "statistic", "df_error", "p", "etasq", "omegasq", "cohens.f"),
    package = c("stats::aov", "stats::aov", "stats::oneway.test", "stats::oneway.test", "stats::oneway.test",
                "effectsize::eta_squared", "effectsize::omega_squared", "sjstats::anova_stats", "effectsize::cohens_f",
                "sjstats::anova_stats", "lsr::etaSquared",
                "stats::oneway.test", "stats::oneway.test", "stats::oneway.test",
                "effectsize::F_to_eta2", "effectsize::F_to_omega2", "effectsize::F_to_f"),
    rwf = c(fisher$ss_effect, fisher$ss_error, fisher$statistic, fisher$df_error, fisher$p,
            fisher$etasq, fisher$omegasq, fisher$partial.omegasq, fisher$cohens.f, fisher$power, fisher$etasq,
            welch$statistic, welch$df_error, welch$p, welch$etasq, welch$omegasq, welch$cohens.f),
    other = c(anova_table[1, "Sum Sq"], anova_table[2, "Sum Sq"],
              unname(oneway_fisher$statistic), unname(oneway_fisher$parameter[2]), oneway_fisher$p.value,
              effectsize::eta_squared(model, partial = FALSE, ci = NULL)$Eta2,
              effectsize::omega_squared(model, partial = FALSE, ci = NULL)$Omega2,
              sjstats_table$partial.omegasq[1],
              effectsize::cohens_f(model, ci = NULL, verbose = FALSE)$Cohens_f,
              sjstats_table$power[1],
              lsr::etaSquared(model)[1, "eta.sq"],
              unname(oneway_welch$statistic), unname(oneway_welch$parameter[2]), oneway_welch$p.value,
              effectsize::F_to_eta2(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Eta2_partial,
              effectsize::F_to_omega2(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Omega2_partial,
              effectsize::F_to_f(welch$statistic, welch$df_effect, welch$df_error, ci = NULL)$Cohens_f_partial))
  comparison$difference <- comparison$rwf - comparison$other
  comparison
}
check_one_way(formula = form, df = df_blood_pressure)
```

The same check on seven datasets with different numbers of groups, equal and unequal group sizes, and equal and unequal variances. `max_abs_difference` is the largest absolute difference across all the comparisons for that dataset:

```{r check_datasets}
datasets <- list(
  list(formula = bp_before ~ agegrp, df = df_blood_pressure),
  list(formula = qsec ~ cyl, df = mtcars),
  list(formula = weight ~ group, df = PlantGrowth),
  list(formula = count ~ spray, df = InsectSprays),
  list(formula = weight ~ feed, df = chickwts),
  list(formula = len ~ dose, df = ToothGrowth),
  list(formula = Sepal.Length ~ Species, df = iris)
)
do.call(rbind, lapply(datasets, function(d) {
  fisher <- compute_one_way_test(formula = d$formula, df = d$df, var.equal = TRUE)
  welch <- compute_one_way_test(formula = d$formula, df = d$df, var.equal = FALSE)
  check <- check_one_way(formula = d$formula, df = d$df)
  data.frame(formula = deparse(d$formula),
             N = fisher$df_error + fisher$df_effect + 1,
             k = fisher$df_effect + 1,
             F_fisher = fisher$statistic,
             F_welch = welch$statistic,
             etasq = fisher$etasq,
             omegasq = fisher$omegasq,
             max_abs_difference = max(abs(check$difference)))
}))
```

## Fisher and Welch with unequal variances

Three groups have the **same** population mean, so every rejection is a Type I error. The smallest group has the largest standard deviation. At $\alpha=0.05$ a test should reject about 5% of the time:

```{r simulation_welch}
set.seed(1)
rejections <- t(replicate(2000, {
  sim <- data.frame(group = rep(c("a", "b", "c"), times = c(10, 20, 40)),
                    score = rnorm(70, mean = 0, sd = rep(c(4, 2, 1), times = c(10, 20, 40))))
  c(fisher = compute_one_way_test(formula = score ~ group, df = sim, var.equal = TRUE)$p < 0.05,
    welch = compute_one_way_test(formula = score ~ group, df = sim, var.equal = FALSE)$p < 0.05)
}))
colMeans(rejections)
```

Fisher's F rejects a true null hypothesis far more often than 5%, while Welch's F stays close to 5%.

## p values shrink, effect sizes do not

Three groups are drawn from normal distributions whose means differ by 0.3 standard deviations, with 10 to 1,000 observations per group. The population effect is the same in every row. As the sample grows, $F$ grows and the p value drops towards 0, while $\eta^2$, $\omega^2$ and $f$ settle around the same values:

```{r simulation_sample_size}
set.seed(1)
do.call(rbind, lapply(c(10, 30, 100, 300, 1000), function(n_per_group) {
  sim <- data.frame(group = rep(c("a", "b", "c"), each = n_per_group),
                    score = rnorm(3 * n_per_group, mean = rep(c(0, 0.3, 0.6), each = n_per_group)))
  cbind(n_per_group, compute_one_way_test(formula = score ~ group, df = sim)[, c("statistic", "p", "etasq", "omegasq", "cohens.f")])
}))
```

The population values are $f=\sqrt{\frac{1}{3}\sum(\mu_j-\bar{\mu})^2}=\sqrt{0.06}\approx0.245$ and $\eta^2=f^2/(1+f^2)\approx0.057$. A p value from one study says little about the size of the effect. The effect size does.

## Bias of eta-squared

When there is no effect at all, the average $\eta^2$ is close to $(k-1)/(N-1)$, while the average $\omega^2$ is close to 0. With $k=3$ groups of 10 the bias is $2/29\approx0.069$, which is already above the "medium" benchmark:

```{r simulation_bias}
set.seed(1)
null_effects <- t(replicate(2000, {
  sim <- data.frame(group = rep(c("a", "b", "c"), each = 10), score = rnorm(30))
  unlist(compute_one_way_test(formula = score ~ group, df = sim)[, c("etasq", "omegasq")])
}))
data.frame(mean_etasq = mean(null_effects[, "etasq"]),
           expected_etasq = (3 - 1) / (30 - 1),
           mean_omegasq = mean(null_effects[, "omegasq"]),
           proportion_negative_omegasq = mean(null_effects[, "omegasq"] < 0))
```

With small samples, report $\omega^2$, or remember that $\eta^2$ is inflated by roughly $(k-1)/(N-1)$.

## Missing values

`compute_one_way_test` removes rows with a missing value in the outcome or the grouping variable before the test, as `stats::oneway.test` does. Missing values in other columns of `df` are ignored. Two blood pressure values are set to missing here, and both functions use the remaining 118 rows:

```{r missing_values}
df_missing <- df_blood_pressure
df_missing$bp_before[c(1, 50)] <- NA
df_missing$bp_after[2] <- NA
compute_one_way_test(formula = form, df = df_missing, var.equal = TRUE)
stats::oneway.test(formula = form, data = df_missing, var.equal = TRUE)
check_one_way(formula = form, df = df_missing)
```

## References

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Delacre, M., Leys, C., Mora, Y. L., & Lakens, D. (2019). Taking parametric assumptions seriously: Arguments for the use of Welch's F-test instead of the classical F-test in one-way ANOVA. *International Review of Social Psychology, 32*(1), 13. https://doi.org/10.5334/irsp.198

Hays, W. L. (1963). *Statistics for psychologists*. Holt, Rinehart and Winston.

Hoenig, J. M., & Heisey, D. M. (2001). The abuse of power: The pervasive fallacy of power calculations for data analysis. *The American Statistician, 55*(1), 19–24. https://doi.org/10.1198/000313001300339897

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate cumulative science: A practical primer for t-tests and ANOVAs. *Frontiers in Psychology, 4*, 863. https://doi.org/10.3389/fpsyg.2013.00863

Olejnik, S., & Algina, J. (2003). Generalized eta and omega squared statistics: Measures of effect size for some common research designs. *Psychological Methods, 8*(4), 434–447. https://doi.org/10.1037/1082-989X.8.4.434

R Core Team. (2026). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria. https://doi.org/10.32614/R.manuals

Sawilowsky, S. S. (2009). New effect size rules of thumb. *Journal of Modern Applied Statistical Methods, 8*(2), 597–599. https://doi.org/10.22237/jmasm/1257035100

Welch, B. L. (1951). On the comparison of several mean values: An alternative approach. *Biometrika, 38*(3/4), 330–336. https://doi.org/10.2307/2332579

Wilkinson, L., & Task Force on Statistical Inference. (1999). Statistical methods in psychology journals: Guidelines and explanations. *American Psychologist, 54*(8), 594–604. https://doi.org/10.1037/0003-066X.54.8.594
