---
title: "GLM Reliability"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM Reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette documents the reliability and validity functions from `working_functions/GLM_RELIABILITY.R`: Cronbach's alpha and its item-level diagnostics, a Multitrait-Multimethod (MTMM) matrix for convergent/discriminant validity, and variance-components-based reliability (Shrout-Fleiss) for person × item × time measurement designs.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### Cronbach's alpha

Cronbach's alpha estimates **internal consistency** — how much the items of a scale covary, as a proxy for how much they all measure the same underlying construct:

$$\alpha = \frac{k}{k-1}\left(1-\frac{\sum \text{diag}(\Sigma)}{\sum \Sigma}\right)$$

where $k$ is the number of items and $\Sigma$ is their covariance matrix. Conventionally, $\alpha \geq 0.70$ is considered acceptable for research purposes and $\geq 0.80$ for applied/clinical decisions — though alpha can be inflated merely by adding more items, regardless of whether they add real construct coverage (Cortina, 1993).

**Item-total correlation** (`r`) is each item's correlation with the total scale score (which includes that item); **corrected item-total correlation** / **r-drop** excludes the item from the total before correlating, avoiding the artificial inflation of correlating an item with a sum that already contains it. Items with r-drop below ~0.30 are usually flagged as poor scale members. **Alpha-if-item-removed** shows what alpha would be without each item — an item whose removal *increases* alpha is actively hurting the scale's internal consistency.

### Multitrait-Multimethod (MTMM) matrix

Campbell & Fiske (1959)'s MTMM design measures several **traits** using several **methods**, then examines four types of correlation among the resulting scale scores:

| Type | Same trait? | Same method? | Interpretation |
|---|---|---|---|
| Monotrait-monomethod | ✓ | ✓ | Reliability (this is Cronbach's alpha) |
| Monotrait-heteromethod | ✓ | ✗ | **Convergent validity** — different methods measuring the same trait should agree |
| Heterotrait-monomethod | ✗ | ✓ | Method variance — different traits measured the same way should *not* agree strongly |
| Heterotrait-heteromethod | ✗ | ✗ | **Discriminant validity** — the weakest correlations of all, since neither trait nor method is shared |

Good construct validity shows high monotrait-heteromethod correlations (convergence) and low heterotrait correlations of both kinds (discrimination).

### Generalizability theory and Shrout-Fleiss reliability

When a measurement design has multiple facets of variation (e.g. multiple raters/items *and* multiple time points), classical reliability doesn't distinguish between different *sources* of unreliability. Extracting **variance components** from a mixed model (person, item, time, and their interactions) lets you compute several distinct reliability coefficients (Shrout & Fleiss, 1979) tailored to how scores will actually be used — a single measurement vs. an average of several, at fixed vs. randomly-sampled time points.

---

## Cronbach's Alpha

```{r alpha_setup}
set.seed(12345)
cm <- matrix(.5, ncol = 6, nrow = 6)
diag(cm) <- 1
df_items <- round(generate_correlation_matrix(cm, nrows = 1000), 0) + 5
```

```{r compute_raw_alpha}
psych::alpha(df_items)$total$raw_alpha
compute_raw_alpha(df = df_items)
```

`compute_raw_alpha` matches `psych::alpha`'s raw alpha exactly, computed directly from the covariance-matrix formula above rather than via `psych`'s more elaborate internals.

### Item diagnostics

```{r compute_alpha_diagnostics}
compute_alpha_diagnostics(df = df_items)
```

Every item here shares the same 0.5 pairwise correlation by construction, so alpha-if-removed and the item-total correlations are near-identical across items — in a real scale, an item with a notably lower `item.total.correlation.r.drop` than the rest would be the first candidate for removal.

### Scale score mean and SD

```{r compute_mean_sd_alpha}
compute_mean_sd_alpha(df_items)
compute_mean_sd_alpha(df_items, divisor = 100)
```

With `divisor = NULL` (default) the scale score is the row mean; with a `divisor` supplied, it's the row sum divided by that value instead — e.g. dividing by the maximum possible sum score to express results as a proportion.

## Full Alpha Report

`report_alpha` runs `psych::alpha` across one or more named scales at once (with optional item reversal and bootstrap CIs), returning scale-level, item-level, and alpha-if-dropped tables together.

```{r report_alpha,results='hide'}
key <- list(f1 = c("X1", "X2", "X3"), f2 = c("X4", "X5", "X6"))
result <- report_alpha(df = df_items, key = key, check.keys = FALSE, n.iter = 2)
```

```{r report_alpha_output}
result$result_total[, c("dimension", "items", "raw_alpha", "alpha_criterion")]
result$result_item_statistics[, c("dimension", "question", "raw_alpha")]
```

The `alpha_criterion` column classifies each scale from "Unacceptable" (<0.60) through "Excellent" (>0.90), giving an immediate at-a-glance verdict alongside the numeric alpha.

## MTMM Matrix

`plot_mtmm` requires long-format data (one row per subject per method) with the same trait items repeated under each method.

```{r mtmm_setup,results='hide'}
set.seed(12345)
population_model <- "t1=~x1+.9*x2+.9*x3
                     t2=~x4+.9*x5+.9*x6
                     t3=~x7+.9*x8+.9*x9"
model_data <- lavaan::simulateData(population_model, sample.nobs = 1000)
model_data <- model_data[sample(1:1000, 1000, TRUE), ]
model_data <- rbind(model_data, model_data, model_data)
model_data$method <- c(rep("m1", 1000), rep("m2", 1000), rep("m3", 1000))
model_data$id <- rep(1:1000, 3)
key_mtmm <- list(t1 = paste0("x", 1:3), t2 = paste0("x", 4:6), t3 = paste0("x", 7:9))
```

```{r mtmm_plot,fig.width=9,fig.height=8}
plot_mtmm(df = model_data, key = key_mtmm, method = "method", subject = "id")
```

Here, the same three latent traits were simply duplicated across three "methods" (`m1`–`m3` are actually identical copies of the same data), so this is a best-case, idealized MTMM: monotrait-monomethod cells (diagonal, reliability) and monotrait-heteromethod cells (convergent validity) should both come out high, since there's no real method variance in this simulation.

`key_to_cfa_model` converts the same `key` list format into `lavaan` CFA syntax, for fitting a formal confirmatory model on the same trait structure:

```{r key_to_cfa}
cat(key_to_cfa_model(key_mtmm))
```

## Variance Components and Shrout-Fleiss Reliability

`extract_components` pulls variance components out of a `mixlm::lm` mixed model (person, item, time, and their interactions specified with `r(...)`), reporting each component's percentage of total variance.

```{r variance_components,fig.width=8,fig.height=5}
design <- expand.grid(time = 1:3, item = 1:3, person = 1:10)
design <- change_data_type(design, type = "factor")
design$response <- rowSums(change_data_type(design[, 1:2], type = "numeric")) + rnorm(90, 0, 0.1)
model_vc <- mixlm::lm(response ~ r(time) * r(person) + r(item) * r(person), data = design)
result_vc <- extract_components(model_vc)
result_vc$components
result_vc$plot
```

`compute_shrout` then converts these variance components into the five Shrout-Fleiss reliability coefficients:

```{r shrout}
vc <- result_vc$components
compute_shrout(
  sperson = vc[vc$component == "person", "VC"],
  spersonitem = vc[vc$component == "item:person", "VC"],
  stime = vc[vc$component == "time", "VC"],
  spersontime = vc[vc$component == "time:person", "VC"],
  serror = vc[vc$component == "Residuals", "VC"],
  m = 3, k = 3
)
```

- **r1f/r1r**: reliability of a *single* measurement, at fixed vs. random time points.
- **rkf/rkr**: reliability of scores *averaged* over `m` items and `k` time points — always at least as high as the single-measurement coefficients, since averaging reduces error.
- **rc**: reliability of *change* across time — a different question from the others (how confidently can you say a person's score has changed?), directly tied to the person×time interaction variance.

---

## Conclusion

`compute_raw_alpha`/`compute_alpha_diagnostics`/`report_alpha` cover classical internal-consistency reliability at both the scale and item level; `plot_mtmm`/`key_to_cfa_model` extend reliability into a full construct-validity design distinguishing convergent from discriminant validity; and `extract_components`/`compute_shrout` bring in generalizability theory for measurement designs with multiple facets of variation (raters, items, and time), producing reliability coefficients tailored to exactly how the scores will be used.

---

*Rendered with R `r getRversion()`*
