---
title: "GLM FRIEDMAN"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM FRIEDMAN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))
load(paste0(directory, "data/df_co2.rda"))
load(paste0(directory, "data/df_blood_pressure.rda"))
```

## Description

This shows the output of `compute_friedman_test` from the package rwf, which runs the Friedman rank-sum test (Friedman, 1937), the non-parametric alternative to the one-way repeated measures ANOVA. The function returns the test statistic $Q$ with its degrees of freedom and p value, Kendall's coefficient of concordance $W$ (Kendall & Babington Smith, 1939) as the effect size and, optionally, a bootstrap confidence interval for $W$. Each section gives the formulas the function uses, computes them by hand and checks the results against `stats`, `coin`, `rstatix` and `effectsize`.

Installation instructions for rwf can be found [here](https://github.com/sedzinfo/rwf)

The code can be found [here](https://github.com/sedzinfo/rwf/blob/main/rwf/R/GLM_ANOVA.R)

## Notation

There are $n$ blocks (usually subjects) and $k$ groups (usually conditions or time points). Each block has exactly one observation in each group: $y_{ij}$ is the observation of block $i$ in group $j$. The observations are ranked **within each block** from 1 to $k$, and tied values get the mean of the ranks they span (mid-ranks). $r_{ij}$ is the rank of $y_{ij}$ within block $i$, $R_j=\sum_{i=1}^{n}r_{ij}$ is the rank sum of group $j$, and $n(k+1)/2$ is the rank sum every group would have if the groups did not differ. $t$ are the sizes of the groups of tied values within blocks.

## The Friedman test

The Friedman test asks whether the blocks rank the groups in a consistent order. Each subject is compared only with itself, so differences between subjects (one plant absorbs more CO₂ than another overall) do not matter, only the order of the conditions within each subject. Its null hypothesis is that, within every block, all orderings of the groups are equally likely.

### The Q statistic

Without ties the statistic is

$$Q_0=\frac{12}{nk(k+1)}\sum_{j=1}^{k}R_j^2-3n(k+1)$$

The general form, which also corrects for ties within blocks, is

$$Q=\frac{12\sum_{j=1}^{k}\left(R_j-\frac{n(k+1)}{2}\right)^2}{nk(k+1)-\frac{\sum\left(t^3-t\right)}{k-1}}$$

Values that occur once in a block have $t=1$ and add nothing to the sum, so $Q=Q_0$ when there are no ties.

### Degrees of freedom and p value

Under the null hypothesis $Q$ approximately follows a chi-squared distribution with $k-1$ degrees of freedom:

$$df=k-1,\qquad p=P\left(\chi^2_{k-1}\ge Q\right)$$

## When to use it and assumptions

### When to use it

Use the Friedman test to compare three or more related measurements, where the same subjects are measured under every condition or the observations come in matched sets (blocks), and the repeated measures ANOVA is not a good fit:

- **The outcome is ordinal**, such as Likert ratings of several products by the same people, or rankings by judges.
- **The outcome is continuous but the residuals are clearly not normal** and there are few subjects.
- **There are outliers** that would dominate the condition means. Ranks within blocks limit the influence of any single value.
- **Sphericity is badly violated** and you would rather not rely on corrections.

Do not use it when:

- **The residuals are normal.** The repeated measures ANOVA is then more powerful, and the gap is larger than for the Kruskal-Wallis test, because the Friedman test throws away the differences between blocks and the sizes of the differences within them. Its asymptotic relative efficiency under normality is $\frac{3}{\pi}\cdot\frac{k}{k+1}$, which is 0.72 for $k=3$ and 0.84 for $k=7$ (Hollander, Wolfe & Chicken, 2014).
- **The groups are independent.** Use the Kruskal-Wallis test, `compute_kruskal_wallis_test`.
- **There are only two conditions.** The Friedman test with $k=2$ is the sign test, which uses only the direction of each difference. The Wilcoxon signed-rank test also uses the sizes of the differences and is usually more powerful.
- **The design has more than one within-subject factor, or between-subject factors.** The test handles a single factor in a complete block design.

With $k=2$ the statistic reduces to $Q=(n_+-n_-)^2/(n_++n_-)$, where $n_+$ and $n_-$ count the blocks with a positive and a negative difference, and tied blocks drop out. This is the squared z of the sign test:

```{r two_conditions}
bp_long <- data.frame(patient = rep(df_blood_pressure$patient, 2),
                      time = rep(c("before", "after"), each = nrow(df_blood_pressure)),
                      bp = c(df_blood_pressure$bp_before, df_blood_pressure$bp_after))
difference <- df_blood_pressure$bp_before - df_blood_pressure$bp_after
data.frame(Q_rwf = compute_friedman_test(formula = bp ~ time | patient, df = bp_long)$Q,
           Q_sign_test = (sum(difference > 0) - sum(difference < 0))^2 / sum(difference != 0))
```

### Assumptions

1. **Independent blocks.** The subjects (blocks) are independent of each other. Observations within a block may be related, which is the point of the design.
2. **One observation per block and group.** The design is an unreplicated complete block design. Blocks with a missing value are dropped (see [Missing values](#missing-values)).
3. **An outcome that is at least ordinal within blocks.** The values within a block must be orderable. Values do not have to be comparable across blocks.
4. **No block by group interaction.** The test detects a consistent shift across conditions. If some subjects go up and others go down, the rank sums cancel out and the test has little power.
5. **Enough blocks for the chi-squared approximation.** With few blocks use an exact or permutation p value, such as `coin::friedman_test(distribution = coin::approximate(nresample = 10000))`.

Normality and sphericity are **not** assumed.

### After a significant result

A significant $Q$ means that at least one condition differs from the others, not which ones. Follow it with pairwise comparisons, such as Wilcoxon signed-rank tests between every pair of conditions with an adjustment for multiple comparisons. `stats::pairwise.wilcox.test` with `paired = TRUE` pairs observations by position, so the data must be sorted by block within each group, as `df_co2` is:

```{r post_hoc}
stats::pairwise.wilcox.test(df_co2$uptake, df_co2$conc, paired = TRUE, p.adjust.method = "holm")
```

## Why effect sizes

A p value answers one question: if the conditions did not differ, how surprising would data like these be? It does not say how large the difference is. Because $Q$ grows with the number of blocks, a trivial difference becomes "significant" with many subjects, and a large difference can be "non-significant" with few.

An effect size answers a different question: how much do the conditions differ? Effect sizes are useful because they:

- **measure magnitude.** They separate practical importance from statistical significance.
- **do not grow with $n$.** The same population effect gives roughly the same effect size with 10 subjects or 1,000.
- **can be compared across studies.** This is what meta-analysis combines.
- **drive power analysis.** The sample size a new study needs depends on the effect size it expects.
- **are expected in reports.** The APA Task Force on Statistical Inference recommends reporting effect sizes for primary outcomes (Wilkinson & Task Force on Statistical Inference, 1999; see also Lakens, 2013).

The [simulation below](#p-values-shrink-effect-sizes-do-not) shows this with data.

## Kendall's W

### Formula

Kendall's coefficient of concordance measures how strongly the blocks agree on the order of the groups:

$$W=\frac{12\sum_{j=1}^{k}\left(R_j-\frac{n(k+1)}{2}\right)^2}{n^2\left(k^3-k\right)-n\sum\left(t^3-t\right)}=\frac{Q}{n(k-1)}$$

$W$ lies in $[0,1]$. $W=0$ means the rank sums are all equal, so the blocks show no common order. $W=1$ means every block ranks the groups in exactly the same order.

### W and the Spearman correlation

Without ties, $W$ is a rescaled mean of the Spearman correlations $r_s$ between all pairs of blocks:

$$\bar{r}_s=\frac{nW-1}{n-1}$$

`df_co2` has one plant with tied values, so the two are close but not identical here:

```{r spearman}
uptake <- tapply(df_co2$uptake, list(df_co2$Plant, df_co2$conc), identity)
spearman <- cor(t(uptake), method = "spearman")
W <- compute_friedman_test(formula = uptake ~ conc | Plant, df = df_co2)$kendall_w
n <- nrow(uptake)
data.frame(mean_spearman = mean(spearman[upper.tri(spearman)]),
           from_W = (n * W - 1) / (n - 1))
```

The relation also shows the bias of $W$. Under the null hypothesis $E[Q]=k-1$, so $E[W]=1/n$ rather than 0, while $\bar{r}_s$ is centred at 0 ([simulation below](#bias-of-kendalls-w)).

### Interpretation

Commonly used benchmarks for $W$ are the Cohen (1988) benchmarks for correlations, which `rstatix::friedman_effsize` also uses:

| Magnitude | $W$ |
|---|---|
| tiny | < 0.1 |
| small | 0.1 to < 0.3 |
| medium | 0.3 to < 0.5 |
| large | ≥ 0.5 |

`effectsize::interpret_kendalls_w` uses agreement labels instead (Landis & Koch, 1977): slight below 0.2, fair from 0.2, moderate from 0.4, substantial from 0.6 and almost perfect from 0.8. These are rough guides. What counts as a meaningful effect depends on the field and the outcome.

## Data

`df_co2` records the CO₂ uptake of 12 grass plants, each measured at 7 ambient CO₂ concentrations. The plants are the blocks and the concentrations are the groups. The formula `uptake ~ conc | Plant` asks whether uptake changes with concentration, comparing each plant only with itself.

```{r data}
head(df_co2)
form <- formula(uptake ~ conc | Plant)
table(df_co2$Plant, df_co2$conc)
```

Each line is one plant. Uptake rises with concentration in almost every plant, even though the plants differ a lot in their overall level:

```{r data_plot,fig.width=7,fig.height=5}
interaction.plot(df_co2$conc, df_co2$Plant, df_co2$uptake, legend = FALSE,
                 xlab = "CO2 concentration", ylab = "uptake", col = "grey40", lty = 1)
```

## compute_friedman_test

```{r compute_friedman_test}
result <- compute_friedman_test(formula = form, df = df_co2)
result
```

The output columns are:

| Column | Formula |
|---|---|
| `formula` | the model formula |
| `method` | the test name |
| `kendall_w` | $W=Q/(n(k-1))$ |
| `Q` | tie-corrected Friedman statistic |
| `df` | $k-1$ |
| `n` | number of complete blocks used |
| `p` | $P\left(\chi^2_{k-1}\ge Q\right)$ |

## Step by step

Every quantity from the formulas above, computed by hand. First the blocks by groups matrix and the ranks within each block:

```{r step_ranks}
y <- tapply(df_co2$uptake, list(df_co2$Plant, df_co2$conc), identity)
n <- nrow(y)
k <- ncol(y)
r <- t(apply(y, 1, rank))
r
R_j <- colSums(r)
R_j
```

$Q$ without ties, the tie correction and the corrected $Q$:

```{r step_Q}
Q_0 <- 12 / (n * k * (k + 1)) * sum(R_j^2) - 3 * n * (k + 1)
ties <- sum(apply(y, 1, function(u) {
  t <- table(u)
  sum(t^3 - t)
}))
Q <- 12 * sum((R_j - n * (k + 1) / 2)^2) / (n * k * (k + 1) - ties / (k - 1))
data.frame(Q_0, ties, Q)
```

Degrees of freedom, p value and Kendall's $W$, from $Q$ and from its own formula:

```{r step_W}
by_hand <- data.frame(kendall_w = Q / (n * (k - 1)),
                      Q = Q,
                      df = k - 1,
                      n = n,
                      p = pchisq(Q, k - 1, lower.tail = FALSE))
by_hand
12 * sum((R_j - n * (k + 1) / 2)^2) / (n^2 * (k^3 - k) - n * ties)
all.equal(by_hand, result[, c("kendall_w", "Q", "df", "n", "p")], check.attributes = FALSE)
```

## Checks against other packages

### stats, coin and rstatix

`stats::friedman.test` (with a formula or with the blocks by groups matrix), `coin::friedman_test` and `rstatix::friedman_test` give the same $Q$, degrees of freedom and p value. `coin` needs the groups and blocks as factors:

```{r check_stats}
stats::friedman.test(formula = form, data = df_co2)
stats::friedman.test(y)
coin::friedman_test(uptake ~ factor(conc) | factor(Plant), data = df_co2)
rstatix::friedman_test(df_co2, form)
```

### rstatix and effectsize

`rstatix::friedman_effsize` and `effectsize::kendalls_w` compute $W$. Their confidence intervals are bootstrapped, so they change slightly from run to run. The [confidence intervals](#confidence-intervals) section compares them with the interval from rwf.

```{r check_effect_sizes}
rstatix::friedman_effsize(df_co2, form, ci = TRUE, conf.level = 0.95, ci.type = "perc", nboot = 100)
effectsize::kendalls_w(form, data = df_co2)
```

### A test function

`check_friedman` puts each rwf value next to the value from the other packages and returns the difference. The other packages need complete blocks, so incomplete blocks are dropped first, as rwf does.

```{r check_function}
check_friedman <- function(formula, df) {
  vars <- all.vars(formula)
  df <- data.frame(y = df[[vars[1]]], g = factor(df[[vars[2]]]), b = factor(df[[vars[3]]]))
  incomplete <- unique(df$b[is.na(df$y)])
  df <- droplevels(df[!df$b %in% incomplete, ])
  rwf <- compute_friedman_test(formula = y ~ g | b, df = df)
  fr <- stats::friedman.test(y ~ g | b, data = df)
  comparison <- data.frame(
    statistic = c("Q", "Q", "Q", "df", "p", "kendall_w", "kendall_w"),
    package = c("stats::friedman.test", "coin::friedman_test", "rstatix::friedman_test", "stats::friedman.test",
                "stats::friedman.test", "rstatix::friedman_effsize", "effectsize::kendalls_w"),
    rwf = c(rwf$Q, rwf$Q, rwf$Q, rwf$df, rwf$p, rwf$kendall_w, rwf$kendall_w),
    other = c(unname(fr$statistic),
              unname(coin::statistic(coin::friedman_test(y ~ g | b, data = df))),
              rstatix::friedman_test(df, y ~ g | b)$statistic,
              unname(fr$parameter),
              fr$p.value,
              rstatix::friedman_effsize(df, y ~ g | b)$effsize,
              suppressWarnings(effectsize::kendalls_w(y ~ g | b, data = df, ci = NULL)$Kendalls_W)))
  comparison$difference <- comparison$rwf - comparison$other
  comparison
}
check_friedman(formula = form, df = df_co2)
```

The same check on seven datasets with different numbers of blocks and groups and with and without ties. `RoundingTimes` is the example from `?friedman.test`: 22 baseball players, each timed rounding first base with three methods. `max_abs_difference` is the largest absolute difference across all the comparisons for that dataset:

```{r check_datasets}
RoundingTimes <- matrix(c(5.40, 5.50, 5.55, 5.85, 5.70, 5.75, 5.20, 5.60, 5.50, 5.55, 5.50, 5.40,
                          5.90, 5.85, 5.70, 5.45, 5.55, 5.60, 5.40, 5.40, 5.35, 5.45, 5.50, 5.35,
                          5.25, 5.15, 5.00, 5.85, 5.80, 5.70, 5.25, 5.20, 5.10, 5.65, 5.55, 5.45,
                          5.60, 5.35, 5.45, 5.05, 5.00, 4.95, 5.50, 5.50, 5.40, 5.45, 5.55, 5.50,
                          5.55, 5.55, 5.35, 5.45, 5.50, 5.55, 5.50, 5.45, 5.25, 5.65, 5.60, 5.40,
                          5.70, 5.65, 5.55, 6.30, 6.30, 6.25),
                        nrow = 22, byrow = TRUE)
df_rounding <- data.frame(time = c(RoundingTimes),
                          method = rep(c("Round Out", "Narrow Angle", "Wide Angle"), each = 22),
                          player = rep(1:22, times = 3))
datasets <- list(
  list(formula = uptake ~ conc | Plant, df = df_co2),
  list(formula = time ~ method | player, df = df_rounding),
  list(formula = bp ~ time | patient, df = bp_long),
  list(formula = decrease ~ treatment | rowpos, df = OrchardSprays),
  list(formula = circumference ~ age | Tree, df = as.data.frame(Orange)),
  list(formula = conc ~ time | Subject, df = as.data.frame(Indometh)),
  list(formula = height ~ age | Seed, df = as.data.frame(Loblolly))
)
do.call(rbind, lapply(datasets, function(d) {
  rwf <- compute_friedman_test(formula = d$formula, df = d$df)
  check <- check_friedman(formula = d$formula, df = d$df)
  data.frame(formula = deparse(d$formula),
             n = rwf$n,
             k = rwf$df + 1,
             Q = rwf$Q,
             kendall_w = rwf$kendall_w,
             max_abs_difference = max(abs(check$difference)))
}))
```

## Confidence intervals

With `ci = TRUE` the function adds a percentile bootstrap confidence interval for $W$. It draws `nboot` samples of $n$ **blocks** with replacement, so each resampled subject keeps all of its repeated measurements, and computes $W$ in each, giving $W^{*}_1,\dots,W^{*}_B$. The interval is

$$\left[W^{*}_{(\alpha/2)},\ W^{*}_{(1-\alpha/2)}\right],\qquad \alpha=1-\texttt{conf.level}$$

where $W^{*}_{(q)}$ is the $q$ quantile of the bootstrap estimates.

```{r confidence_intervals}
set.seed(1)
compute_friedman_test(formula = form, df = df_co2, ci = TRUE, conf.level = 0.95, nboot = 5000)
```

`rstatix` and `effectsize` also resample blocks with the percentile method. Each bootstrap draws different resamples, so the limits agree only to about two decimals, and they get closer as `nboot` grows:

```{r confidence_intervals_check}
set.seed(1)
rwf <- compute_friedman_test(formula = form, df = df_co2, ci = TRUE, nboot = 5000)
set.seed(1)
rstatix_w <- rstatix::friedman_effsize(df_co2, form, ci = TRUE, ci.type = "perc", nboot = 5000)
set.seed(1)
effectsize_w <- suppressWarnings(effectsize::kendalls_w(form, data = df_co2, alternative = "two.sided", iterations = 5000))
data.frame(package = c("rwf", "rstatix::friedman_effsize", "effectsize::kendalls_w"),
           estimate = c(rwf$kendall_w, rstatix_w$effsize, effectsize_w$Kendalls_W),
           lower = c(rwf$kendall_w_lower, rstatix_w$conf.low, effectsize_w$CI_low),
           upper = c(rwf$kendall_w_upper, rstatix_w$conf.high, effectsize_w$CI_high))
```

`rstatix` rounds its limits to two decimals. `effectsize` reports one-sided intervals by default, so `alternative = "two.sided"` is set here.

## p values shrink, effect sizes do not

Each simulated subject has its own overall level (standard deviation 1) and is measured under three conditions whose means differ by 0.3 standard deviations, with 10 to 1,000 subjects. The population effect is the same in every row. As the number of subjects grows, $Q$ grows and the p value drops towards 0, while $W$ settles around the same value:

```{r simulation_sample_size}
set.seed(1)
simulate_blocks <- function(n_subjects, shift) {
  subject_level <- rnorm(n_subjects)
  data.frame(subject = rep(seq_len(n_subjects), times = 3),
             condition = rep(c("a", "b", "c"), each = n_subjects),
             score = rep(subject_level, times = 3) + rep(c(0, 1, 2) * shift, each = n_subjects) + rnorm(3 * n_subjects))
}
do.call(rbind, lapply(c(10, 30, 100, 300, 1000), function(n_subjects) {
  sim <- simulate_blocks(n_subjects, shift = 0.3)
  cbind(n_subjects, compute_friedman_test(formula = score ~ condition | subject, df = sim)[, c("Q", "p", "kendall_w")])
}))
```

A p value from one study says little about the size of the effect. The effect size does.

## Bias of Kendall's W

When there is no effect at all, the average $W$ is close to $1/n$, while the average $\bar{r}_s=(nW-1)/(n-1)$ is close to 0. With 10 subjects the bias is $1/10=0.1$, which already reaches the "small" benchmark:

```{r simulation_bias}
set.seed(1)
null_W <- replicate(2000, {
  sim <- simulate_blocks(10, shift = 0)
  compute_friedman_test(formula = score ~ condition | subject, df = sim)$kendall_w
})
data.frame(mean_kendall_w = mean(null_W),
           expected_kendall_w = 1 / 10,
           mean_spearman = mean((10 * null_W - 1) / (10 - 1)))
```

With few subjects, judge $W$ against $1/n$, or report $\bar{r}_s$ alongside it.

## Missing values

`compute_friedman_test` drops every block with a missing value and reports in `n` how many blocks remain. This is what the matrix method of `stats::friedman.test` does. The formula method of `stats::friedman.test` stops with an error instead. Here one measurement of the first plant is set to missing:

```{r missing_values}
df_missing <- df_co2
df_missing$uptake[1] <- NA
compute_friedman_test(formula = form, df = df_missing)
tryCatch(stats::friedman.test(formula = form, data = df_missing), error = conditionMessage)
stats::friedman.test(tapply(df_missing$uptake, list(df_missing$Plant, df_missing$conc), identity))
check_friedman(formula = form, df = df_missing)
```

A block with two observations for the same group is an error, because the design is no longer an unreplicated complete block design:

```{r replicated}
tryCatch(compute_friedman_test(formula = form, df = rbind(df_co2, df_co2[1, ])), error = conditionMessage)
```

## References

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Friedman, M. (1937). The use of ranks to avoid the assumption of normality implicit in the analysis of variance. *Journal of the American Statistical Association, 32*(200), 675–701. https://doi.org/10.1080/01621459.1937.10503522

Hollander, M., Wolfe, D. A., & Chicken, E. (2014). *Nonparametric statistical methods* (3rd ed.). Wiley.

Kendall, M. G., & Babington Smith, B. (1939). The problem of m rankings. *The Annals of Mathematical Statistics, 10*(3), 275–287. https://doi.org/10.1214/aoms/1177732186

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate cumulative science: A practical primer for t-tests and ANOVAs. *Frontiers in Psychology, 4*, 863. https://doi.org/10.3389/fpsyg.2013.00863

Landis, J. R., & Koch, G. G. (1977). The measurement of observer agreement for categorical data. *Biometrics, 33*(1), 159–174. https://doi.org/10.2307/2529310

Wilkinson, L., & Task Force on Statistical Inference. (1999). Statistical methods in psychology journals: Guidelines and explanations. *American Psychologist, 54*(8), 594–604. https://doi.org/10.1037/0003-066X.54.8.594
