---
title: "GLM IRT Unidimensional"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM IRT Unidimensional}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
library(ggplot2)
```

## Description

`working_functions/GLM_IRT_U.R` is a from-scratch, base-R implementation of unidimensional IRT's two core computations: the item response function (with optional guessing/inattentiveness parameters) and the item/test information used to derive standard errors — plus a hand-rolled Newton-Raphson routine for Maximum Likelihood ability estimation. Six functions in total: `compute_unidimensional_theta`, `compute_unidimensional_ability`, `compute_info_1pl`, `compute_info_2pl`, `compute_info_3pl`, and `compute_se_theta`.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### The item response function

For a single item with discrimination $a$, difficulty $b$, and guessing $g$, the probability of a correct response at ability $\theta$ is

$$P(\theta) = g + \frac{1-g}{1+e^{-Da(\theta-b)}}$$

where $D$ is a scaling constant (1, or 1.702 to make the logistic curve closely approximate the normal ogive). Setting $g=0$ collapses this to the 2PL model; additionally fixing $a$ across items gives the 1PL/Rasch model.

### Newton-Raphson Maximum Likelihood ability estimation

Given a fixed set of item parameters and a response pattern $u_1,\ldots,u_n$, the ML ability estimate is found by iterating

$$\theta_{t+1} = \theta_t - \frac{L'(\theta_t)}{L''(\theta_t)}$$

where $L'$ and $L''$ are the first and second derivatives of the log-likelihood, summed across items:

$$L'(\theta)=\sum_i \frac{Da_i(u_i-P_i)(P_i-g_i)}{P_i(1-g_i)} \qquad L''(\theta)=\sum_i \left(\frac{Da_i}{1-g_i}\right)^2\frac{(P_i-g_i)(1-P_i)(u_ig_i-P_i^2)}{P_i^2}$$

Iteration continues until the update is smaller than a tolerance (or a maximum iteration count is hit), and the final estimate is clamped to `lim_theta` to avoid runaway estimates for all-correct/all-incorrect patterns (a known weakness of ML estimation, discussed further in the [EAP vignette](EXAMPLE_EAP.rmd), which uses a Bayesian prior instead to avoid this failure mode entirely).

### Item and test information

Information quantifies precision at a given $\theta$; its formula depends on the model:

$$I_{1PL}(\theta)=P(1-P) \qquad I_{2PL}(\theta)=a^2P(1-P) \qquad I_{3PL}(\theta)=a^2\frac{1-P}{P}\left(\frac{P-g}{1-g}\right)^2$$

and the standard error of the ability estimate at that point is $SE(\theta)=1/\sqrt{I(\theta)}$ — more information means a smaller (better) standard error.

---

## The Item Response Function

```{r theta_curves,fig.width=8,fig.height=6}
x <- seq(-3, 3, by = 0.01)
df_curves <- data.frame(
  theta = x,
  a5_b0  = compute_unidimensional_theta(a = 5, b = 0, theta = x),
  a1_b0  = compute_unidimensional_theta(a = 1, b = 0, theta = x),
  a10_b0 = compute_unidimensional_theta(a = 10, b = 0, theta = x),
  a5_bm1 = compute_unidimensional_theta(a = 5, b = -1, theta = x),
  a5_bp1 = compute_unidimensional_theta(a = 5, b = 1, theta = x)
)
df_long <- reshape2::melt(df_curves, id.vars = "theta", variable.name = "curve", value.name = "P")
ggplot(df_long, aes(x = theta, y = P, color = curve)) +
  geom_line(linewidth = 1) +
  labs(title = "Effect of discrimination (a) and difficulty (b)", x = expression(theta), y = expression(P(theta))) +
  theme_bw(base_size = 12)
```

Higher `a` produces a steeper curve (sharper discrimination around `b`); `b` shifts the curve's midpoint left or right along the ability scale.

### Checking the guessing parameter

```{r guessing_check,fig.width=8,fig.height=6}
df_guess <- data.frame(
  theta = x,
  g0.0 = compute_unidimensional_theta(a = 10, b = 0, g = 0, theta = x),
  g0.1 = compute_unidimensional_theta(a = 10, b = 0, g = .1, theta = x),
  g0.5 = compute_unidimensional_theta(a = 10, b = 0, g = .5, theta = x)
)
df_guess_long <- reshape2::melt(df_guess, id.vars = "theta", variable.name = "g", value.name = "P")
ggplot(df_guess_long, aes(x = theta, y = P, color = g)) +
  geom_line(linewidth = 1) +
  labs(title = "Effect of guessing (g): raises the lower asymptote", x = expression(theta), y = expression(P(theta))) +
  theme_bw(base_size = 12)
```

Increasing `g` raises the curve's lower asymptote — even at very low ability, the probability of a correct response never drops below `g`, exactly as intended for a 3PL guessing parameter.

### The inattentiveness parameter `i` has no effect

The function signature also accepts an inattentiveness parameter `i` (documented as making the function compute a "4PL score" when `i != 1`) — but checking the source shows the function body never actually uses `i` in its formula. This is verifiable directly:

```{r i_parameter_check}
data.frame(
  theta = x[1:5],
  i_1.0 = compute_unidimensional_theta(a = 10, b = 0, g = 0, i = 1, theta = x[1:5]),
  i_0.9 = compute_unidimensional_theta(a = 10, b = 0, g = 0, i = .9, theta = x[1:5]),
  i_0.6 = compute_unidimensional_theta(a = 10, b = 0, g = 0, i = .6, theta = x[1:5])
)
```

All three columns are identical regardless of `i` — so despite the documentation, this function only ever computes a 3PL (or 2PL/1PL) curve, never a true 4PL curve with a lowered upper asymptote. If a genuine 4PL upper asymptote is needed, the numerator `renum <- 1-g` inside the function would need to become `renum <- i-g`.

## Newton-Raphson Ability Estimation

`compute_unidimensional_ability` is validated directly against the three worked examples documented in its own roxygen block, each of which records the exact expected output:

```{r ability_validation}
a <- c(0.39, 0.45, 0.52, 0.3, 0.35, 0.43, 0.42, 0.44, 0.34, 0.42)
b <- c(-1.96, -1.9, -1.38, -0.58, 0.48, -0.81, -0.35, 1.59, 1.33, 2.93)
u <- c(1, 1, 1, 1, 0, 0, 1, 0, 1, 0)
theta1 <- compute_unidimensional_ability(a = a, b = b, u = u, d = 1.7, g = NULL)
c(estimated = theta1, expected = 0.48402574251176, match = isTRUE(all.equal(theta1, 0.48402574251176, tolerance = 1e-6)))

a <- c(1.27, 0.9, 0.94, 0.95, 0.55, 0.6, 0.44, 0.4)
b <- c(-0.54, 0.18, 0.21, 1.26, 1.73, -0.87, 1.72, 2.67)
u <- c(1, 1, 1, 1, 0, 0, 0, 0)
theta2 <- compute_unidimensional_ability(a = a, b = b, u = u, d = 1.7, g = NULL)
c(estimated = theta2, expected = 1.04621621510192, match = isTRUE(all.equal(theta2, 1.04621621510192, tolerance = 1e-6)))

a <- c(0.41, 0.32, 0.33, 1.2, 0.63, 0.62, 0.7, 0.61, 0.38, 0.53, 0.6, 1.16)
b <- c(-1.4, -1.3, -1.17, 0.2, 0.71, 0.86, -0.12, 0.12, 2.06, 1.38, 1.18, -0.33)
u <- c(1, 0, 1, 1, 0, 0, 0, 1, 1, 0, 1, 0)
theta3 <- compute_unidimensional_ability(a = a, b = b, u = u, d = 1.7, g = NULL)
c(estimated = theta3, expected = 0.0860506282671103, match = isTRUE(all.equal(theta3, 0.0860506282671103, tolerance = 1e-6)))
```

All three match to at least 6 decimal places, confirming the Newton-Raphson implementation is numerically correct for these response patterns.

## Item and Test Information

```{r info_curves,fig.width=8,fig.height=6}
theta_grid <- seq(-6, 6, by = 0.01)
df_info <- data.frame(
  theta = theta_grid,
  `1PL` = compute_info_1pl(b = 1, theta = theta_grid),
  `2PL (a=1)` = compute_info_2pl(a = 1, b = -2, theta = theta_grid),
  `2PL (a=2)` = compute_info_2pl(a = 2, b = 0, theta = theta_grid),
  `2PL (a=3)` = compute_info_2pl(a = 3, b = 2, theta = theta_grid),
  `3PL (g=0.2)` = compute_info_3pl(a = 1.5, b = 1, g = .2, theta = theta_grid),
  check.names = FALSE
)
df_info_long <- reshape2::melt(df_info, id.vars = "theta", variable.name = "model", value.name = "information")
ggplot(df_info_long, aes(x = theta, y = information, color = model)) +
  geom_line(linewidth = 1) +
  labs(title = "Item information by model and parameters", x = expression(theta), y = "Information") +
  theme_bw(base_size = 12)
```

Every curve peaks near its own difficulty `b`. Higher discrimination (`a`) produces a taller, narrower peak — more precision at the target ability, less everywhere else. Adding a guessing parameter (3PL) *lowers* peak information relative to an otherwise-equivalent 2PL item, because a nonzero floor on the response probability makes wrong answers less diagnostic of low ability.

## Standard Error from Information

```{r se_curve,fig.width=8,fig.height=6}
info_2pl <- compute_info_2pl(a = 1.5, b = 0, theta = theta_grid)
df_se <- data.frame(theta = theta_grid, information = info_2pl, se = compute_se_theta(info_2pl))
df_se_long <- reshape2::melt(df_se, id.vars = "theta", variable.name = "quantity", value.name = "value")
ggplot(df_se_long, aes(x = theta, y = value, color = quantity)) +
  geom_line(linewidth = 1) +
  facet_wrap(~quantity, scales = "free_y", ncol = 1) +
  labs(title = "Information and its corresponding SE", x = expression(theta), y = NULL) +
  theme_bw(base_size = 12)
```

As information peaks near $\theta=b=0$, SE simultaneously reaches its minimum there — precision and information are two views of the same underlying quantity, related by $SE=1/\sqrt{I}$.

---

## Conclusion

This file implements the two pillars of unidimensional IRT scoring entirely in base R: the item response function (validated to behave correctly for discrimination, difficulty, and guessing, though the documented `i`/4PL inattentiveness behavior turned out not to be implemented) and Newton-Raphson ML ability estimation (validated exactly against its own documented worked examples), plus the 1PL/2PL/3PL information formulas that convert a fitted model into a standard error at any ability level.

---

*Rendered with R `r getRversion()`*
