---
title: "GLM Linear Regression"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM Linear Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
options(future.show.progress=FALSE)
options(width=1000)
knitr::opts_chunk$set(echo=TRUE)
hook_output<-function(x, options) {
  paste0('<pre class="r-output">',knitr::knit_print(x),'</pre>')
}
knitr::knit_hooks$set(output=hook_output)
```

```{r load,include=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette demonstrates linear regression functions from the `rwf` package, covering scatterplot visualisation and full regression reporting for both simple and multiple regression models.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Plot Scatterplot

`plot_scatterplot` produces scatterplots with a fitted regression line and confidence band. It supports pairwise plots across all columns of a dataframe, specific variable combinations, and symmetric vs. non-symmetric axis scaling.

The examples show:

- **Full `mtcars` matrix**: all pairwise scatterplots stored in `result` for later multiplot use.
- **Single pair, equal axes** (`coord_equal = TRUE`): both axes share the same scale — useful when variables are on comparable units.
- **Single pair, free axes** (`coord_equal = FALSE`): each axis scales independently — better for variables with very different ranges.
- **Custom combinations**: only selected x–y pairs are plotted using the `combinations` argument, avoiding a full matrix when you only care about specific relationships.
- **All orders** (`all_orders = TRUE`): plots both x~y and y~x for a given pair, showing how the regression line changes with axis assignment.
- **Random simulation**: x and y generated with near-perfect correlation (sd = 0.1 noise) — confirms the function handles near-linear data cleanly.
- **Negative correlation simulation**: a matrix with r = −0.999 is generated and plotted — a stress test for strong negative linear relationships.

```{r, plot_scatterplot,fig.width=8,fig.height=8,message=FALSE,warning=FALSE,results='hide'}
result<-plot_scatterplot(df=mtcars,title="",coord_equal=TRUE,base_size=10)
plot_scatterplot(df=mtcars[,1:2],base_size=10,coord_equal=TRUE,all_orders=FALSE)
plot_scatterplot(df=mtcars[,1:2],base_size=10,coord_equal=FALSE,all_orders=FALSE)
plot_scatterplot(df=mtcars,base_size=10,coord_equal=TRUE,all_orders=FALSE,
                 combinations=data.frame(x=c("mpg","mpg","mpg"),
                                         y=c("cyl","hp","mpg")))
plot_scatterplot(df=mtcars,base_size=10,coord_equal=TRUE,all_orders=TRUE,
                 combinations=data.frame(x=c("mpg"),y=c("cyl")))
x<-rnorm(1000)
y<-x+rnorm(x,sd=.1)
plot_scatterplot(df=data.frame(x,y),title="Random Simulation",coord_equal=TRUE)
df<-data.frame(matrix(-.999,ncol=2,nrow=2))
correlation_martix<-as.matrix(df)
diag(correlation_martix)<-1
df<-generate_correlation_matrix(correlation_martix,nrows=1000)
plot_scatterplot(df,title="Simulation of -.999 Correlation",coord_equal=TRUE,base_size=10)
```

## Plot Multiplot

`plot_multiplot` arranges multiple ggplot objects into a grid. Here it displays the first 12 pairwise scatterplots from the full `mtcars` matrix generated above, laid out in 4 columns. This is useful for quickly scanning all pairwise relationships in a dataset side by side.

```{r, plot_multiplot,fig.width=20,fig.height=20}
plot_multiplot(plotlist=result[1:12],cols=4)
```

## Report Regression

`report_regression` fits a linear model and returns a structured report with model fit statistics (R², adjusted R², F, p), a coefficient table with standardized and unstandardized betas, confidence intervals, and tolerance/VIF collinearity diagnostics.

Two models are demonstrated:

- **Simple regression** (`mpg ~ qsec`): one predictor, with `plot_diagnostics = TRUE` to produce residual diagnostic plots (residuals vs. fitted, Q-Q plot, scale-location, leverage). These help assess whether linearity, homoscedasticity, and normality assumptions hold.
- **Multiple regression with interactions** (`mpg ~ qsec * hp * wt * drat`): all main effects and their two-, three-, and four-way interactions. This tests how the function handles a highly parameterised model and whether collinearity among interaction terms is flagged in the VIF output.

```{r, report_regression,fig.width=8,fig.height=8}
form<-formula(mpg~qsec)
regressionmodel<-lm(form,data=mtcars)
multipleregressionmodel<-lm(mpg~qsec*hp*wt*drat,data=mtcars)
res<-report_regression(model=regressionmodel,plot_diagnostics=TRUE)
res<-report_regression(model=multipleregressionmodel)
res
```
