---
title: "GLM_MEANS"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM_MEANS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
options(width=1000)
hook_output <- function(x, options) {
  paste0('<pre class="r-output">', knitr::knit_print(x), '</pre>')
}
knitr::knit_hooks$set(output = hook_output)
knitr::opts_chunk$set(echo=TRUE)
options(future.show.progress = FALSE)
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette demonstrates mean comparison functions from the `rwf` package — `report_ttests` and `report_wtests`. Both functions run pairwise comparisons across all combinations of independent variable levels and return a structured table with test statistics, effect sizes, confidence intervals, and Bonferroni-adjusted significance flags.

- Use `report_ttests` when data meet the normality assumption (parametric).
- Use `report_wtests` when data are ordinal or normality cannot be assumed (non-parametric).

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Data

The `df_insurance` dataset is used throughout this vignette. It contains insurance data for 1338 individuals with the following variables:

Format: A data frame with 1338 rows and 7 variables:

- **age**: Age of the individual
- **sex**: Sex of the individual (male/female)
- **bmi**: Body Mass Index
- **children**: Number of children covered by the insurance
- **smoker**: Smoking status (yes/no)
- **region**: Region of residence (southwest, southeast, northwest, northeast)
- **charges**: Medical insurance charges billed

Source: researchpy repo

```{r, data}
head(df_insurance)
cdf(df_insurance)
```

## T-Test

`report_ttests` runs independent-samples t-tests for all pairwise combinations of levels of the specified independent variable(s) against a dependent variable. When multiple IVs are provided, it tests all level combinations across variables.

The calls below demonstrate the key options:

- **Single IV** (`iv = 2`, sex): one pairwise comparison (male vs. female).
- **Single IV** (`iv = 4`, children): pairwise comparisons across all child-count groups.
- **Multiple IVs** (`iv = c(2,4)`): all combinations of sex and children levels tested simultaneously.
- **`alternative`**: controls the direction of the hypothesis — `"two.sided"` (default), `"less"`, or `"greater"`.
- **`var.equal = TRUE`**: uses Student's t-test assuming equal variances; default (`FALSE`) uses Welch's t-test which does not assume equal variances and is generally safer.

```{r, report_ttests}
report_ttests(df=df_insurance,dv=7,iv=c(2))
report_ttests(df=df_insurance,dv=7,iv=c(4))
report_ttests(df=df_insurance,dv=7,iv=c(2,4))
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="two.sided")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="less")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="greater")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),var.equal=TRUE)
```

## Explanation of columns
| Column   | Description                                                                 |
|----------|-----------------------------------------------------------------------------|
|DV                  |dependent variable|
|IV                  |independent variable|
|level1              |level 1|
|level2              |level 2|
|n1                  |sample size for level 1|
|n2                  |sample size for level 2|
|t                   |t statistic|
|df                  |degrees of freedom for t statistic|
|p                   |p value|
|CI_l                |confidence interval lower bound|
|CI_u                |confidence interval upper bound|
|alternative         |alternative hypothesis|
|method              |name of method should be displayed here|
|mean1               |mean for level 1|
|mean2               |mean for level 2|
|sd1                 |standard deviation for level 1|
|sd2                 |standard deviation for level 2|
|sd_pooled           |pooled standard deviation combined standard deviation of both levels|
|d                   |measure of effect size cohen's d|
|d                   |Very small 0.01 Sawilowsky (2009)|
|d                   |Small 0.20 Cohen (1988)|
|d                   |Medium 0.50 Cohen (1988)|
|d                   |Large 0.80 Cohen (1988)|
|d                   |Very large 1.20 Sawilowsky (2009)|
|d                   |Huge 12.0 Sawilowsky (2009)|
|r                   |measure of effect size correlation effect size|
|k_squared bartlett  |bartlett test for homogeneity of variances|
|df bartlett         |bartlett test for homogeneity of variances|
|p bartlett          |bartlett test for homogeneity of variances significant values indicate heteroscedasticity|
|bonferroni_p        |bonferroni adjusted critical value for a=0.05|
|significant         |if TRUE result is significant after bonferroni adjustment|

## Wilcoxon Test

`report_wtests` runs the Wilcoxon rank-sum test (Mann-Whitney U), the non-parametric equivalent of the independent-samples t-test. It makes no assumption about the distribution of the data and compares group medians via rank ordering. Use it when the normality assumption is violated or the outcome is ordinal.

The calls mirror those in the T-Test section exactly, allowing direct comparison of parametric and non-parametric results on the same data. The key difference in the output is the **W statistic** (Wilcoxon rank-sum) in place of t, and the absence of pooled standard deviation as a meaningful quantity under rank-based testing.

```{r, report_wtests}
report_wtests(df=df_insurance,dv=7,iv=c(2))
report_wtests(df=df_insurance,dv=7,iv=c(4))
report_wtests(df=df_insurance,dv=7,iv=c(2,4))
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="two.sided")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="less")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="greater")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),var.equal=TRUE)
```

## Explanation of columns
| Column   | Description                                                                 |
|----------|-----------------------------------------------------------------------------|
|DV                  |dependent variable|
|IV                  |independent variable|
|level1              |level 1|
|level2              |level 2|
|n1                  |sample size for level 1|
|n2                  |sample size for level 2|
|W                   |wilcoxon statistic|
|df                  |degrees of freedom for t statistic|
|p                   |p value|
|CI_l                |confidence interval lower bound|
|CI_u                |confidence interval upper bound|
|alternative         |alternative hypothesis|
|method              |name of method should be displayed here|
|mean1               |mean for level 1|
|mean2               |mean for level 2|
|sd1                 |standard deviation for level 1|
|sd2                 |standard deviation for level 2|
|sd_pooled           |pooled standard deviation combined standard deviation of both levels|
|d                   |measure of effect size cohen's d|
|d                   |Very small 0.01 Sawilowsky (2009)|
|d                   |Small 0.20 Cohen (1988)|
|d                   |Medium 0.50 Cohen (1988)|
|d                   |Large 0.80 Cohen (1988)|
|d                   |Very large 1.20 Sawilowsky (2009)|
|d                   |Huge 12.0 Sawilowsky (2009)|
|r                   |measure of effect size correlation effect size|
|k_squared bartlett  |bartlett test for homogeneity of variances|
|df bartlett         |bartlett test for homogeneity of variances|
|p bartlett          |bartlett test for homogeneity of variances significant values indicate heteroscedasticity|
|bonferroni_p        |bonferroni adjusted critical value for a=0.05|
|significant         |if TRUE result is significant after bonferroni adjustment|
