---
title: "GLM_MEANS"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM_MEANS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))

```

## Description

This vignette demonstrates mean comparison functions from the `rwf` package — `report_ttests` and `report_wtests`. Both functions run pairwise comparisons across all combinations of independent variable levels and return a structured table with test statistics, effect sizes, confidence intervals, and Bonferroni-adjusted significance flags.

- Use `report_ttests` when data meet the normality assumption (parametric).
- Use `report_wtests` when data are ordinal or normality cannot be assumed (non-parametric).

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Data

The `df_insurance` dataset is used throughout this vignette. It contains insurance data for 1338 individuals with the following variables:

Format: A data frame with 1338 rows and 7 variables:

- **age**: Age of the individual
- **sex**: Sex of the individual (male/female)
- **bmi**: Body Mass Index
- **children**: Number of children covered by the insurance
- **smoker**: Smoking status (yes/no)
- **region**: Region of residence (southwest, southeast, northwest, northeast)
- **charges**: Medical insurance charges billed

Source: researchpy repo

```{r, data}
load(paste0(directory,"data/df_insurance.rda"))
head(df_insurance)
cdf(df_insurance)
```

## T-Test

`report_ttests` runs independent-samples t-tests for all pairwise combinations of levels of the specified independent variable(s) against a dependent variable. When multiple IVs are provided, it tests all level combinations across variables.

The calls below demonstrate the key options:

- **Single IV** (`iv = 2`, sex): one pairwise comparison (male vs. female).
- **Single IV** (`iv = 4`, children): pairwise comparisons across all child-count groups.
- **Multiple IVs** (`iv = c(2,4)`): all combinations of sex and children levels tested simultaneously.
- **`alternative`**: controls the direction of the hypothesis — `"two.sided"` (default), `"less"`, or `"greater"`.
- **`var.equal = TRUE`**: uses Student's t-test assuming equal variances; default (`FALSE`) uses Welch's t-test which does not assume equal variances and is generally safer.

```{r, report_ttests}
report_ttests(df=df_insurance,dv=7,iv=c(2))
report_ttests(df=df_insurance,dv=7,iv=c(4))
report_ttests(df=df_insurance,dv=7,iv=c(2,4))
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="two.sided")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="less")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),alternative="greater")
report_ttests(df=df_insurance,dv=7,iv=c(2,4),var.equal=TRUE)
```

## Explanation of columns
| Column   |           |Description                            |  Reference               |
|----------|-----------|---------------------------------------|--------------------------|
|DV                  ||dependent variable|
|IV                  ||independent variable|
|level1              ||level 1|
|level2              ||level 2|
|n1                  ||sample size for level 1|
|n2                  ||sample size for level 2|
|t                   ||t statistic|
|df                  ||degrees of freedom for t statistic|
|p                   ||p value|
|CI_l                ||confidence interval lower bound|
|CI_u                ||confidence interval upper bound|
|alternative         ||alternative hypothesis|
|method              ||name of method should be displayed here|
|mean1               ||mean for level 1|
|mean2               ||mean for level 2|
|sd1                 ||standard deviation for level 1|
|sd2                 ||standard deviation for level 2|
|sd_pooled           ||pooled standard deviation combined standard deviation of both levels|
|d                   ||measure of effect size cohen's d|
|d                   |Very small| 0.01| Sawilowsky (2009)|
|d                   |Small|      0.20| Cohen (1988)|
|d                   |Medium|     0.50| Cohen (1988)|
|d                   |Large|      0.80| Cohen (1988)|
|d                   |Very large| 1.20| Sawilowsky (2009)|
|d                   |Huge|       12.0| Sawilowsky (2009)|
|r                   ||measure of effect size correlation effect size|
|k_squared bartlett  ||bartlett test for homogeneity of variances|
|df bartlett         ||bartlett test for homogeneity of variances|
|p bartlett          ||bartlett test for homogeneity of variances significant values indicate heteroscedasticity|
|bonferroni_p        ||bonferroni adjusted critical value for a=0.05|
|significant         ||if TRUE result is significant after bonferroni adjustment|

## Wilcoxon Test

`report_wtests` runs the Wilcoxon rank-sum test (Mann-Whitney U), the non-parametric equivalent of the independent-samples t-test. It makes no assumption about the distribution of the data and compares group medians via rank ordering. Use it when the normality assumption is violated or the outcome is ordinal.

The calls mirror those in the T-Test section exactly, allowing direct comparison of parametric and non-parametric results on the same data. The key difference in the output is the **W statistic** (Wilcoxon rank-sum) in place of t, and the absence of pooled standard deviation as a meaningful quantity under rank-based testing.

```{r, report_wtests}
report_wtests(df=df_insurance,dv=7,iv=c(2))
report_wtests(df=df_insurance,dv=7,iv=c(4))
report_wtests(df=df_insurance,dv=7,iv=c(2,4))
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="two.sided")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="less")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),alternative="greater")
report_wtests(df=df_insurance,dv=7,iv=c(2,4),var.equal=TRUE)
```

## Explanation of columns
| Column   |           |Description                            |  Reference               |
|----------|-----------|---------------------------------------|--------------------------|
|DV                  ||dependent variable|
|IV                  ||independent variable|
|level1              ||level 1|
|level2              ||level 2|
|n1                  ||sample size for level 1|
|n2                  ||sample size for level 2|
|W                   ||wilcoxon statistic|
|df                  ||degrees of freedom for t statistic|
|p                   ||p value|
|CI_l                ||confidence interval lower bound|
|CI_u                ||confidence interval upper bound|
|alternative         ||alternative hypothesis|
|method              ||name of method should be displayed here|
|mean1               ||mean for level 1|
|mean2               ||mean for level 2|
|sd1                 ||standard deviation for level 1|
|sd2                 ||standard deviation for level 2|
|sd_pooled           ||pooled standard deviation combined standard deviation of both levels|
|d                   ||measure of effect size cohen's d|
|d                   |Very small| 0.01 Sawilowsky (2009)|
|d                   |Small|      0.20| Cohen (1988)|
|d                   |Medium|     0.50| Cohen (1988)|
|d                   |Large|      0.80| Cohen (1988)|
|d                   |Very large| 1.20| Sawilowsky (2009)|
|d                   |Huge|       12.0| Sawilowsky (2009)|
|r                   ||measure of effect size correlation effect size|
|k_squared bartlett  ||bartlett test for homogeneity of variances|
|df bartlett         ||bartlett test for homogeneity of variances|
|p bartlett          ||bartlett test for homogeneity of variances significant values indicate heteroscedasticity|
|bonferroni_p        ||bonferroni adjusted critical value for a=0.05|
|significant         ||if TRUE result is significant after bonferroni adjustment|


## Formulas

Group 1 = `level1`, group 2 = `level2`; $x_{1i}$ ($i=1,\dots,n_1$) and $x_{2j}$ ($j=1,\dots,n_2$) are the DV values of the complete cases in each group, $g_i$ is the IV value of case $i$; $N=n_1+n_2$; $\alpha=0.05$; $m$ = number of rows in the output table; $\Phi$, $\varphi$ = standard normal CDF and density, $z_q=\Phi^{-1}(q)$; $t_{q,\,df}$ = $q$ quantile of the t distribution.

### report_ttests

**`n1`, `n2`**

$$n_1=\sum_i \mathbf{1}[g_i=\text{level1}],\qquad n_2=\sum_i \mathbf{1}[g_i=\text{level2}]$$

**`mean1`, `mean2`**

$$\bar{x}_1=\frac{1}{n_1}\sum_{i=1}^{n_1}x_{1i},\qquad \bar{x}_2=\frac{1}{n_2}\sum_{j=1}^{n_2}x_{2j}$$

**`sd1`, `sd2`**

$$s_1=\sqrt{\frac{1}{n_1-1}\sum_{i=1}^{n_1}(x_{1i}-\bar{x}_1)^2},\qquad s_2=\sqrt{\frac{1}{n_2-1}\sum_{j=1}^{n_2}(x_{2j}-\bar{x}_2)^2}$$

**`t`**

$$\text{Welch (default):}\quad t=\frac{\bar{x}_1-\bar{x}_2}{SE},\qquad SE=\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$

$$\text{Student (var.equal = TRUE):}\quad t=\frac{\bar{x}_1-\bar{x}_2}{SE},\qquad SE=s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}},\qquad s_p^2=\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{N-2}$$

**`df`**

$$\text{Welch:}\quad df=\frac{\left(\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}\right)^2}{\frac{(s_1^2/n_1)^2}{n_1-1}+\frac{(s_2^2/n_2)^2}{n_2-1}},\qquad \text{Student:}\quad df=N-2$$

**`p`**

$$p_{\text{two.sided}}=2\,P(T_{df}\ge\lvert t\rvert),\qquad p_{\text{less}}=P(T_{df}\le t),\qquad p_{\text{greater}}=P(T_{df}\ge t),\qquad T_{df}\sim t(df)$$

**`CI_l`, `CI_u`**

$$\text{two.sided:}\quad \left[(\bar{x}_1-\bar{x}_2)-t_{1-\alpha/2,\,df}\,SE,\ \ (\bar{x}_1-\bar{x}_2)+t_{1-\alpha/2,\,df}\,SE\right]$$

$$\text{less:}\quad \left(-\infty,\ \ (\bar{x}_1-\bar{x}_2)+t_{1-\alpha,\,df}\,SE\right],\qquad \text{greater:}\quad \left[(\bar{x}_1-\bar{x}_2)-t_{1-\alpha,\,df}\,SE,\ \ \infty\right)$$

**`sd_pooled`**

$$s_{pooled}=\sqrt{\frac{s_1^2+s_2^2}{2}}$$

**`d`**

$$d=\frac{\lvert\bar{x}_2-\bar{x}_1\rvert}{s_{pooled}}$$

**`r`**

$$r=\frac{d}{\sqrt{d^2+\frac{N^2}{n_1n_2}}}$$

**`k_squared[bartlett]`**

$$K^2=\frac{(N-2)\ln s_p^2-(n_1-1)\ln s_1^2-(n_2-1)\ln s_2^2}{1+\frac{1}{3}\left(\frac{1}{n_1-1}+\frac{1}{n_2-1}-\frac{1}{N-2}\right)},\qquad s_p^2=\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{N-2}$$

**`df[bartlett]`**

$$df_{bartlett}=k-1=2-1=1$$

**`p[bartlett]`**

$$p_{bartlett}=P\left(\chi^2_{1}\ge K^2\right)$$

**`bonferroni_p`**

$$\alpha_{bonferroni}=\frac{\alpha}{m}=\frac{0.05}{m}$$

**`significant`**

$$\text{significant}=\begin{cases}\text{TRUE} & p<\alpha_{bonferroni}\\ \text{FALSE} & p\ge\alpha_{bonferroni}\end{cases}$$

### report_wtests

`n1`, `n2`, `mean1`, `mean2`, `sd1`, `sd2`, `sd_pooled`, `d`, `k_squared[bartlett]`, `df[bartlett]`, `p[bartlett]`, `bonferroni_p`, `significant`: as in `report_ttests`.

**`W`**

$$W=R_1-\frac{n_1(n_1+1)}{2},\qquad R_1=\sum_{i=1}^{n_1}\operatorname{rank}(x_{1i})\ \ \text{(ranks in the pooled sample of } N \text{ values, ties averaged)}$$

**`p`**

$$\text{exact } (n_1<50,\ n_2<50)\text{:}\quad p_{\text{two.sided}}=\min\{1,\ 2P(W_0\le W),\ 2P(W_0\ge W)\},\qquad p_{\text{less}}=P(W_0\le W),\qquad p_{\text{greater}}=P(W_0\ge W)$$

$$\text{normal approximation (otherwise):}\quad p_{\text{two.sided}}=2\min\{\Phi(z),\ 1-\Phi(z)\},\qquad p_{\text{less}}=\Phi(z),\qquad p_{\text{greater}}=1-\Phi(z)$$

$$z=\frac{W-\frac{n_1n_2}{2}-c}{\sigma_W},\qquad \sigma_W=\sqrt{\frac{n_1n_2}{12}\left[(N+1)-\frac{\sum_k(t_k^3-t_k)}{N(N-1)}\right]},\qquad c=\begin{cases}0.5\,\operatorname{sign}\left(W-\frac{n_1n_2}{2}\right) & \text{two.sided}\\ 0.5 & \text{greater}\\ -0.5 & \text{less}\end{cases}$$

$W_0$ = null distribution of $W$ (conditional on the ties), $t_k$ = number of values in tie group $k$.

**`CI_l`, `CI_u`**

$$\text{exact, no ties:}\quad \text{two.sided: }\left[D_{(k)},\ D_{(n_1n_2-k+1)}\right],\quad \text{greater: }\left[D_{(k)},\ \infty\right),\quad \text{less: }\left(-\infty,\ D_{(n_1n_2-k+1)}\right]$$

$$k=\min\{w:\ P(W_0\le w)>\alpha/2\}\ \text{(two.sided)},\qquad k=\min\{w:\ P(W_0\le w)>\alpha\}\ \text{(one-sided)}$$

$$\text{normal approximation:}\quad \text{two.sided: } z(CI_l)=z_{1-\alpha/2},\ z(CI_u)=z_{\alpha/2},\qquad \text{greater: } z(CI_l)=z_{1-\alpha},\qquad \text{less: } z(CI_u)=z_{\alpha}$$

$D_{(1)}\le\dots\le D_{(n_1n_2)}$ = sorted differences $x_{1i}-x_{2j}$; $z(\Delta)$ = $z$ computed on $x_{1i}-\Delta$ and $x_{2j}$.

**`r`**

$$r=\frac{\lvert Z\rvert}{\sqrt{N}},\qquad Z=\Phi^{-1}\left(1-\frac{p}{2}\right)$$

## References

Bartlett, M. S. (1937). Properties of sufficiency and statistical tests. *Proceedings of the Royal Society of London. Series A, 160*(901), 268–282. https://doi.org/10.1098/rspa.1937.0109

Bonferroni, C. E. (1936). Teoria statistica delle classi e calcolo delle probabilità. *Pubblicazioni del R Istituto Superiore di Scienze Economiche e Commerciali di Firenze, 8*, 3–62.

Borenstein, M., Hedges, L. V., Higgins, J. P. T., & Rothstein, H. R. (2009). *Introduction to meta-analysis*. Wiley. https://doi.org/10.1002/9780470743386

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Dunn, O. J. (1961). Multiple comparisons among means. *Journal of the American Statistical Association, 56*(293), 52–64. https://doi.org/10.1080/01621459.1961.10482090

Hodges, J. L., & Lehmann, E. L. (1963). Estimates of location based on rank tests. *The Annals of Mathematical Statistics, 34*(2), 598–611. https://doi.org/10.1214/aoms/1177704172

Mann, H. B., & Whitney, D. R. (1947). On a test of whether one of two random variables is stochastically larger than the other. *The Annals of Mathematical Statistics, 18*(1), 50–60. https://doi.org/10.1214/aoms/1177730491

Rosenthal, R. (1991). *Meta-analytic procedures for social research* (Rev. ed.). Sage. https://doi.org/10.4135/9781412984997

Satterthwaite, F. E. (1946). An approximate distribution of estimates of variance components. *Biometrics Bulletin, 2*(6), 110–114. https://doi.org/10.2307/3002019

Sawilowsky, S. S. (2009). New effect size rules of thumb. *Journal of Modern Applied Statistical Methods, 8*(2), 597–599. https://doi.org/10.22237/jmasm/1257035100

Student. (1908). The probable error of a mean. *Biometrika, 6*(1), 1–25. https://doi.org/10.1093/biomet/6.1.1

Welch, B. L. (1947). The generalization of "Student's" problem when several different population variances are involved. *Biometrika, 34*(1–2), 28–35. https://doi.org/10.1093/biomet/34.1-2.28

Wilcoxon, F. (1945). Individual comparisons by ranking methods. *Biometrics Bulletin, 1*(6), 80–83. https://doi.org/10.2307/3001968
