---
title: "GLM Logistic Regression"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM Logistic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette documents the logistic regression functions from `working_functions/GLM_LOGISTIC_REGRESSION.R`: the logistic response function, a visualisation of a fitted logistic model against its predictors, a likelihood-ratio test for comparing nested models, and a full diagnostic report (`report_logistic`) covering pseudo-R², coefficients, residual diagnostics, and confusion-matrix-based classification performance.

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### The logistic function

Logistic regression models the probability of a binary outcome as

$$P(y=1\mid x) = \frac{1}{1+e^{-(\beta_0+\beta_1 x)}}$$

which maps any real-valued linear predictor onto the $(0,1)$ probability range via the S-shaped logistic curve.

### Maximum likelihood, deviance, and pseudo-R²

Unlike OLS regression, logistic regression coefficients are found by **maximum likelihood** — the coefficients that make the observed 0/1 outcomes most probable. **Deviance** ($-2 \times$ log-likelihood) plays the role that residual sum of squares plays in OLS; it follows a $\chi^2$ distribution and can be tested for significance.

Because there's no single, universally agreed-upon $R^2$ for logistic regression (unlike OLS, where $R^2$ has one unambiguous definition), several **pseudo-$R^2$** measures exist, all comparing the fitted model's deviance to the null (intercept-only) model's deviance:

- **Hosmer-Lemeshow $R^2_L$**: $1 - \text{deviance}_{model}/\text{deviance}_{null}$ — the proportional reduction in deviance.
- **Cox and Snell $R^2_{CS}$**: based on the likelihood ratio, but capped below 1 even for a perfect model.
- **Nagelkerke $R^2_N$**: a rescaling of Cox & Snell that stretches it to reach a true maximum of 1.

### The likelihood-ratio test

To test whether an additional predictor (or set of predictors) significantly improves a model, compare the deviance of two *nested* models:

$$X^2 = \text{deviance}_{smaller} - \text{deviance}_{larger} \sim \chi^2_{df}$$

with degrees of freedom equal to the difference in the number of parameters between the two models. A significant result means the larger model fits significantly better than the smaller one.

### Diagnostics

Logistic regression carries over the same influence diagnostics as OLS — standardized/studentized residuals, `dfbeta`, hat values (leverage), and VIF (multicollinearity) — interpreted the same way, but computed on the deviance/Pearson residual scale rather than raw residuals.

---

## The Logistic Function

```{r compute_y_logistic,fig.width=7,fig.height=5}
x <- -10:10
plot(x, compute_y_logistic(intercept = 0, coefficient = 1, x = x), type = "l",
     xlab = "x", ylab = "P(y=1)", main = "Logistic function (intercept=0, coefficient=1)")
grid()
abline(h = .5, lty = 2)
```

The curve crosses $P=0.5$ exactly at $x=-\text{intercept}/\text{coefficient}$ (here, $x=0$); larger `coefficient` values make the transition steeper.

## Plotting a Fitted Model

`plot_logistic_model` overlays a fitted logistic curve for every predictor in a data frame against a single binary outcome, letting you compare how strongly each predictor separates the two outcome classes.

```{r plot_logistic_model,fig.width=8,fig.height=6}
set.seed(12345)
df <- data.frame(
  outcome = c(rep(1, 10), rep(0, 10)),
  pd1 = c(rep(1, 11), rep(0, 9)),
  pd2 = c(rep(1, 9), rep(0, 11)),
  pc1 = c(rnorm(10, mean = 5), rnorm(10, mean = 10)),
  pc2 = c(rnorm(10, mean = 5), rnorm(10, mean = 20))
)
plot_logistic_model(df = df, base_size = 15)
```

`pc2` separates the two outcome groups far more sharply than `pc1` (means 5 vs. 20, rather than 5 vs. 10) — its fitted curve accordingly rises more steeply, visually previewing that it will be the stronger predictor once modeled formally.

## Comparing Nested Models

`output_compare_model_logistic` runs the likelihood-ratio test described above between two fitted `glm` models.

```{r compare_models}
modelcategoricalpredictor <- glm(case ~ education, data = infert, family = binomial)
modelcontinuouspredictor  <- glm(case ~ age, data = infert, family = binomial)
modeltwopredictors        <- glm(case ~ education * age, data = infert, family = binomial)

anova(modelcategoricalpredictor, modelcontinuouspredictor)
output_compare_model_logistic(model1 = modelcategoricalpredictor, model2 = modeltwopredictors)
output_compare_model_logistic(model1 = modelcontinuouspredictor, model2 = modeltwopredictors)
output_compare_model_logistic(model1 = modelcontinuouspredictor, model2 = modelcategoricalpredictor)
```

Note the last comparison: `model1` and `model2` here are **not nested** (`education` and `age` are different, non-overlapping predictor sets) — the likelihood-ratio test's $\chi^2$ approximation is only valid for nested models, so a non-nested comparison like this one should not be interpreted as a formal significance test, only as a raw deviance difference.

## Full Model Report

`report_logistic` runs the model through a comprehensive diagnostic pipeline: pseudo-$R^2$, coefficient table, residual/influence diagnostics, VIF, and confusion-matrix-based classification performance (with an optimal cut-point search).

```{r report_logistic,results='hide'}
model_simple <- glm(case ~ education, data = infert, family = binomial)
model_two    <- glm(case ~ education + stratum, data = infert, family = binomial)
result_simple <- report_logistic(model = model_simple)
result_two    <- report_logistic(model = model_two)
```

```{r report_logistic_output}
result_simple$result_X2_logistic
result_simple$result_R2_logistic
result_simple$coefficients
result_two$VIF
```

- **`result_X2_logistic`**: the omnibus test of the full model against the null (intercept-only) model — analogous to the F-test in OLS regression.
- **`result_R2_logistic`**: all three pseudo-$R^2$ measures side by side.
- **`coefficients`**: standard `glm` coefficient table (estimate, SE, z, p).
- **`VIF`**: only populated when the model has more than one predictor — flags multicollinearity among predictors (values above 10 are conventionally considered problematic).

### Validation on held-out data

`report_logistic` accepts a `validation_data` argument to evaluate classification performance on data the model wasn't fit to, rather than on its own training data (which otherwise gives an optimistic, resubstitution estimate of accuracy):

```{r validation,results='hide'}
result_validated <- report_logistic(model = model_simple, validation_data = infert)
```

```{r validation_output}
result_validated$score_descriptives
```

## Family Robustness

Because `report_logistic` operates on the generic `glm` object, it works across GLM families beyond binomial — though pseudo-$R^2$ and the confusion-matrix performance metrics are only meaningful for binary outcomes:

```{r other_families,results='hide'}
model_gaussian <- glm(case ~ education, data = infert, family = gaussian)
model_poisson  <- glm(case ~ education, data = infert, family = poisson)
result_gaussian <- report_logistic(model = model_gaussian)
result_poisson  <- report_logistic(model = model_poisson)
```

```{r other_families_output}
result_gaussian$logistic_output
result_poisson$logistic_output
```

---

## Conclusion

`compute_y_logistic` and `plot_logistic_model` make the shape of the fitted response curve visible and comparable across predictors; `output_compare_model_logistic` formalizes whether adding predictors significantly improves fit (for genuinely nested models); and `report_logistic` bundles pseudo-$R^2$, coefficients, influence diagnostics, and out-of-sample classification performance into a single report suitable for a full model write-up.

---

*Rendered with R `r getRversion()`*
