---
title: "GLM Correlation"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{GLM Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r library,message=FALSE,warning=FALSE}
source(paste0(directory, "code_rwf.R"))
```

## Description

This vignette documents the correlation functions from `working_functions/GLM_CORRELATION.R` in the `rwf` package: a correlation heatmap plotter, power analysis for correlation tests, a full bivariate correlation report (r, p, CI, adjusted p, sample size — all as lower-triangle matrices), and special correlation types for dichotomous/ordinal data (tetrachoric, polychoric, biserial, polyserial).

Installation instructions for rwf can be found on the [rwf GitHub repository](https://github.com/sedzinfo/rwf)

## Theoretical Background

### Pearson, Spearman, and Kendall correlation

The Pearson correlation coefficient measures *linear* association:

$$r = \frac{\sum (x_i-\bar x)(y_i-\bar y)}{\sqrt{\sum(x_i-\bar x)^2}\sqrt{\sum(y_i-\bar y)^2}}$$

Spearman's $\rho$ and Kendall's $\tau$ instead correlate the *ranks* of the data, making them robust to outliers and appropriate for monotonic-but-not-linear relationships or ordinal data.

### Multiple-comparison adjustment

When many pairwise correlations are tested at once (as in a full correlation matrix), the chance of at least one false positive grows with the number of tests. `report_correlation` adjusts p-values (Holm by default; also supports Bonferroni, Hochberg, BH/FDR, etc.) so the reported significance accounts for the whole matrix, not just one pair at a time.

### Statistical power for a correlation test

Power is the probability of detecting a true effect of a given size. For a bivariate correlation, power depends on the sample size $n$, the true correlation $r$, and the significance threshold $\alpha$ — `compute_power_r` sweeps over $n$ (via `pwr::pwr.r.test`) to show how power accumulates as sample size grows for a fixed $r$.

### Tetrachoric, polychoric, biserial, and polyserial correlation

Pearson correlation assumes continuous, normally distributed variables. When one or both variables are discrete, the naive Pearson correlation on the raw 0/1 or ordinal codes systematically *underestimates* the true association between the underlying continuous constructs. Four corrections exist depending on variable types:

| Correlation | X | Y |
|---|---|---|
| Tetrachoric | dichotomous (assumed to reflect an underlying continuous, normal latent variable) | dichotomous |
| Polychoric | ordinal (few categories) | ordinal |
| Biserial | continuous | dichotomous |
| Polyserial | continuous | ordinal |

All four estimate the correlation between the *latent continuous* variables assumed to underlie the observed discrete/dichotomous scores, rather than the raw observed correlation.

---

## Correlation Matrix Plot

`plot_corrplot` draws a correlation matrix as a colour-coded, annotated heatmap — only the lower triangle is shown (the matrix is symmetric, so the upper triangle is redundant).

```{r plot_corrplot,fig.width=8,fig.height=8,dev='png',fig.retina=2}
plot_corrplot(stats::cor(mtcars), title = "mtcars correlation matrix")
```

Blue cells indicate positive correlation, red negative, and white near-zero — `fill_limits` controls the midpoint and endpoints of this gradient (default `c(-1, 0, 1)`).

## Power Analysis for Correlation

`compute_power_r` sweeps sample size from 10 up to `n` and computes the statistical power to detect a correlation of magnitude `r` at the given significance level, via `pwr::pwr.r.test`.

```{r compute_power_r,fig.width=7,fig.height=5,dev='png',fig.retina=2}
result <- compute_power_r(n = 100, r = 0.5, sig.level = 0.05, alternative = "two.sided")
result$plot
head(result$power_table)
```

Power climbs steeply then flattens as it approaches 1 — the sample size at which the curve crosses the conventional 0.80 threshold is the minimum recommended sample for reliably detecting an effect of this size.

`compute_power_r_matrix` applies this to an entire correlation matrix at once, running the power sweep for the matrix's minimum, maximum, mean, and median absolute correlation (off-diagonal), and arranging all four power curves in one multiplot — a quick way to see the range of power your dataset's correlations imply.

```{r compute_power_r_matrix,fig.width=10,fig.height=8,dev='png',fig.retina=2}
result_matrix <- compute_power_r_matrix(m = stats::cor(mtcars, use = "pairwise.complete.obs"), n = 100)
result_matrix$plot
```

## Full Correlation Report

`report_correlation` wraps `psych::corr.test` and returns every quantity you'd need for a correlation table in a paper: r, r², p (raw and adjusted), t, n, and SE — each as a lower-triangle matrix — plus confidence intervals and (optionally) all pairwise scatterplots.

```{r report_correlation,fig.width=8,fig.height=8,dev='png',fig.retina=2,results='hide'}
result <- report_correlation(x = mtcars[, 1:5], scatterplot = TRUE)
```

```{r report_correlation_output}
result$r_lower
result$p_lower
result$p_lower_adjusted
result$n_lower
head(result$ci)
```

The `scatterplot` argument (when `TRUE`) additionally returns a full set of pairwise scatterplots (via `plot_scatterplot`), so you can visually confirm that a Pearson correlation is appropriate (i.e. the relationship looks linear, without strong outliers) before trusting the numeric table.

## Tetrachoric, Polychoric, Biserial, and Polyserial Correlation

`report_choric_serial` dispatches to the appropriate `psych` function based on `type`, and produces both a correlation heatmap and (when `file` is set) an Excel export.

```{r choric_serial_setup}
set.seed(12345)
binary_data <- generate_data(min = 0, max = 1, type = "uniform")
ordinal_data <- generate_data(min = 1, max = 5, type = "uniform")
```

```{r tetrachoric,fig.width=6,fig.height=6,dev='png',fig.retina=2}
result_tetrachoric <- report_choric_serial(binary_data, type = "tetrachoric")
result_tetrachoric$rho
```

```{r polychoric,fig.width=6,fig.height=6,dev='png',fig.retina=2}
result_polychoric <- report_choric_serial(ordinal_data, type = "polychoric")
result_polychoric$rho
```

```{r polyserial,fig.width=6,fig.height=6,dev='png',fig.retina=2}
result_polyserial <- report_choric_serial(x = psych::lsat6, y = psych::lsat6, type = "polyserial")
result_polyserial
```

```{r biserial,fig.width=6,fig.height=6,dev='png',fig.retina=2}
result_biserial <- report_choric_serial(x = psych::lsat6, y = psych::lsat6, type = "biserial")
result_biserial
```

Notice the tetrachoric/polychoric correlations tend to run *higher in magnitude* than a naive Pearson correlation on the same raw codes would — this is the latent-variable correction described above, not an inflation artifact.

---

## Conclusion

`plot_corrplot` gives a quick visual read of a correlation matrix; `compute_power_r`/`compute_power_r_matrix` tell you whether your sample size is adequate to detect the correlations you care about; `report_correlation` produces a full, publication-ready correlation table with multiple-comparison-adjusted significance; and `report_choric_serial` extends correlation analysis to dichotomous and ordinal variables via their appropriate latent-variable estimators.

---

*Rendered with R `r getRversion()`*
