---
title: "Airport Weather Analysis with NOAA ISD"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

## Overview

This tutorial shows how to download, clean, and visualise long-term weather data for any airport in the world using the **NOAA Integrated Surface Database (ISD)** — no FTP, no API key, no paid data.

The example uses **Athens Eleftherios Venizelos (LGAV)**, but every step works for any ICAO station code.

**Outputs:**

| Plot | What it shows |
|------|--------------|
| Daily ribbon | Temperature range (min–max) for every day |
| Monthly ribbon | Smoothed monthly range with gradient line |
| Heatmap max | Monthly max temperature by year |
| Heatmap min | Monthly min temperature by year |
| Decomposition | Trend + seasonal + residual components |

---

## Packages

```{r packages}
library(rnoaa)
library(dplyr)
library(lubridate)
library(ggplot2)
```

```{r install, eval=FALSE}
# remotes::install_github("ropensci/rnoaa")   # use the GitHub version
# install.packages(c("dplyr", "lubridate", "ggplot2"))
```

---

## 1. Find the station

NOAA ISD identifies stations by **USAF** and **WBAN** codes. The easiest way to find them is by ICAO airport code.

```{r station}
stations <- isd_stations(refresh = FALSE)

# explore Greek stations
stations[stations$ctry %in% "GR", c("station_name","icao","usaf","wban","lat","lon")]
```

```{r lgav}
lgav    <- stations[stations$icao %in% "LGAV", ]
station <- lgav$station_name
usaf    <- lgav$usaf[1]
wban    <- lgav$wban[1]

message("Station : ", station)
message("USAF    : ", usaf, "   WBAN: ", wban)
message("Location: ", round(lgav$lat[1], 3), "°N  ", round(lgav$lon[1], 3), "°E")
```

> Change `"LGAV"` to any ICAO code — `"EGLL"` for Heathrow, `"KJFK"` for New York, etc.

---

## 2. Download hourly ISD data

`rnoaa::isd()` replaces the old wget + FTP workflow. It downloads, decompresses, and parses the ISD binary format automatically. Each call fetches one year; a loop collects all years into a list.

```{r download}
years     <- 2001:2026
data_list <- list()

for (yr in years) {
  tryCatch({
    d <- rnoaa::isd(usaf = usaf, wban = wban, year = yr, progress = FALSE)
    data_list[[as.character(yr)]] <- d
    message("  downloaded ", yr, " (", nrow(d), " rows)")
  }, error = function(e) message("  skip ", yr, ": ", conditionMessage(e)))
}

df_raw <- bind_rows(data_list)
message("Total rows: ", nrow(df_raw))
```

> Years where data are unavailable are silently skipped by `tryCatch`. 2026 data may be partial depending on when you run this.

---

## 3. Clean: sentinel values and date parsing

### Why sentinel values matter

NOAA ISD encodes missing measurements as **9999** in the raw file. After dividing by 10 (ISD stores all values as tenths of the unit), missing temperature becomes **999.9°C** — a physically impossible value that will corrupt any aggregate (mean, max, min) if not removed.

```{r clean}
df_data <- df_raw |>
  mutate(
    date     = as.Date(date, format = "%Y%m%d"),   # isd() returns "YYYYMMDD" strings
    temp     = as.numeric(temperature) / 10,
    dew      = as.numeric(temperature_dewpoint) / 10,
    wind_spd = as.numeric(wind_speed) / 10,
    slp      = as.numeric(air_pressure) / 10
  ) |>
  mutate(across(c(temp, dew, wind_spd, slp),
                ~ ifelse(. > 900 | . < -200, NA, .))) |>
  filter(!is.na(date))

df_data[df_data == 999.9] <- NA          # catch any remaining sentinels
df_data$month <- sub("-\\d{2}$", "", df_data$date)   # "YYYY-MM" column
```

**Two-pass sentinel removal:**

1. `ifelse(. > 900 | . < -200, NA, .)` — catches the scaled sentinels immediately after division
2. `df_data[df_data == 999.9] <- NA` — safety net for any that slipped through

---

## 4. Daily and monthly aggregates

`plyr::ddply` with `numcolwise` applies a function (here `max` or `min`) to every numeric column, grouped by date or month. Columns are then renamed with `_max` / `_min` suffixes before merging.

```{r aggregate}
daily_max <- plyr::ddply(df_data, "date",  plyr::numcolwise(max, na.rm = TRUE))
daily_min <- plyr::ddply(df_data, "date",  plyr::numcolwise(min, na.rm = TRUE))
month_max <- plyr::ddply(df_data, "month", plyr::numcolwise(max, na.rm = TRUE))
month_min <- plyr::ddply(df_data, "month", plyr::numcolwise(min, na.rm = TRUE))

names(daily_max)[2:length(daily_max)] <- paste0(names(daily_max)[2:length(daily_max)], "_max")
names(daily_min)[2:length(daily_min)] <- paste0(names(daily_min)[2:length(daily_min)], "_min")
names(month_max)[2:length(month_max)] <- paste0(names(month_max)[2:length(month_max)], "_max")
names(month_min)[2:length(month_min)] <- paste0(names(month_min)[2:length(month_min)], "_min")

daily <- merge(daily_max, daily_min)
month <- merge(month_max, month_min)

daily$date       <- as.Date(daily$date)
month$month_date <- as.Date(paste0(month$month, "-01"))
month            <- month[order(month$month_date), ]
```

```{r aggregate_preview}
head(daily[, c("date","temp_max","temp_min","wind_spd_max")], 6)
```

---

## 5. Plots

### Daily temperature ribbon

`geom_ribbon` fills the area between the daily min and max. The fill colour is mapped to `temp_max` through a diverging gradient — cold days are blue, hot days are red.

```{r plot_daily}
plot_daily <- ggplot(daily, aes(x = date)) +
  geom_ribbon(aes(ymin = temp_min, ymax = temp_max, fill = temp_max), alpha = 0.7) +
  scale_fill_gradientn(
    colours = c("#2166ac","#74add1","#fee090","#f46d43","#a50026"),
    name    = "Max °C"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  geom_hline(yintercept = 0, color = "white", linewidth = 0.4, linetype = "dashed") +
  labs(
    title    = str_aes(station),
    subtitle = "Daily temperature range",
    x        = NULL, y = "Temperature (°C)",
    caption  = "Source: NOAA ISD via rnoaa"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.background    = element_rect(fill = "white", color = NA),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.minor.x = element_blank()
  )

plot_daily
```

### Monthly ribbon

Aggregating to monthly level removes the noise and makes the seasonal pattern clearer. The max temperature gets a coloured gradient line; the min stays white.

```{r plot_monthly}
plot_monthly <- ggplot(month, aes(x = month_date)) +
  geom_ribbon(aes(ymin = temp_min, ymax = temp_max), fill = "#74add1", alpha = 0.45) +
  geom_line(aes(y = temp_max, color = temp_max), linewidth = 0.8) +
  geom_line(aes(y = temp_min), color = "white", linewidth = 0.6, alpha = 0.8) +
  scale_color_gradientn(
    colours = c("#2166ac","#74add1","#fee090","#f46d43","#a50026"),
    name    = "Max °C"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  geom_hline(yintercept = 0, color = "white", linewidth = 0.4, linetype = "dashed") +
  labs(
    title    = str_aes(station),
    subtitle = "Monthly temperature range",
    x        = NULL, y = "Temperature (°C)",
    caption  = "Source: NOAA ISD via rnoaa"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.background    = element_rect(fill = "white", color = NA),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.minor.x = element_blank()
  )

plot_monthly
```

### Heatmaps

A calendar heatmap puts **year on the x-axis** and **month on the y-axis**, with temperature encoded as fill colour. This format is ideal for spotting anomalies — an unusually hot August or a cold February stands out immediately.

```{r heatmaps}
month$year <- substr(month$month, 1, 4)
month$mon  <- substr(month$month, 6, 7)

plot_heat_max <- ggplot(month, aes(x = year, y = mon, fill = temp_max)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(temp_max, 1), "°C")),
            color = "grey20", size = 3) +
  scale_fill_gradient2(
    low      = "#256abf", mid = "#f0efec", high = "#e34948",
    midpoint = median(month$temp_max, na.rm = TRUE),
    name     = "°C"
  ) +
  labs(title = "Monthly Max Temperature", x = NULL, y = NULL,
       caption = "Source: NOAA ISD via rnoaa") +
  theme_minimal(base_size = 20) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

plot_heat_min <- ggplot(month, aes(x = year, y = mon, fill = temp_min)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(temp_min, 1), "°C")),
            color = "grey20", size = 3) +
  scale_fill_gradient2(
    low      = "#256abf", mid = "#f0efec", high = "#e34948",
    midpoint = median(month$temp_min, na.rm = TRUE),
    name     = "°C"
  ) +
  labs(title = "Monthly Min Temperature", x = NULL, y = NULL,
       caption = "Source: NOAA ISD via rnoaa") +
  theme_minimal(base_size = 20) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

plot_heat_max
plot_heat_min
```

### All four together

```{r multiplot,fig.height = 20}
plot_multiplot(
  plotlist = list(plot_daily, plot_monthly, plot_heat_max, plot_heat_min),
  cols = 1
)
```

---

## 6. Seasonal decomposition

`decompose()` splits a time series into three additive components:

- **Trend** — the long-term direction (is the station getting warmer?)
- **Seasonal** — the repeating annual cycle (summer peaks, winter troughs)
- **Random** — what's left after removing trend and seasonal (unusual events, measurement noise)

The `frequency = 365` tells R the series repeats annually. `yday()` sets the correct starting position within the first year.

```{r decomp}
ts_temp <- ts(
  daily$temp_max,
  frequency = 365,
  start     = c(year(min(daily$date)), yday(min(daily$date)))
)

fit <- decompose(ts_temp)

autoplot(fit) +
  labs(title = paste("Seasonal decomposition —", str_aes(station), "daily max")) +
  theme_minimal(base_size = 14)
```

> **Reading the decomposition:** If the trend panel slopes upward over 20+ years, that is a local warming signal. The random panel should look like white noise — large spikes indicate extreme weather events.

---

## Adapting to another airport

Change only these two lines and re-run everything:

```{r other_station, eval=FALSE}
lgav    <- stations[stations$icao %in% "EGLL", ]   # London Heathrow
years   <- 2000:2026
```

Any ICAO code in the ISD station list will work. To search by country:

```{r search, eval=FALSE}
stations[stations$ctry %in% "UK", c("station_name","icao","usaf","wban")]
```

---

## Data source

| Resource | URL |
|----------|-----|
| NOAA ISD | `https://www.ncei.noaa.gov/products/land-based-station/integrated-surface-database` |
| rnoaa package | `https://github.com/ropensci/rnoaa` |
| Station list | `isd_stations()` — cached locally after first call |

---

*Rendered with R `r getRversion()` · rnoaa · ggplot2 · plyr · lubridate*
