---
title: "R Colors: Names, RGB/HSV, and Building a Color Chart"
author: "efg, Stowers Institute for Medical Research (original script); annotated version"
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

# Background

R ships with a built-in palette of named colors, returned by `colors()`. As of this
writing (and for a long time now) it holds **657** names — things like `"steelblue"`,
`"tomato3"`, or `"gray42"`. This document walks through a classic script by efg
(Stowers Institute) that turns that list into a printable reference chart, and along
the way explains the underlying R color machinery: how names map to RGB, how RGB maps
to HSV, and how to pick readable text/background combinations automatically.

The original script writes everything to a file called `COLOR_CHART.pdf`. Here we
reproduce each piece as an inline, explained chunk so it renders directly in this
document; the full original (PDF-producing) script is included as an appendix at the
end for reference.

```{r sanity-check}
# The chart layout below hard-codes 25 columns x 27 rows = 675 slots for 657 colors.
# This assumes the palette size hasn't changed between R versions.
length(colors())
```

# How R represents colors

R gives you three interchangeable ways to specify a color:

1. **By name** — a string from `colors()`, e.g. `"firebrick"`.
2. **By hex code** — `"#FF0000"`, the familiar web-style RRGGBB (optionally with alpha, RRGGBBAA).
3. **By numeric index** — a legacy palette (`palette()`), rarely used now.

`col2rgb()` converts any of the first two into the underlying 0–255 RGB triplet:

```{r col2rgb-demo}
col2rgb("yellow")
col2rgb("steelblue")
```

RGB is how the color is stored, but it is not how humans reason about color. For
that, R (via `rgb2hsv()`) can convert to **HSV** — Hue, Saturation, Value — which
separates *what color* (hue, as an angle 0–1 representing 0–360°) from *how vivid*
(saturation) and *how bright* (value):

```{r rgb2hsv-demo}
rgb2hsv(col2rgb("steelblue"))
```

Sorting colors by hue-then-saturation-then-value is what makes the second chart
below ("sorted by hue") group visually similar colors together, instead of the
arbitrary alphabetical-ish order `colors()` returns them in.

# Choosing readable text color: a contrast heuristic

Rectangles need a label (the color's name and index) drawn *on top of* them. White
text on `"yellow"` is unreadable; black text on `"navy"` is unreadable. The script
uses a cheap luminance heuristic: average the R, G, and B channels, and if that
average is above the midpoint (127.5), the background is "light" so use black text;
otherwise use white text.

```{r contrast-function}
SetTextContrastColor <- function(color) {
  ifelse(mean(col2rgb(color)) > 127, "black", "white")
}

SetTextContrastColor("white")
SetTextContrastColor("black")
SetTextContrastColor("red")
SetTextContrastColor("yellow")
```

This is a simplification — perceptual luminance weights green much more heavily
than blue (roughly `0.299*R + 0.587*G + 0.114*B`), so a plain channel average can
misjudge some saturated blues or reds. It's good enough for a quick reference
chart, though, which is the point here.

We precompute the contrast color for every entry in `colors()` once, so the
plotting loops below can just look it up instead of recomputing it per rectangle:

```{r contrast-vector}
TextContrastColor <- unlist(lapply(colors(), SetTextContrastColor))
table(TextContrastColor)
```

# Chart 1a: colors in index order

The simplest chart just walks `colors()` in the order R returns it, laying out
25 colors per row. Each rectangle is drawn with `rect()`, positioned by its
row/column, filled with the color itself, and labeled with its numeric index
(`SetTextContrastColor` decides whether that label is black or white).

```{r chart-index-order, fig.height=11}
colCount <- 25  # number of rectangles per row
rowCount <- ceiling(length(colors()) / colCount)

oldpar <- par(mar = c(1, 1, 2, 1))

plot(c(1, colCount), c(0, rowCount), type = "n", ylab = "", xlab = "",
     axes = FALSE, ylim = c(rowCount, 0))
title("R colors (index order)")

for (j in 0:(rowCount - 1)) {
  base <- j * colCount
  remaining <- length(colors()) - base
  RowSize <- ifelse(remaining < colCount, remaining, colCount)
  rect((1:RowSize) - 0.5, j - 0.5, (1:RowSize) + 0.5, j + 0.5,
       border = "black", col = colors()[base + (1:RowSize)])
  text((1:RowSize), j, paste(base + (1:RowSize)), cex = 0.7,
       col = TextContrastColor[base + (1:RowSize)])
}

par(oldpar)
```

Notice how little visual structure there is: R's `colors()` order is alphabetical
by name (with numbered variants like `"blue1"`..`"blue4"` interleaved), not by
appearance, so neighboring cells jump around the color wheel.

# Chart 1b: colors sorted by hue, saturation, value

To get a chart that actually reads as a gradient, we convert every color to HSV
and sort by hue first, then saturation, then value. `order()` gives us the
permutation of indices that achieves that sort, and we index into `colors()`
with it.

```{r chart-hue-order, fig.height=11}
RGBColors <- col2rgb(colors())
HSVColors <- rgb2hsv(RGBColors[1, ], RGBColors[2, ], RGBColors[3, ], maxColorValue = 255)
HueOrder <- order(HSVColors[1, ], HSVColors[2, ], HSVColors[3, ])

oldpar <- par(mar = c(1, 1, 2, 1))

plot(0, type = "n", ylab = "", xlab = "", axes = FALSE,
     ylim = c(rowCount, 0), xlim = c(1, colCount))
title("R colors -- sorted by hue, saturation, value")

for (j in 0:(rowCount - 1)) {
  for (i in 1:colCount) {
    k <- j * colCount + i
    if (k <= length(colors())) {
      rect(i - 0.5, j - 0.5, i + 0.5, j + 0.5,
           border = "black", col = colors()[HueOrder[k]])
      text(i, j, paste(HueOrder[k]), cex = 0.7,
           col = TextContrastColor[HueOrder[k]])
    }
  }
}

par(oldpar)
```

The label in each cell is still the *original* index into `colors()` — so you can
look up a color visually here, then use that number (or better, `colors()[index]`)
to reference it by name in code.

# Getting a color's hex and decimal RGB

For a reference sheet you usually want the exact hex code (for CSS/HTML/design
tools) alongside the decimal RGB triplet (for R/plotting code). `sprintf()` builds
both from `col2rgb()`'s output in one line:

```{r hex-and-decimal}
GetColorHexAndDecimal <- function(color) {
  c <- col2rgb(color)
  sprintf("#%02X%02X%02X %3d %3d %3d", c[1], c[2], c[3], c[1], c[2], c[3])
}

GetColorHexAndDecimal("yellow")
GetColorHexAndDecimal("steelblue")
```

# Chart 2: paginated detail sheet (name + hex + decimal)

The final part of the original script produces a multi-page listing: two columns
of 50 colors each per page, each row showing a filled swatch, the color's index,
its name, and its hex/decimal RGB (in a monospace font so the numbers align).
Below is a single representative page reproduced inline; the appendix has the
full paging loop that generates every page (this is what the original script
writes across the multi-page PDF).

```{r chart-detail-page, fig.height=11, fig.width=8}
index <- paste(1:length(colors()))
HexAndDec <- unlist(lapply(colors(), GetColorHexAndDecimal))
PerColumn <- 50
PerPage <- 2 * PerColumn

page <- 1
oldpar <- par(mar = c(1, 1, 1, 1))

plot(0, type = "n", ylab = "", xlab = "", axes = FALSE,
     ylim = c(PerColumn, 0), xlim = c(0, 1))
title("R colors")
mtext(paste("page", page), side = 1, adj = 1, line = -1)

base <- PerPage * (page - 1)

# Column 1
remaining <- length(colors()) - base
ColumnSize <- ifelse(remaining < PerColumn, remaining, PerColumn)
rect(0.00, 0:(ColumnSize - 1), 0.49, 1:ColumnSize,
     border = "black", col = colors()[(base + 1):(base + ColumnSize)])
text(0.045, 0.45 + (0:(ColumnSize - 1)), adj = 1, index[(base + 1):(base + ColumnSize)],
     cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
text(0.06, 0.45 + (0:(ColumnSize - 1)), adj = 0, colors()[(base + 1):(base + ColumnSize)],
     cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
save <- par(family = "mono")
text(0.25, 0.45 + (0:(ColumnSize - 1)), adj = 0, HexAndDec[(base + 1):(base + ColumnSize)],
     cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
par(save)

# Column 2
if (remaining > PerColumn) {
  remaining2 <- remaining - PerColumn
  ColumnSize2 <- ifelse(remaining2 < PerColumn, remaining2, PerColumn)
  rect(0.51, 0:(ColumnSize2 - 1), 1.00, 1:ColumnSize2,
       border = "black", col = colors()[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)])
  text(0.545, 0.45 + (0:(ColumnSize2 - 1)), adj = 1,
       index[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)],
       cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)])
  text(0.56, 0.45 + (0:(ColumnSize2 - 1)), adj = 0,
       colors()[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)],
       cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)])
  save <- par(family = "mono")
  text(0.75, 0.45 + (0:(ColumnSize2 - 1)), adj = 0,
       HexAndDec[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)],
       cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize2)])
  par(save)
}

par(oldpar)
```

To generate *every* page (not just page 1) and get the full multi-page reference
sheet, wrap that same block in a loop over `page in 1:ceiling(length(colors()) / PerPage)`
and, when knitting, either emit one PDF page per iteration or (for HTML) let each
iteration produce its own plot — see the appendix for the loop as written in the
original script.

# Takeaways

- `colors()` is R's built-in named palette (657 entries); `col2rgb()` and
  `rgb2hsv()` convert between name/hex, RGB, and HSV representations.
- Sorting by HSV hue (not the default alphabetical order) is what makes a color
  chart visually navigable.
- A simple average-of-channels test is a cheap-but-imperfect way to auto-pick
  readable black/white text on an arbitrary background color.
- `sprintf("#%02X%02X%02X ...")` is the idiomatic way to format an RGB triplet
  as both hex and decimal in one string.

# Appendix: original script (writes `COLOR_CHART.pdf`)

```{r original-script, eval=FALSE}
# efg, Stowers Institute for Medical Research
# efg's Research Notes: http://research.stowers-institute.org/efg/R/Color/Chart 6 July 2004. Modified 23 May 2005.
pdf("COLOR_CHART.pdf", width = 10, height = 10)
# save to reset at end
oldparameters <- par(mar = c(1, 1, 2, 1), mfrow = c(2, 1))
# Be cautious in case definition of "colors" changes. Use some hard-coded constants since this is not expected to change.
stopifnot(length(colors()) == 657)
# 0. Setup
SetTextContrastColor <- function(color) {ifelse(mean(col2rgb(color)) > 127, "black", "white")}
TextContrastColor <- unlist(lapply(colors(), SetTextContrastColor))
# 1a. Plot matrix of R colors, in index order, 25 per row.
colCount <- 25
rowCount <- 27
plot(c(1, colCount), c(0, rowCount), type = "n", ylab = "", xlab = "", axes = FALSE, ylim = c(rowCount, 0))
title("R colors")
mtext("http://research.stowers-institute.org/efg/R/Color/Chart", cex = 0.6)
for (j in 0:(rowCount - 1)) {
  base <- j * colCount
  remaining <- length(colors()) - base
  RowSize <- ifelse(remaining < colCount, remaining, colCount)
  rect((1:RowSize) - 0.5, j - 0.5, (1:RowSize) + 0.5, j + 0.5, border = "black", col = colors()[base + (1:RowSize)])
  text((1:RowSize), j, paste(base + (1:RowSize)), cex = 0.7, col = TextContrastColor[base + (1:RowSize)])
}
# 1b. Plot matrix of R colors, in "hue" order, 25 per row.
RGBColors <- col2rgb(colors()[1:length(colors())])
HSVColors <- rgb2hsv(RGBColors[1, ], RGBColors[2, ], RGBColors[3, ], maxColorValue = 255)
HueOrder <- order(HSVColors[1, ], HSVColors[2, ], HSVColors[3, ])
plot(0, type = "n", ylab = "", xlab = "", axes = FALSE, ylim = c(rowCount, 0), xlim = c(1, colCount))
title("R colors -- Sorted by Hue, Saturation, Value")
for (j in 0:(rowCount - 1)) {
  for (i in 1:colCount) {
    k <- j * colCount + i
    if (k <= length(colors())) {
      rect(i - 0.5, j - 0.5, i + 0.5, j + 0.5, border = "black", col = colors()[HueOrder[k]])
      text(i, j, paste(HueOrder[k]), cex = 0.7, col = TextContrastColor[HueOrder[k]])
    }
  }
}
# 2. Create 7-page color chart showing rectangle block of color, along with index, color name, and RGB constants in hex and decimal.
GetColorHexAndDecimal <- function(color) {
  c <- col2rgb(color)
  sprintf("#%02X%02X%02X %3d %3d %3d", c[1], c[2], c[3], c[1], c[2], c[3])
}
par(oldparameters)
oldparameters <- par(mar = c(1, 1, 1, 1))
index <- paste(1:length(colors()))
HexAndDec <- unlist(lapply(colors(), GetColorHexAndDecimal))
PerColumn <- 50
PerPage <- 2 * PerColumn
for (page in 1:(trunc(length(colors()) + (PerPage - 1)) / PerPage)) {
  plot(0, type = "n", ylab = "", xlab = "", axes = FALSE, ylim = c(PerColumn, 0), xlim = c(0, 1))
  title("R colors")
  mtext(paste("page ", page), SOUTH <- 1, adj = 1, line = -1)
  base <- PerPage * (page - 1)
  # Column 1
  remaining <- length(colors()) - base
  ColumnSize <- ifelse(remaining < PerColumn, remaining, PerColumn)
  rect(0.00, 0:(ColumnSize - 1), 0.49, 1:ColumnSize, border = "black", col = colors()[(base + 1):(base + ColumnSize)])
  text(0.045, 0.45 + (0:(ColumnSize - 1)), adj = 1, index[(base + 1):(base + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
  text(0.06, 0.45 + (0:(ColumnSize - 1)), adj = 0, colors()[(base + 1):(base + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
  save <- par(family = "mono")
  text(0.25, 0.45 + (0:(ColumnSize - 1)), adj = 0, HexAndDec[(base + 1):(base + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + 1):(base + ColumnSize)])
  par(save)
  # Column 2
  if (remaining > PerColumn) {
    remaining <- remaining - PerColumn
    ColumnSize <- ifelse(remaining < PerColumn, remaining, PerColumn)
    rect(0.51, 0:(ColumnSize - 1), 1.00, 1:ColumnSize, border = "black", col = colors()[(base + PerColumn + 1):(base + PerColumn + ColumnSize)])
    text(0.545, 0.45 + (0:(ColumnSize - 1)), adj = 1, index[(base + PerColumn + 1):(base + PerColumn + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize)])
    text(0.56, 0.45 + (0:(ColumnSize - 1)), adj = 0, colors()[(base + PerColumn + 1):(base + PerColumn + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize)])
    save <- par(family = "mono")
    text(0.75, 0.45 + (0:(ColumnSize - 1)), adj = 0, HexAndDec[(base + PerColumn + 1):(base + PerColumn + ColumnSize)], cex = 0.6, col = TextContrastColor[(base + PerColumn + 1):(base + PerColumn + ColumnSize)])
    par(save)
  }
}
par(oldparameters)
dev.off()
```
