---
title: "Arctic Sea Ice Is Melting"
subtitle: "Visualising 45 years of decline with real NSIDC data"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

## Background

Every September the Arctic sea ice reaches its annual minimum after the summer melt season. The **National Snow and Ice Data Center (NSIDC)** has been tracking this since 1979 using satellite passive-microwave sensors. Their **Sea Ice Index v4.0** is the authoritative public record.

This document shows you exactly how to download the real data and reproduce the charts below — no manual downloads, no hardcoded numbers.

---

## Packages

```{r packages}
# Install any missing packages before running
# install.packages(c("ggplot2", "dplyr", "sf", "rnaturalearth", "rnaturalearthdata"))

library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
```

---

## Part 1 — Time series

### Download the data

NSIDC publishes one CSV per month. September (month 09) is the annual minimum.

```{r download-csv}
csv_url <- "https://noaadata.apps.nsidc.org/NOAA/G02135/north/monthly/data/N_09_extent_v4.0.csv"

arctic <- read.csv(csv_url, strip.white = TRUE) |>
  select(year, extent) |>
  filter(year >= 1979)

head(arctic)
```

### Plot

```{r timeseries}
avg_baseline <- arctic |> filter(year <= 1990) |> pull(extent) |> mean()
loss_pct     <- round((1 - min(arctic$extent) / avg_baseline) * 100)

col_fill   <- "#B3E5FC"
col_line   <- "#0277BD"
col_trend  <- "#C62828"
col_record <- "#E65100"
col_ref    <- "#78909C"
col_title  <- "#0D1B40"

records <- filter(arctic, year %in% c(2007, 2012, 2020))

ggplot(arctic, aes(x = year, y = extent)) +
  geom_area(fill = col_fill, alpha = 0.55) +
  geom_hline(yintercept = avg_baseline, linetype = "dotted",
             color = col_ref, linewidth = 0.7) +
  annotate("text", x = 1981, y = avg_baseline + 0.18,
           label = "1979-1990 mean", size = 2.9, color = col_ref, hjust = 0) +
  geom_smooth(method = "lm", se = TRUE,
              color = col_trend, fill = "#FFCDD2", alpha = 0.25,
              linewidth = 1.4, linetype = "dashed") +
  geom_line(color = col_line, linewidth = 1.1) +
  geom_point(data = records, color = col_record, fill = col_record,
             size = 4, shape = 21, stroke = 1.8) +
  annotate("segment", x = 2012, xend = 2012,
           y = min(arctic$extent), yend = min(arctic$extent) - 0.6,
           color = col_record, linewidth = 0.6) +
  annotate("text", x = 2012, y = min(arctic$extent) - 0.75,
           label = paste0("Record low\n", round(min(arctic$extent), 2),
                          " M km²  (", arctic$year[which.min(arctic$extent)], ")"),
           color = col_record, size = 2.9, fontface = "bold", hjust = 0.5) +
  annotate("text", x = max(arctic$year), y = max(arctic$extent) * 0.93,
           label = paste0("-", loss_pct, "% vs.\n1979-1990 avg"),
           color = col_trend, size = 4.5, fontface = "bold", hjust = 1) +
  labs(
    title    = "Arctic Sea Ice Is Melting",
    subtitle = "September minimum sea ice extent, 1979 to present",
    x        = NULL,
    y        = "Extent  (million km²)",
    caption  = "Source: NSIDC Sea Ice Index v4.0 | noaadata.apps.nsidc.org"
  ) +
  scale_x_continuous(breaks = seq(1980, 2030, 5), expand = c(0.01, 0)) +
  scale_y_continuous(labels = function(x) paste0(x, " M")) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F0F8FF", color = NA),
    panel.grid.major = element_line(color = "gray88", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 17, color = col_title),
    plot.subtitle    = element_text(size = 11, color = "gray45"),
    plot.caption     = element_text(size = 8, color = "gray60", hjust = 1),
    axis.text        = element_text(color = "gray40", size = 9),
    axis.title.y     = element_text(size = 10, color = "gray40", margin = margin(r = 8)),
    plot.margin      = margin(18, 18, 12, 18)
  )
```

---

## Part 2 — Maps (overview)

### How the shapefiles work

NSIDC also publishes the sea ice boundary as a polygon shapefile for every month. We download and cache them locally, then read with `sf`.

```{r shapefile-function}
month_dirs <- c("01_Jan","02_Feb","03_Mar","04_Apr","05_May","06_Jun",
                "07_Jul","08_Aug","09_Sep","10_Oct","11_Nov","12_Dec")

fetch_ice_extent <- function(year, month = "09", pole = "N",
                             tmpdir = file.path(tempdir(), "nsidc_ice")) {
  dir.create(tmpdir, showWarnings = FALSE)
  mm       <- sprintf("%02d", as.integer(month))
  yymm     <- paste0(year, mm)
  pole_dir <- ifelse(pole == "N", "north", "south")
  mon_dir  <- month_dirs[as.integer(mm)]
  url <- paste0(
    "https://noaadata.apps.nsidc.org/NOAA/G02135/",
    pole_dir, "/monthly/shapefiles/shp_extent/", mon_dir, "/",
    "extent_", pole, "_", yymm, "_polygon_v4.0.zip"
  )
  zip_path <- file.path(tmpdir, paste0("ice_", pole, "_", yymm, ".zip"))
  if (!file.exists(zip_path) || file.size(zip_path) < 1000) {
    message("Downloading ", year, "-", mm, " ...")
    tryCatch(
      download.file(url, zip_path, quiet = TRUE, mode = "wb"),
      error = function(e) { message("  failed: ", e$message); return(NULL) }
    )
  }
  if (!file.exists(zip_path) || file.size(zip_path) < 1000) return(NULL)
  exdir <- file.path(tmpdir, paste0("ice_", pole, "_", yymm))
  suppressWarnings(unzip(zip_path, exdir = exdir, overwrite = FALSE))
  shp <- list.files(exdir, "\\.shp$", full.names = TRUE)[1]
  if (is.na(shp)) return(NULL)
  shape <- tryCatch(sf::st_read(shp, quiet = TRUE), error = function(e) NULL)
  if (is.null(shape)) return(NULL)
  shape$year <- year
  shape
}
```

### Download shapefiles (every 5 years)

```{r download-shp-overview}
years_overview <- c(1990, 1995, 2000, 2005, 2010, 2015, 2020, 2024)

ice_list <- lapply(years_overview, fetch_ice_extent, month = "09", pole = "N")
ice_list <- Filter(Negate(is.null), ice_list)
ice_all  <- do.call(rbind, ice_list)
```

### Build the map

```{r map-overview}
world     <- ne_countries(scale = "medium", returnclass = "sf")
crs_polar <- "+proj=stere +lat_0=90 +lat_ts=70 +lon_0=0 +datum=WGS84 +units=m"

ice_proj   <- st_transform(ice_all,  crs_polar)
world_proj <- st_transform(world,    crs_polar)

clip_large <- st_buffer(
  st_sfc(st_point(c(0, 90)), crs = 4326) |> st_transform(crs_polar),
  dist = 1e7
)
world_clip <- st_intersection(world_proj, clip_large)

ggplot() +
  geom_sf(data = clip_large, fill = "#AED6F1", color = NA) +
  geom_sf(data = ice_proj,   fill = "white",  color = "white",   linewidth = 0.1) +
  geom_sf(data = world_clip, fill = "#8D6E63", color = "#6D4C41", linewidth = 0.15) +
  facet_wrap(~year, ncol = 4) +
  labs(
    title   = "Arctic Sea Ice Extent - September",
    caption = "Source: NSIDC Sea Ice Index v4.0"
  ) +
  theme_void(base_size = 10) +
  theme(
    plot.background = element_rect(fill = "#1A237E", color = NA),
    strip.text      = element_text(color = "white", face = "bold", size = 9),
    plot.title      = element_text(color = "white", face = "bold", size = 14,
                                   margin = margin(12, 0, 8, 12)),
    plot.caption    = element_text(color = "#546E7A", size = 7, hjust = 1,
                                   margin = margin(8, 12, 8, 0)),
    panel.spacing   = unit(0.4, "lines")
  )
```

---

## Part 3 — Maps (zoomed comparison)

Four key years in a closer view, with the actual extent area computed from the polygon geometry.

```{r download-shp-zoom}
years_zoom <- c(1985, 2000, 2012, 2024)

ice_zoom_list <- lapply(years_zoom, fetch_ice_extent, month = "09", pole = "N")
ice_zoom_list <- Filter(Negate(is.null), ice_zoom_list)
ice_zoom_all  <- do.call(rbind, ice_zoom_list)
```

```{r map-zoom, fig.width=14, fig.height=5}
ice_zoom_proj <- st_transform(ice_zoom_all, crs_polar)

clip_tight <- st_buffer(
  st_sfc(st_point(c(0, 90)), crs = 4326) |> st_transform(crs_polar),
  dist = 5.5e6
)
world_clip_tight <- st_intersection(world_proj, clip_tight)
ice_zoom_clip    <- st_intersection(ice_zoom_proj, clip_tight)

# Compute area per year (million km²) from the actual geometry
ice_areas <- sapply(years_zoom, function(y) {
  sub <- ice_zoom_proj[ice_zoom_proj$year == y, ]
  if (nrow(sub) == 0) return(NA)
  as.numeric(sum(st_area(sub))) / 1e12
})
base_area <- ice_areas[1]

labels_df <- data.frame(
  year  = years_zoom,
  label = ifelse(seq_along(years_zoom) == 1,
                 sprintf("%.2f M km²\n(baseline)", ice_areas),
                 sprintf("%.2f M km²\n(-%.0f%% vs %d)",
                         ice_areas, (1 - ice_areas / base_area) * 100, years_zoom[1]))
)
ice_zoom_clip <- merge(ice_zoom_clip, labels_df, by = "year")

ggplot() +
  geom_sf(data = clip_tight,        fill = "#AED6F1", color = NA) +
  geom_sf(data = ice_zoom_clip,     fill = "white",   color = "#c8e0f0", linewidth = 0.15) +
  geom_sf(data = world_clip_tight,  fill = "#6D4C41", color = "#4E342E", linewidth = 0.2) +
  geom_sf_text(
    data = ice_zoom_clip |> group_by(year) |> slice(1),
    aes(label = label, geometry = geometry),
    stat = "sf_coordinates",
    color = "white", size = 3, fontface = "bold",
    nudge_x = -3.2e6, nudge_y = -4.0e6, hjust = 0
  ) +
  facet_wrap(~year, ncol = 4) +
  labs(
    title    = "Arctic Sea Ice Shrinkage - September Minimum",
    subtitle = "Polar stereographic projection, zoomed to 55N+",
    caption  = "Source: NSIDC Sea Ice Index v4.0"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "#0D1B2A", color = NA),
    strip.text      = element_text(color = "white",  face = "bold", size = 12),
    plot.title      = element_text(color = "white",  face = "bold", size = 16,
                                   margin = margin(14, 0, 4, 14)),
    plot.subtitle   = element_text(color = "#90CAF9", size = 9,
                                   margin = margin(0, 0, 10, 14)),
    plot.caption    = element_text(color = "#546E7A", size = 7.5, hjust = 1,
                                   margin = margin(8, 14, 10, 0)),
    panel.spacing   = unit(0.8, "lines")
  )
```

---

## Data source

All data in this document comes directly from **NSIDC Sea Ice Index v4.0**:

- **Monthly extent CSV**: `https://noaadata.apps.nsidc.org/NOAA/G02135/north/monthly/data/`
- **Monthly shapefiles**: `https://noaadata.apps.nsidc.org/NOAA/G02135/north/monthly/shapefiles/shp_extent/`

> Fetterer, F., K. Knowles, W. N. Meier, M. Savoie, and A. K. Windnagel. 2017, updated daily. *Sea Ice Index, Version 3*. Boulder, Colorado USA. NSIDC: National Snow and Ice Data Center. https://doi.org/10.7265/N5K072F8.
