---
title: "BFI-44 Validation"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: "%\\VignetteIndexEntry{BFI-44 Validation} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

```{r options, echo=FALSE, warning=FALSE, message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r load, include=FALSE}
source(paste0(directory, "code_rwf.R"))
library(psych)
library(ggplot2)
library(reshape2)

load("data/df_personality.rda")

# Factor structure
factors <- list(
  Extraversion      = c(1, 6, 11, 16, 21, 26, 31, 36),
  Agreeableness     = c(2, 7, 12, 17, 22, 27, 32, 37, 42),
  Conscientiousness = c(3, 8, 13, 18, 23, 28, 33, 38, 43),
  Neuroticism       = c(4, 9, 14, 19, 24, 29, 34, 39),
  Openness          = c(5, 10, 15, 20, 25, 30, 35, 40, 41, 44)
)

# Reverse-scored items (1-6 scale: subtract from 7)
reverse_items <- c(2, 6, 8, 9, 12, 18, 21, 23, 24, 27, 31, 34, 35, 37, 41, 43)

df <- df_personality
for (i in reverse_items) {
  col <- sprintf("pers%02d", i)
  df[[col]] <- 7 - df[[col]]
}

# Rename columns: dimension prefix + item number + R if reverse-scored
dim_prefix <- c(Extraversion="E", Agreeableness="A",
                Conscientiousness="C", Neuroticism="N", Openness="O")
item_name_map <- character(44)
for (f in names(factors)) {
  for (i in factors[[f]]) {
    r <- if (i %in% reverse_items) "R" else ""
    item_name_map[i] <- sprintf("%s%02d%s", dim_prefix[f], i, r)
  }
}
names(item_name_map) <- sprintf("pers%02d", 1:44)
colnames(df) <- item_name_map[colnames(df_personality)]

factor_order <- unlist(lapply(names(factors), function(f) {
  item_name_map[sprintf("pers%02d", factors[[f]])]
}))
item_labels        <- unlist(lapply(names(factors), function(f) rep(f, length(factors[[f]]))))
names(item_labels) <- factor_order

# Lookup table for item descriptions
df_lookup <- data.frame(
  code = sprintf("pers%02d", 1:44),
  label = item_name_map,
  dimension = item_labels[item_name_map],
  reverse = sprintf("pers%02d", 1:44) %in% sprintf("pers%02d", reverse_items),
  description = c(
    "Is talkative",
    "Tends to find fault with others",
    "Does a thorough job",
    "Is depressed, blue",
    "Is original, comes up with new ideas",
    "Is reserved",
    "Is helpful and unselfish with others",
    "Can be somewhat careless",
    "Is relaxed, handles stress well",
    "Is curious about many different things",
    "Is full of energy",
    "Starts quarrels with others",
    "Is a reliable worker",
    "Can be tense",
    "Is ingenious, a deep thinker",
    "Generates a lot of enthusiasm",
    "Has a forgiving nature",
    "Tends to be disorganized",
    "Worries a lot",
    "Has an active imagination",
    "Tends to be quiet",
    "Is generally trusting",
    "Tends to be lazy",
    "Is emotionally stable, not easily upset",
    "Is inventive",
    "Has an assertive personality",
    "Can be cold and aloof",
    "Perseveres until the task is finished",
    "Can be moody",
    "Values artistic, aesthetic experiences",
    "Is sometimes shy, inhibited",
    "Is considerate and kind to almost everyone",
    "Does things efficiently",
    "Remains calm in tense situations",
    "Prefers work that is routine",
    "Is outgoing, sociable",
    "Is sometimes rude to others",
    "Makes plans and follows through with them",
    "Gets nervous easily",
    "Likes to reflect, play with ideas",
    "Has few artistic interests",
    "Likes to cooperate with others",
    "Is easily distracted",
    "Is sophisticated in art, music, or literature"
  ),
  stringsAsFactors = FALSE
)
```

## Description

The **Big Five Inventory (BFI-44)** was developed by John & Srivastava (1999) as a short measure of the five major dimensions of personality. Participants rated how much each statement applied to them on a scale from 1 (*Disagree strongly*) to 6 (*Agree strongly*). The dataset contains 433 respondents.

Items are prefixed by their dimension (**E**xtraversion, **A**greeableness, **C**onscientiousness, **N**euroticism, **O**penness) followed by the item number. Items marked **R** are reverse-scored before analysis.

**John, O. P., & Srivastava, S. (1999).** The Big Five trait taxonomy: History, measurement, and theoretical perspectives. In L. A. Pervin & O. P. John (Eds.), *Handbook of personality: Theory and research* (2nd ed., pp. 102–138). Guilford Press.

## Item List

```{r itemlist, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  df_lookup[, c("label", "dimension", "reverse", "description")],
  col.names = c("Label", "Dimension", "Reverse-scored", "Item wording"),
  row.names = FALSE
)
```

## Inter-Item Correlation Matrix

```{r corrmatrix, message=FALSE, warning=FALSE, results='show', dev='png'}
cor_mat  <- cor(df[, factor_order], use="pairwise.complete.obs")
cor_long <- melt(cor_mat)
cor_long$Var1 <- factor(cor_long$Var1, levels=factor_order)
cor_long$Var2 <- factor(cor_long$Var2, levels=factor_order)

ggplot(cor_long, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color="white", linewidth=0.2) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="r") +
  theme_bw(base_size=9) +
  theme(axis.text.x=element_text(angle=90, hjust=1, size=7),
        axis.text.y=element_text(size=7),
        axis.title=element_blank(),
        panel.grid=element_blank()) +
  ggtitle("BFI-44 inter-item correlation matrix\nItems ordered by dimension; R = reverse-scored")
```

## Subscale Correlations

```{r subscalecor, message=FALSE, warning=FALSE, dev='png', results='show'}
subscale_scores <- as.data.frame(sapply(names(factors), function(f) {
  cols <- item_name_map[sprintf("pers%02d", factors[[f]])]
  rowMeans(df[, cols], na.rm=TRUE)
}))

cat("Subscale descriptives\n")
print(psych::describe(subscale_scores))

cat("\nSubscale intercorrelations\n")
sub_cor <- round(cor(subscale_scores, use="pairwise.complete.obs"), 2)
print(sub_cor)

sub_long <- melt(sub_cor)
ggplot(sub_long, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color="white", linewidth=0.5) +
  geom_text(aes(label=value), size=3.5) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="r") +
  theme_bw(base_size=10) +
  theme(axis.title=element_blank(), panel.grid=element_blank()) +
  ggtitle("BFI-44 subscale intercorrelations")
```

## Exploratory Factor Analysis (5 Factors)

```{r efa, message=FALSE, warning=FALSE, results='show', dev='png'}
set.seed(42)
efa <- fa(df, nfactors=5, rotate="oblimin", fm="ml", scores="regression")

print(efa$loadings, cutoff=0.30, sort=TRUE)

loadings_mat           <- as.data.frame(unclass(efa$loadings))
colnames(loadings_mat) <- paste0("F", 1:5)
loadings_mat$item      <- rownames(loadings_mat)
loadings_mat$expected  <- item_labels[loadings_mat$item]

load_long      <- melt(loadings_mat, id.vars=c("item","expected"),
                       variable.name="factor", value.name="loading")
load_long$item <- factor(load_long$item, levels=factor_order)

ggplot(load_long, aes(x=factor, y=item, fill=loading)) +
  geom_tile(color="white", linewidth=0.3) +
  geom_text(aes(label=ifelse(abs(loading)>=0.30, round(loading,2), "")), size=2.5) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="loading") +
  facet_grid(expected ~ ., scales="free_y", space="free_y") +
  theme_bw(base_size=9) +
  theme(strip.text.y=element_text(angle=0, size=8),
        axis.text.y=element_text(size=7),
        axis.title=element_blank(),
        panel.grid=element_blank()) +
  ggtitle("EFA factor loadings (oblimin, ML)\nValues >= |0.30| shown; R = reverse-scored item")
```

## Parallel Analysis

```{r parallel, message=FALSE, warning=FALSE, results='show', dev='png'}
fa.parallel(df, fm="ml", fa="fa",
            main="Parallel analysis scree plot (BFI-44)")
```
