---
title: "IPIP-50 / OCEAN Validation"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{OCEAN Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, echo=FALSE, warning=FALSE, message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "code_options.R"))
```

```{r load, include=FALSE}
source(paste0(directory, "code_rwf.R"))
library(psych)
library(ggplot2)
library(reshape2)

load("data/df_ocean.rda")

item_cols <- c(paste0("E", 1:10), paste0("N", 1:10),
               paste0("A", 1:10), paste0("C", 1:10), paste0("O", 1:10))

reverse_items <- c("E2","E4","E6","E8","E10",
                   "N2","N4",
                   "A1","A3","A5","A7",
                   "C2","C4","C6","C8",
                   "O2","O4","O6")

df <- df_ocean[, item_cols]
for (v in reverse_items) df[[v]] <- 5 - df[[v]]

new_names <- ifelse(item_cols %in% reverse_items,
                    paste0(item_cols, "R"),
                    item_cols)
colnames(df) <- new_names

factors <- list(
  Extraversion      = new_names[1:10],
  Neuroticism       = new_names[11:20],
  Agreeableness     = new_names[21:30],
  Conscientiousness = new_names[31:40],
  Openness          = new_names[41:50]
)

factor_order <- unlist(factors)
item_labels  <- unlist(lapply(names(factors), function(f) rep(f, 10)))
names(item_labels) <- factor_order

df_lookup <- data.frame(
  original = item_cols,
  label    = new_names,
  dimension = item_labels[new_names],
  reverse   = item_cols %in% reverse_items,
  description = c(
    # Extraversion E1-E10
    "Am the life of the party.",
    "Don't talk a lot.",
    "Feel comfortable around people.",
    "Keep in the background.",
    "Start conversations.",
    "Have little to say.",
    "Talk to a lot of different people at parties.",
    "Don't like to draw attention to myself.",
    "Don't mind being the center of attention.",
    "Am quiet around strangers.",
    # Neuroticism N1-N10
    "Get stressed out easily.",
    "Am relaxed most of the time.",
    "Worry about things.",
    "Seldom feel blue.",
    "Am easily disturbed.",
    "Get upset easily.",
    "Change my mood a lot.",
    "Have frequent mood swings.",
    "Get irritated easily.",
    "Often feel blue.",
    # Agreeableness A1-A10
    "Feel little concern for others.",
    "Am interested in people.",
    "Insult people.",
    "Sympathize with others' feelings.",
    "Am not interested in other people's problems.",
    "Have a soft heart.",
    "Am not really interested in others.",
    "Take time out for others.",
    "Feel others' emotions.",
    "Make people feel at ease.",
    # Conscientiousness C1-C10
    "Am always prepared.",
    "Leave my belongings around.",
    "Pay attention to details.",
    "Make a mess of things.",
    "Get chores done right away.",
    "Often forget to put things back in their proper place.",
    "Like order.",
    "Shirk my duties.",
    "Follow a schedule.",
    "Am exacting in my work.",
    # Openness O1-O10
    "Have a rich vocabulary.",
    "Have difficulty understanding abstract ideas.",
    "Have a vivid imagination.",
    "Am not interested in abstract ideas.",
    "Have excellent ideas.",
    "Do not have a good imagination.",
    "Am quick to understand things.",
    "Use difficult words.",
    "Spend time reflecting on things.",
    "Am full of ideas."
  ),
  stringsAsFactors = FALSE
)
```

## Description

The **IPIP-50** (International Personality Item Pool; Goldberg, 1992, 1999) is a 50-item public-domain measure of the Big Five personality dimensions. Respondents rated each item on a 1–5 scale (*Very Inaccurate* to *Very Accurate*); 0 encodes missing data. The dataset contains responses from a large online sample.

Items are grouped into five dimensions: **E**xtraversion, **N**euroticism, **A**greeableness, **C**onscientiousness, and **O**penness to Experience. Items marked **R** are reverse-keyed and are recoded (5 − *x*) before analysis.

**Goldberg, L. R. (1992).** The development of markers for the Big-Five factor structure. *Psychological Assessment*, *4*(1), 26–42.

**Goldberg, L. R. (1999).** A broad-bandwidth, public domain, personality inventory measuring the lower-level facets of several Five-Factor models. In I. Mervielde, I. Deary, F. De Fruyt, & F. Ostendorf (Eds.), *Personality Psychology in Europe* (Vol. 7, pp. 7–28). Tilburg University Press.

## Item List

```{r itemlist, echo=FALSE}
knitr::kable(
  df_lookup[, c("label", "dimension", "reverse", "description")],
  col.names = c("Label", "Dimension", "Reverse-scored", "Item wording"),
  row.names = FALSE
)
```

## Inter-Item Correlation Matrix

```{r corrmatrix, fig.width=10, fig.height=9, dev='png', fig.retina=2}
cor_mat  <- cor(df[, factor_order], use="pairwise.complete.obs")
cor_long <- melt(cor_mat)
cor_long$Var1 <- factor(cor_long$Var1, levels=factor_order)
cor_long$Var2 <- factor(cor_long$Var2, levels=factor_order)

ggplot(cor_long, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color="white", linewidth=0.2) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="r") +
  theme_bw(base_size=9) +
  theme(axis.text.x=element_text(angle=90, hjust=1, size=7),
        axis.text.y=element_text(size=7),
        axis.title=element_blank(),
        panel.grid=element_blank()) +
  ggtitle("IPIP-50 inter-item correlation matrix\nItems ordered by dimension; R = reverse-scored (already recoded)")
```

## Subscale Correlations

```{r subscalecor, fig.width=6, fig.height=5, dev='png', fig.retina=2}
subscale_scores <- as.data.frame(sapply(names(factors), function(f) {
  rowMeans(df[, factors[[f]]], na.rm=TRUE)
}))

cat("Subscale descriptives\n")
print(psych::describe(subscale_scores))

cat("\nSubscale intercorrelations\n")
sub_cor <- round(cor(subscale_scores, use="pairwise.complete.obs"), 2)
print(sub_cor)

sub_long <- melt(sub_cor)
ggplot(sub_long, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color="white", linewidth=0.5) +
  geom_text(aes(label=value), size=3.5) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="r") +
  theme_bw(base_size=10) +
  theme(axis.title=element_blank(), panel.grid=element_blank()) +
  ggtitle("IPIP-50 subscale intercorrelations")
```

## Exploratory Factor Analysis (5 Factors)

```{r efa, fig.width=7, fig.height=10, dev='png', fig.retina=2}
set.seed(42)
efa <- fa(df, nfactors=5, rotate="oblimin", fm="ml", scores="regression")

print(efa$loadings, cutoff=0.30, sort=TRUE)

loadings_mat           <- as.data.frame(unclass(efa$loadings))
colnames(loadings_mat) <- paste0("F", 1:5)
loadings_mat$item      <- rownames(loadings_mat)
loadings_mat$expected  <- item_labels[loadings_mat$item]

load_long      <- melt(loadings_mat, id.vars=c("item","expected"),
                       variable.name="factor", value.name="loading")
load_long$item <- factor(load_long$item, levels=factor_order)

ggplot(load_long, aes(x=factor, y=item, fill=loading)) +
  geom_tile(color="white", linewidth=0.3) +
  geom_text(aes(label=ifelse(abs(loading)>=0.30, round(loading,2), "")), size=2.5) +
  scale_fill_gradient2(low="#2166ac", mid="white", high="#d6604d",
                       midpoint=0, limits=c(-1,1), name="loading") +
  facet_grid(expected ~ ., scales="free_y", space="free_y") +
  theme_bw(base_size=9) +
  theme(strip.text.y=element_text(angle=0, size=8),
        axis.text.y=element_text(size=7),
        axis.title=element_blank(),
        panel.grid=element_blank()) +
  ggtitle("EFA factor loadings (oblimin rotation, 5 factors)\nValues >= |0.30| shown; R = reverse-scored item")
```

## Parallel Analysis

```{r parallel, fig.width=8, fig.height=5, dev='png', fig.retina=2}
fa.parallel(df, fm="ml", fa="fa",
            main="Parallel analysis scree plot (IPIP-50)")
```
